---
title: "Experiment 2 (MLP): Does a hidden-layer nonlinearity entangle class and item?"
author: "Representation Separability project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
    fig_width: 8
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 110, fig.align = "center")
```

# What this analyses

The MLP experiment trains three models on the same synthetic data, differing only
in the readout: **A** (free embedding), **B** (separable-by-construction control),
**C** (entangled-by-construction control), each with a hidden layer
`e -> act(W1 e) -> W2 h`, `act` in {identity, relu, tanh}. We measure separability
of class vs. item structure at **both** sites (embedding `e` and hidden `h`), with
**two** metrics:

- `ratio_*`   — the raw variance-alignment ratio (gauge-*dependent*, kept for
  comparison only).
- `wratio_*`  — the **whitened**, gauge-*invariant* ratio (the one to trust). See
  `NOTES.md` for why the raw ratio is unusable on a hidden layer.

The headline question: **does the nonlinearity make a separable representation
inseparable at the hidden layer?** The decisive quantity is `wratio_hidden -
wratio_embedding` for the separable control **B** — a linear map (identity) cannot
change it (that's our acceptance test), so any increase under relu/tanh is the
nonlinearity's doing.

```{r libs}
library(tidyverse)
library(scales)

# consistent look
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(),
                  plot.title = element_text(face = "bold")))
act_levels <- c("identity", "relu", "tanh")
act_cols   <- c(identity = "grey45", relu = "#D55E00", tanh = "#0072B2")
```

# Load and prepare the data

We read the results CSV and derive three flags used throughout:

- **`converged`** — the model actually learned (expected loss well below the
  uniform-distribution loss `log(vocab_size)`). Cells at/above uniform diverged and
  are excluded.
- **`well_sampled`** — the whitened metric is exact only when there are enough
  samples to estimate the covariance (`n_verbs >= 4 * d`); undersampled cells fall
  back to shrinkage (approximate). Main claims use well-sampled cells only.
- **`delta_w`** / **`delta_raw`** — hidden minus embedding separability, per model.
  This is the core signal.

```{r load}
# adjust the path if you knit from elsewhere
results_path <- file.path("..", "data", "experiment2_mlp_results.csv")

num <- function(x) suppressWarnings(as.numeric(x))

dat <- read_csv(results_path, show_col_types = FALSE) %>%
  mutate(across(c(ratio_embedding, ratio_hidden,
                  wratio_embedding, wratio_hidden, final_loss), num),
         activation = factor(activation, levels = act_levels)) %>%
  mutate(
    uniform_loss = log(vocab_size),          # loss of a uniform predictor
    ceiling      = d / (n_classes - 1),       # max whitened ratio (C's ceiling)
    converged    = final_loss < uniform_loss - 0.4,
    well_sampled = n_verbs >= 4 * d,
    delta_w      = wratio_hidden  - wratio_embedding,   # gauge-invariant signal
    delta_raw    = ratio_hidden   - ratio_embedding
  )

glimpse(dat)
```

## Coverage

How much of the grid is present (the sweep may still be running), and how the
converged / well-sampled filters bite.

```{r coverage}
dat %>%
  filter(model == "B") %>%                     # one row per cell
  count(activation, d, name = "cells") %>%
  pivot_wider(names_from = d, values_from = cells, names_prefix = "d=") %>%
  knitr::kable(caption = "Cells present per activation x d (Model B = 1 per cell)")

dat %>% summarise(
  total_cells      = n() / 3,
  diverged_cells   = sum(!converged) / 3,
  undersampled     = sum(!well_sampled) / 3
) %>% knitr::kable(caption = "Exclusion summary")
```

# Validation 1 — the ruler (Model C)

Model C is entangled by construction: its item variance lies exactly on the class
subspace, so its **raw embedding** ratio must track the closed-form ceiling
`d / (n_classes - 1)`. If it does, the measurement pipeline is trustworthy. (We use
the raw ratio here because C's ceiling is a raw-ratio property; C validates the
ruler regardless of activation.)

```{r ruler}
dat %>%
  filter(model == "C") %>%
  ggplot(aes(ceiling, ratio_embedding, colour = activation)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.5, size = 1.6) +
  scale_colour_manual(values = act_cols) +
  labs(title = "Model C tracks its theoretical ceiling",
       subtitle = "points on the dashed line = the ruler is calibrated",
       x = "theoretical ceiling  d / (n_classes - 1)",
       y = "measured raw ratio (C, embedding)")
```

# Validation 2 — undertraining / divergence

Any cell whose expected loss approaches the uniform-predictor loss learned nothing
(its geometry is near initialisation and reads separable trivially). Cells *above*
uniform diverged outright. We exclude both.

```{r undertraining}
u <- unique(dat$uniform_loss)[1]

dat %>%
  ggplot(aes(final_loss, fill = converged)) +
  geom_histogram(bins = 60) +
  geom_vline(xintercept = u, linetype = 2) +
  annotate("text", x = u, y = Inf, label = "uniform", vjust = 1.5, hjust = -0.1) +
  scale_fill_manual(values = c("TRUE" = "#0072B2", "FALSE" = "#D55E00"),
                    name = "converged") +
  labs(title = "Training quality", x = "expected cross-entropy", y = "cells")

# where do the failures live?
dat %>% filter(!converged) %>% count(activation, d) %>%
  knitr::kable(caption = "Diverged / undertrained cells (excluded) — concentrated at high d")
```

# Validation 3 — the metric works (acceptance test)

The gauge-invariant metric must satisfy a theorem: for the **identity** (linear)
condition, a linear map cannot change linear recoverability, so `wratio_hidden`
must equal `wratio_embedding`. We check that the whitened metric hugs the diagonal
for identity — and that the **raw** metric does *not* (its failure is exactly why
we whiten).

```{r acceptance}
acc <- dat %>% filter(activation == "identity", well_sampled, converged, model == "B")

bind_rows(
  acc %>% transmute(metric = "whitened", emb = wratio_embedding, hid = wratio_hidden),
  acc %>% transmute(metric = "raw",      emb = ratio_embedding,  hid = ratio_hidden)
) %>%
  ggplot(aes(emb, hid, colour = metric)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.6) +
  scale_colour_manual(values = c(whitened = "#0072B2", raw = "#D55E00")) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Identity (linear) control: hidden vs embedding separability",
       subtitle = "whitened hugs the diagonal (correct); raw does not (gauge artifact)",
       x = "embedding separability", y = "hidden separability")
```

# Headline — does the nonlinearity entangle the hidden layer?

The core result. For each activation we plot `delta_w = wratio_hidden -
wratio_embedding` on well-sampled, converged cells. **Model B is the key**: its
embedding is separable by construction, so a positive `delta_w` means the hidden
layer made a *separable* representation entangled.

```{r headline}
core <- dat %>% filter(well_sampled, converged)

core %>%
  filter(model == "B") %>%
  ggplot(aes(activation, delta_w, colour = activation)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_jitter(width = 0.12, alpha = 0.25, size = 1) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.15) +
  scale_colour_manual(values = act_cols, guide = "none") +
  labs(title = "Hidden-layer entanglement of the separable control (Model B)",
       subtitle = "delta > 0 means the hidden layer entangled a separable input",
       x = NULL, y = expression(wratio[hidden] - wratio[embedding]))
```

Identity and tanh sit at ~0 (identity by theorem; tanh because the network keeps
it near-linear — see `NOTES.md`), while **relu entangles**.

# The alpha transfer function — architectural floor vs. data-driven rise

Does the entanglement track the data's separability (`alpha`), or is it present even
on fully separable data (`alpha = 0`)? A non-zero relu value at `alpha = 0` means
the nonlinearity imposes entanglement with *nothing in the data to justify it*.

```{r alpha}
core %>%
  filter(model == "B", d < 256) %>%
  group_by(activation, alpha) %>%
  summarise(delta = mean(delta_w), se = sd(delta_w) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(alpha, delta, colour = activation, fill = activation)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_ribbon(aes(ymin = delta - se, ymax = delta + se), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = act_cols) + scale_fill_manual(values = act_cols) +
  labs(title = "Entanglement vs. data interaction (Model B, d < 256)",
       subtitle = "relu is non-zero even at alpha = 0 (architectural floor) and rises with alpha",
       x = expression(alpha~"(data class-item interaction)"),
       y = expression(wratio[hidden] - wratio[embedding]))
```

# The capacity (d) dependence — real but bounded

Raw `delta_w` grows with `d`, but so does the metric's ceiling (`d/(n_classes-1)`).
To separate a genuine effect from metric scaling we also plot `delta_w` **as a
fraction of the ceiling**. If the fraction keeps rising, capacity genuinely
increases relative entanglement; if it plateaus, the raw growth was mostly scaling.

```{r capacity}
cap <- core %>%
  filter(model == "B", activation == "relu") %>%
  group_by(d) %>%
  summarise(delta = mean(delta_w),
            frac  = mean(delta_w / ceiling), .groups = "drop") %>%
  pivot_longer(c(delta, frac),
               names_to = "view",
               values_to = "value") %>%
  mutate(view = recode(view,
                       delta = "raw (hidden - embedding)",
                       frac  = "as fraction of ceiling"))

cap %>%
  ggplot(aes(factor(d), value, group = 1)) +
  geom_line(colour = act_cols["relu"]) +
  geom_point(colour = act_cols["relu"], size = 2.5) +
  facet_wrap(~view, scales = "free_y") +
  labs(title = "ReLU entanglement vs. capacity (Model B)",
       subtitle = "raw grows with d; the ceiling-fraction rises then plateaus by ~d=64-128",
       x = "embedding / hidden width d", y = NULL)
```

# Summary

```{r summary, echo=FALSE, results='asis'}
cat("
- **Ruler + metric validated.** Model C tracks its ceiling; the whitened metric
  passes the identity acceptance test (hidden ~= embedding) where the raw metric
  fails it.
- **ReLU entangles a separable representation** at the hidden layer, with an
  **architectural floor at alpha = 0** (entanglement with no data interaction) that
  **rises with alpha**.
- **tanh and identity do not** — tanh because the network keeps it near-linear
  (avoidable nonlinearity); identity by construction.
- The capacity effect is **real but bounded**: relative entanglement rises to
  ~d = 64 then plateaus, so the story is established without the (diverging,
  excluded) d = 256 cells.
")
```
