---
title: "Experiment 3 (conversion): does a non-additive interaction force entanglement robustly across lr?"
author: "Representation Separability project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
    fig_width: 8
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 110, fig.align = "center")
```

# What this analyses

Experiment 2 found that a hidden nonlinearity entangles class and item — but the
*magnitude* was lr-dependent (additive data can be fit near-linearly, so
entanglement was optional). Experiment 3 asks whether a genuinely **non-additive**
linguistic interaction — **conversion** (same form used as noun and verb, with
form-specific collocates by POS) — makes entanglement **task-set and robust to
lr** instead.

`interaction_strength` is the dose-response axis: the per-word level of how much
of each form's collocate mass is form-specific-by-POS (non-additive) vs. shared
(additive). 0 = fully additive (reproduces Exp-2's lr-sensitivity), 1 = strongly
non-additive.

Models: **shared** (form-embedding shared across POS — the parameter-sharing model
*forced* to bind on non-additive data), **free** (per-lexeme memorizer — can
*dodge* binding), **C** (entangled ceiling — the ruler). Classes = POS (2), so the
ceiling is `d/(n_pos-1) = d`. The trustworthy metric is the whitened (gauge-invariant)
one, `wratio_*`.

**The key expectation:** additive (strength 0) entanglement should *rise with lr*
(the Exp-2 fragility); conversion (high strength) entanglement should be *flat
across lr* (a forced floor). The contrast is cleanest at low lr.

```{r libs}
library(tidyverse); library(scales)
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold")))
```

# Load and prepare

```{r load}
results_path <- file.path("..", "data", "experiment3_conversion_results.csv")
num <- function(x) suppressWarnings(as.numeric(x))
dat <- read_csv(results_path, show_col_types = FALSE) %>%
  mutate(across(c(ratio_embedding, ratio_hidden, wratio_embedding, wratio_hidden,
                  final_loss, specialization), num),
         model = factor(model, levels = c("shared", "free", "C")),
         activation = factor(activation, levels = c("identity", "relu", "tanh"))) %>%
  mutate(uniform_loss = log(2000),
         ceiling = d,                                  # d/(n_pos-1), n_pos=2
         converged = final_loss < uniform_loss - 0.4,
         well_sampled = n_lexemes >= 4 * d,
         delta_w = wratio_hidden - wratio_embedding)   # the entanglement signal
glimpse(dat)
```

```{r coverage}
dat %>% filter(model == "shared") %>%
  count(interaction_strength, lr, name = "cells") %>%
  pivot_wider(names_from = lr, values_from = cells, names_prefix = "lr=") %>%
  knitr::kable(caption = "Cells present (shared model) by strength x lr")
```

# Validations

## The ruler (Model C)

C is entangled by construction, so its raw embedding ratio should track the
ceiling `d`.

```{r ruler}
dat %>% filter(model == "C") %>%
  ggplot(aes(ceiling, ratio_embedding, colour = activation)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.5) +
  labs(title = "Model C tracks its ceiling (ruler check)",
       x = "ceiling = d", y = "measured raw ratio (C, embedding)")
```

## Undertraining / divergence

```{r undertraining}
u <- unique(dat$uniform_loss)[1]
dat %>% ggplot(aes(final_loss, fill = converged)) +
  geom_histogram(bins = 60) + geom_vline(xintercept = u, linetype = 2) +
  scale_fill_manual(values = c("TRUE" = "#0072B2", "FALSE" = "#D55E00")) +
  labs(title = "Training quality", x = "expected cross-entropy", y = "cells")
```

## Metric acceptance test (identity is flat)

The linear baseline (`activation = identity`) must have `hid ≈ emb` at every
strength (a linear map can't change recoverability). This is also the linear
*measurement control* re-run on the conversion data.

```{r acceptance}
dat %>% filter(activation == "identity", model == "shared", well_sampled, converged) %>%
  ggplot(aes(factor(interaction_strength), delta_w)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_jitter(width = 0.1, alpha = 0.4) +
  stat_summary(fun = mean, geom = "point", colour = "#D55E00", size = 3) +
  labs(title = "Identity (linear) hidden-embedding delta ~ 0 at all strengths",
       x = "interaction_strength", y = expression(wratio[hidden]-wratio[embedding]))
```

# Headline A — entanglement vs. interaction strength (dose-response)

Shared model, ReLU. Does entanglement rise as the task becomes non-additive?

```{r headline-strength}
core <- dat %>% filter(model == "shared", activation == "relu", well_sampled, converged)
core %>% group_by(interaction_strength, lr) %>%
  summarise(d_ = mean(delta_w), se = sd(delta_w)/sqrt(n()), .groups = "drop") %>%
  ggplot(aes(interaction_strength, d_, colour = factor(lr), fill = factor(lr))) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_ribbon(aes(ymin = d_-se, ymax = d_+se), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "Hidden entanglement vs. non-additivity (shared, relu)",
       colour = "lr", fill = "lr",
       x = "interaction_strength", y = expression(wratio[hidden]-wratio[embedding]))
```

# Headline B — the robustness figure (entanglement vs. lr)

**This is the key contrast.** Additive (strength 0) entanglement should *rise
with lr* (Exp-2 fragility); conversion (high strength) should be *flat* (forced
floor, lr-robust).

```{r headline-lr}
core %>% group_by(lr, interaction_strength) %>%
  summarise(d_ = mean(delta_w), se = sd(delta_w)/sqrt(n()), .groups = "drop") %>%
  ggplot(aes(lr, d_, colour = factor(interaction_strength), fill = factor(interaction_strength))) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_ribbon(aes(ymin = d_-se, ymax = d_+se), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_x_log10() + scale_colour_viridis_d() + scale_fill_viridis_d() +
  labs(title = "Additive is lr-fragile; conversion is lr-robust (shared, relu)",
       subtitle = "flat lines = entanglement forced by the task, not the optimizer",
       colour = "strength", fill = "strength",
       x = "learning rate (log)", y = expression(wratio[hidden]-wratio[embedding]))
```

# Parameter-sharing vs. memorizer vs. ceiling

The forcing is specific to the parameter-sharing (**shared**) model; the
per-lexeme **memorizer (free)** can dodge binding; **C** is the ceiling.

```{r models}
dat %>% filter(activation == "relu", well_sampled, converged, lr == 0.003) %>%
  group_by(model, interaction_strength) %>%
  summarise(d_ = mean(delta_w), .groups = "drop") %>%
  ggplot(aes(interaction_strength, d_, colour = model)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "shared (forced) vs. free (dodges) vs. C (ceiling), relu, lr=0.003",
       x = "interaction_strength", y = expression(wratio[hidden]-wratio[embedding]))
```

# Specialization — does the model FIT the conversion?

A secondary diagnostic (interpret cautiously — softmax lets even the linear model
partially specialize): does a lexeme favour its own-POS specific collocates?

```{r spec}
dat %>% filter(model == "shared", well_sampled, converged, !is.na(specialization)) %>%
  group_by(activation, interaction_strength) %>%
  summarise(sp = mean(specialization), .groups = "drop") %>%
  ggplot(aes(interaction_strength, sp, colour = activation)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "Fit of the conversion behaviour (shared)",
       x = "interaction_strength", y = "specialization (own-POS minus other-POS prob)")
```

# Summary

```{r summary, echo=FALSE, results='asis'}
cat("
- **Ruler + metric validated on the new data** (C at ceiling; identity flat at all strengths).
- **Headline B** is the point: if additive (strength 0) rises with lr while high-strength
  lines stay flat, entanglement is *forced by the non-additive task*, not by the optimizer
  -- resolving the Exp-2 lr-sensitivity.
- The forcing is **specific to the parameter-sharing (shared) model**; the memorizer dodges.
- Caveats: entanglement magnitude is modest (read the *contrast* vs. additive, cleanest at
  low lr), and specialization is a weak secondary point (linear partially fits via softmax).
")
```
