---
title: "Experiment 5b: does compositional (additive) structure entangle class and item more than interacting structure — at matched budget?"
author: "Representation Separability project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
    fig_width: 8
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 110, fig.align = "center")
```

# What this analyses

Exp 4/5 hinted that *additive* (compositional) class structure is more inseparable
than *interacting* structure, but that comparison confounded structure with **cue
richness** (the interacting generator carried more category tokens). Experiment 5b
removes the confound: every config gets the **same** number of category tokens `S`,
and an **interaction fraction φ** only reallocates that fixed budget from additive
per-(factor,level) cues (φ=0) to interacting per-(factor-pair) cues (φ=1). So the
*amount* of category signal is held constant and only additive↔interacting
**structure** varies.

**The question:** at matched budget (and matched capacity), does inseparability change
with φ? If `cvwh` falls as φ rises (additive most inseparable), compositional structure
genuinely entangles class and item more — it is not cue richness.

Measure = cross-validated whitened separability (`cvwh_hidden`), with the corrected
**parallel-analysis rank** for the class subspace (a lenient rank threshold over-counts
for low-rank additive class means and manufactures inseparability — see METHOD.md).

**Regime note:** these runs use batch=2048 (GPU). Larger batch = less gradient noise =
more separable, so absolute `cvwh` is ~4x smaller than the batch=64 numbers we saw
earlier. The *learned regime* (relu above the identity floor) and the φ ordering were
verified to survive this regime; read the **contrasts**, not the absolute scale.

```{r libs}
library(tidyverse); library(splines)
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold")))
```

# Load and prepare

```{r load}
num <- function(x) suppressWarnings(as.numeric(x))
dat <- read_csv(file.path("..", "data", "experiment5b_interaction_matched_results.csv"),
                show_col_types = FALSE) %>%
  mutate(across(c(phi, capacity, cvwh_hidden, m_eff, k_item, final_loss, n_over_d), num),
         d = as.integer(d), K = as.integer(K),
         activation = factor(activation, levels = c("relu", "identity"))) %>%
  mutate(uniform_loss = log(12000),
         converged = final_loss < uniform_loss - 0.4,
         artifact  = cvwh_hidden > 1.05)          # bounded whitened measure: >1 = undersampling/rank artifact
glimpse(dat)
```

```{r coverage}
dat %>% count(d, K, activation) %>%
  pivot_wider(names_from = activation, values_from = n) %>%
  knitr::kable(caption = "Cells present, by d x K x activation")
```

# Validation

The measure must be clean: identity (linear) at the floor, and essentially no
`cvwh > 1` artifacts (those are undersampling / rank over-counting, not real).

```{r validation}
dat %>% group_by(d) %>%
  summarise(identity_mean = mean(cvwh_hidden[activation == "identity"], na.rm = TRUE),
            frac_artifact = mean(cvwh_hidden > 1.05, na.rm = TRUE),
            mean_n_over_d = mean(n_over_d, na.rm = TRUE), n = n(), .groups = "drop") %>%
  knitr::kable(digits = 3, caption = "identity ~0 and frac_artifact ~0 => clean measurement (esp. check d=64)")
```

# Headline: inseparability vs. interaction fraction φ

relu only, converged, artifacts dropped. If `cvwh` **falls as φ rises**, additive
(compositional) structure is more inseparable than interacting — at matched budget.

```{r headline}
relu <- dat %>% filter(activation == "relu", converged, !artifact)
relu %>%
  group_by(d, phi) %>%
  summarise(cvwh = mean(cvwh_hidden), se = sd(cvwh_hidden) / sqrt(n()), n = n(), .groups = "drop") %>%
  ggplot(aes(phi, cvwh, colour = factor(d), fill = factor(d))) +
  geom_ribbon(aes(ymin = cvwh - se, ymax = cvwh + se), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "Inseparability falls as structure goes additive -> interacting (matched budget)",
       subtitle = "phi=0 additive/compositional, phi=1 interacting; per-config category tokens held fixed",
       x = "interaction fraction phi", y = "CV-whitened inseparability (relu)", colour = "d", fill = "d")
```

## Binned means (with the identity floor for reference)

```{r binned}
dat %>% filter(converged, !artifact) %>%
  group_by(activation, phi) %>%
  summarise(cvwh = mean(cvwh_hidden), .groups = "drop") %>%
  pivot_wider(names_from = activation, values_from = cvwh) %>%
  knitr::kable(digits = 4, caption = "relu falls with phi; identity stays at the floor at every phi")
```

# Is it structure, or residual capacity? (regression)

φ is designed to hold capacity roughly fixed, but we confirm it with a flexible
capacity control: the φ coefficient is the structure effect *at matched capacity*.

```{r regression}
fit_dat <- relu %>% filter(!is.na(capacity), !is.na(cvwh_hidden))
fit <- lm(cvwh_hidden ~ ns(capacity, 4) + phi + factor(d), data = fit_dat)
co <- as.data.frame(coef(summary(fit))); co$term <- rownames(co)
co %>% filter(term == "phi") %>% select(term, Estimate, `Std. Error`, `Pr(>|t|)`) %>%
  knitr::kable(digits = 4, caption = "phi effect at matched capacity. <0 = additive more inseparable.")
cat(sprintf("capacity is ~flat across phi: %s\n",
            paste(round(tapply(fit_dat$capacity, fit_dat$phi, mean), 2), collapse = " / ")))
```

# By K (robustness)

```{r by-k}
relu %>% group_by(K, phi) %>% summarise(cvwh = mean(cvwh_hidden), .groups = "drop") %>%
  ggplot(aes(phi, cvwh, colour = factor(K))) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  labs(title = "phi effect across K (number of factors)",
       x = "phi", y = "cvwh (relu)", colour = "K")
```

# Summary

```{r summary, echo=FALSE, results='asis'}
cat("
- **Measurement clean** if identity ~0 and no cvwh>1 artifacts across d (check d=64).
- **Headline**: if cvwh falls monotonically with phi (and the phi coefficient is <0 at
  matched capacity), then compositional/additive category structure entangles class and
  item MORE than interacting structure -- a genuine STRUCTURE effect, not cue richness
  (which is held fixed here) and not capacity (matched).
- **Regime**: batch=2048 -> absolute cvwh is small; read the phi contrast and the
  relu-vs-identity gap, not the raw magnitude. The learned regime survives this batch.
- **Mechanism (hypothesis)**: additive cues force the model to COMPOSE factor-signals to
  identify a config, smearing the category across factor directions that overlap item;
  interacting cues are config-specific landmarks the model reads directly, keeping the
  category localized and item separable.
")
```
