---
title: "Experiment 5: does class structure affect inseparability beyond capacity?"
author: "Representation Separability project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: show
    fig_width: 8
    fig_height: 4.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 110, fig.align = "center")
```

# What this analyses

Experiment 4 showed inseparability rises with **capacity** = `(m_eff + k_item)/d`,
and that class *structure* seemed to matter — but it compared structures at matched
*load* (same K), which confounds structure with capacity (interacting and additive
factored structures have different effective class rank). Experiment 5 fixes two
things:

1. **Measurement**: `n_forms = 6·d`, so n/d ≥ 24 everywhere and the CV-whitened
   estimate no longer inflates for the ReLU representation at larger d (Exp 4's
   high-d cells were contaminated at n/d ≈ 8).
2. **The confound**: the three structures (`single`, `factored_add`, `factored_int`)
   are swept over **overlapping capacity**, and matched on *measured* capacity here
   via `cvwh ~ ns(capacity) + structure`. A structure effect at matched capacity
   means organization matters beyond capacity; none means Exp 4's gap was capacity.

The measure is the cross-validated whitened separability (`cvwh_hidden`):
gauge-invariant, self-calibrating (separable ≈ 0, chance ≈ 1), undersampling-robust.

**Caveat carried into the interpretation:** `factored_int` has more category tokens
than `factored_add` (per-factor + factor-pair), so "interaction" is confounded with
*cue richness* — read the additive-vs-interacting direction with that in mind.

```{r libs}
library(tidyverse); library(splines)
theme_set(theme_minimal(base_size = 12) +
            theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold")))
```

# Load and prepare

```{r load}
num <- function(x) suppressWarnings(as.numeric(x))
dat <- read_csv(file.path("..", "data", "experiment5_capacity_matched_results.csv"),
                show_col_types = FALSE) %>%
  mutate(across(c(capacity, cvwh_hidden, m_eff, k_item, final_loss, n_over_d), num),
         d = as.integer(d),
         structure = factor(structure, levels = c("single", "factored_add", "factored_int")),
         activation = factor(activation)) %>%
  mutate(uniform_loss = log(1000),
         converged = final_loss < uniform_loss - 0.4,
         artifact  = cvwh_hidden > 1.05 & activation == "identity")  # gauge-invariant linear can't exceed ~1 unless mis-estimated
glimpse(dat)
```

```{r coverage}
dat %>% count(structure, d, activation) %>%
  pivot_wider(names_from = activation, values_from = n) %>%
  knitr::kable(caption = "Cells present so far, by structure x d x activation")
```

# Measurement validation (did the n_forms = 6d fix work?)

The linear (`identity`) representation is gauge-invariantly separable, so it must
read ~0 everywhere below the capacity wall, and `cvwh` should essentially never
exceed ~1. If identity sits at the floor and there are no `cvwh > 1` artifacts,
the high-d contamination from Exp 4 is gone.

```{r validation}
dat %>% group_by(d, activation) %>%
  summarise(mean_cvwh = mean(cvwh_hidden, na.rm = TRUE),
            frac_gt1 = mean(cvwh_hidden > 1.05, na.rm = TRUE),
            mean_n_over_d = mean(n_over_d, na.rm = TRUE), n = n(), .groups = "drop") %>%
  knitr::kable(digits = 3, caption = "Identity should be ~0; frac_gt1 should be ~0 (clean measurement)")
```

# The learned regime: ReLU entangles below the capacity wall

Identity marks the *floor* (separation achievable); where ReLU sits above it at
capacity < 1, the network is choosing partial inseparability with room to spare.

```{r learned-regime}
dat %>% filter(converged) %>%
  ggplot(aes(capacity, cvwh_hidden, colour = activation)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey60") +
  geom_hline(yintercept = 1, linetype = 3, colour = "grey70") +
  geom_point(alpha = 0.35, size = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.9, linewidth = 1) +
  labs(title = "Inseparability vs. capacity: relu rises below the wall, identity stays at the floor",
       subtitle = "dashed = capacity 1 (theoretical wall); dotted = chance",
       x = "capacity = (m_eff + k_item)/d", y = "CV-whitened inseparability")
```

# Headline: structure at matched capacity

The key figure — ReLU only, `cvwh` vs. capacity, one smooth per structure. If the
smooths **separate vertically**, structure matters at matched capacity.

```{r headline-structure}
relu <- dat %>% filter(activation == "relu", converged, !is.na(capacity))
relu %>%
  ggplot(aes(capacity, cvwh_hidden, colour = structure, fill = structure)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey60") +
  geom_point(alpha = 0.35, size = 1.2) +
  geom_smooth(method = "loess", se = TRUE, span = 1.0, alpha = 0.12, linewidth = 1) +
  labs(title = "Class structure shifts inseparability at matched capacity",
       subtitle = "vertical separation between smooths = a structure effect beyond capacity",
       x = "capacity = (m_eff + k_item)/d", y = "CV-whitened inseparability (relu)")
```

## Binned view (matched-capacity means)

```{r binned}
relu %>%
  mutate(cap_bin = cut(capacity, c(0.4, 0.6, 0.8, 1.0, 1.3, 9))) %>%
  group_by(cap_bin, structure) %>%
  summarise(cvwh = mean(cvwh_hidden), mean_cap = mean(capacity), n = n(), .groups = "drop") %>%
  select(cap_bin, structure, mean_cap, cvwh, n) %>%
  pivot_wider(names_from = structure, values_from = c(cvwh, n, mean_cap)) %>%
  knitr::kable(digits = 3, caption = "relu cvwh by capacity bin x structure (check mean_cap ~ matched within bin)")
```

## The regression: structure effect controlling for capacity

`ns(capacity, 4)` flexibly absorbs the capacity trend; the `structure` coefficients
are then the shifts *at matched capacity*, relative to `single`. Restricted to the
overlapping-capacity region.

```{r regression}
fit_dat <- relu %>% filter(capacity >= 0.5, capacity <= 1.3)
fit <- lm(cvwh_hidden ~ ns(capacity, 4) + structure, data = fit_dat)
co <- as.data.frame(coef(summary(fit))); co$term <- rownames(co)
co %>% filter(grepl("structure", term)) %>%
  select(term, Estimate, `Std. Error`, `Pr(>|t|)`) %>%
  knitr::kable(digits = 4, caption = "Structure shift at matched capacity (vs single). >0 = more inseparable.")
cat(sprintf("n = %d cells in the matched-capacity regression\n", nrow(fit_dat)))
```

# Interpretation & caveats

```{r summary, echo=FALSE, results='asis'}
cat("
- **Measurement fix confirmed** if identity ~0 and no cvwh>1 artifacts across d
  (including d=64): the Exp-4 high-d contamination is gone.
- **Learned regime**: relu elevated above the identity floor for capacity < 1 =
  the model represents class and item as partially inseparable even when a
  separable code is achievable.
- **Structure beyond capacity**: if the `structure` coefficients are non-zero, class
  *organization* matters, not just class *amount* -- the answer to Exp 5's question.
- **Read additive-vs-interacting cautiously**: `factored_int` carries more category
  tokens (factor-pair landmarks) than `factored_add`, so the direction likely
  reflects *cue richness* (more landmarks -> cleaner category -> more separable),
  not interaction per se. A token-budget-matched generator is the clean follow-up.
")
```
