---
title: "Are Lexical Representations Separable? Item and Category in Learned Distributed Representations"
author:
  - name: "Anonymous"
    affiliations:
      - name: "Anonymous Institution"
    email: anon@example.edu
abstract: |
  PLACEHOLDER --- author to write.
bibliography: refs.bib
acl-mode: review
number-sections: true
tbl-cap-location: bottom
format:
  acl-pdf:
    keep-tex: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr); library(dplyr); library(tidyr); library(ggplot2); library(kableExtra)
library(patchwork); library(scales)

DATA <- file.path("..", "data")
rd <- function(f) suppressMessages(read_csv(file.path(DATA, f), show_col_types = FALSE))

# UO palette, as in the companion papers. Green = the linear reference, gold = the
# nonlinearity; amber and blue are held for the model families in Experiments 2-3.
UO <- c(green = "#154733", green2 = "#007030", gold = "#F0B323",
        amber = "#875E00", blue = "#005581", grey = "#6E6E6E")
ACT_COL <- c(identity = UO[["green2"]], relu = UO[["gold"]])
ACT_LAB <- c(identity = "linear", relu = "ReLU")

theme_paper <- function(base = 9) {
  theme_minimal(base_size = base) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, colour = "grey88"),
          strip.text = element_text(face = "bold", size = base),
          legend.position = "top", legend.margin = margin(0, 0, -4, 0),
          legend.key.height = unit(9, "pt"),
          axis.title = element_text(size = base))
}

# ---- Experiment 1: the artificial-language grid ----
# 7,560 cells, measured with unified_split -- the SAME function and arguments Experiment 2
# uses. The two independent estimates are two halves of each cell's observations, so there is
# no alignment step and no projection. Capacity is a reported variable, never a filter.
toy <- rd("experiment10_abc_grid.csv") |>
  mutate(act = factor(activation, c("identity", "relu"), c("linear", "ReLU")),
         # the share of a random vector that lands in the class direction by chance. An
         # orthogonality reading means nothing without it, and it moves 16-fold across the
         # widths used here, so it is carried alongside rather than argued about.
         floor = k_class / d)

# The linear arm is a negative control by ALGEBRA, not by stipulation: with additive inputs a
# linear map gives cell means of the form A(f)+B(c), whose interaction is identically zero
# however much the language contains. Whatever it reports is the measure's own noise, read at
# that exact width, sample size and context level.
toy_null <- toy |> group_by(act, planted = w_int > 0) |>
  summarise(n = n(), med = median(size_interaction),
            p95 = quantile(size_interaction, .95),
            flagged = mean(sig_interaction), .groups = "drop")

# Does the measured interaction track the planted one? Binned, because the achieved share is
# continuous (neither the softmax nor the normalisation is linear in the weights).
toy_track <- toy |> filter(w_item > 0, w_class > 0) |>
  mutate(bin = cut(ach_int, c(-Inf, .005, .05, .15, .30, .55, Inf),
                   labels = c("none", ".005-.05", ".05-.15", ".15-.30", ".30-.55", ">.55"))) |>
  group_by(act, bin) |>
  summarise(n = n(), ach = median(ach_int), size_int = median(size_interaction),
            lo = quantile(size_interaction, .25), hi = quantile(size_interaction, .75),
            flagged = mean(sig_interaction), .groups = "drop")

# Capacity: below 1 every component can have its own axis, above 1 they cannot.
toy_cap <- toy |> filter(w_int > 0, w_item > 0, w_class > 0) |>
  group_by(act, capacity) |>
  summarise(n = n(), size_int = median(size_interaction),
            overlap = median(leak_item_into_class, na.rm = TRUE),
            floor = median(floor),
            ratio = median(leak_item_into_class / floor, na.rm = TRUE), .groups = "drop")

# Context is the nuisance dial: how much of the target is about something other than the item,
# the class and their conjunction. The high end is the regime Experiment 2 sits in.
toy_noise <- toy |> filter(w_int > 0, w_item > 0, w_class > 0) |>
  mutate(cbin = cut(ach_ctx, c(-Inf, .2, .5, .8, Inf),
                    labels = c("<20%", "20-50%", "50-80%", ">80%"))) |>
  group_by(act, cbin) |>
  summarise(n = n(), size_int = median(size_interaction),
            flagged = mean(sig_interaction), .groups = "drop")

g <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
cm <- function(x) format(x, big.mark = ",")
nullget <- function(a, pl, col) toy_null[[col]][toy_null$act == a & toy_null$planted == pl]
capget <- function(a, cap, col) toy_cap[[col]][toy_cap$act == a & toy_cap$capacity == cap]

# ---- Experiment 2: language model representations ----
grid_stats <- rd("methods_grid_stats.csv")
gs <- function(constr, col) {
  v <- grid_stats[[col]][grid_stats$construction == constr]
  if (is.numeric(v)) round(median(v)) else v[1]
}

MODELS <- tibble::tribble(
  ~model,                                       ~family,  ~size,   ~d,
  "znhoughton/opt-babylm-125m-20eps-seed964",   "BabyLM", "125M",  768,
  "znhoughton/opt-babylm-350m-20eps-seed964",   "BabyLM", "350M",  1024,
  "znhoughton/opt-babylm-1.3B-20eps-seed964",   "BabyLM", "1.3B",  2048,
  "EleutherAI/pythia-160m",                     "Pythia", "160M",  768,
  "EleutherAI/pythia-410m",                     "Pythia", "410M",  1024,
  "EleutherAI/pythia-1.4b",                     "Pythia", "1.4B",  2048)

# The three same-token constructions carried in the body. Each file reports the same measure;
# the POS file prefixes its columns std_ because it also stores an unstandardized read.
llm <- bind_rows(
  rd("llm_unified.csv") |>
    transmute(model, layer, d, n_items, construction = "POS (noun/verb)",
              size_item = std_size_item, size_class = std_size_class,
              size_int = std_size_interaction, sig = std_sig_interaction,
              leak = std_leak_item_into_class),
  rd("llm_role.csv") |>
    transmute(model, layer, d, n_items, construction = "role (subject/object)",
              size_item, size_class, size_int = size_interaction,
              sig = sig_interaction, leak = leak_item_into_class),
  rd("llm_metaphor.csv") |>
    transmute(model, layer, d, n_items, construction = "metaphor (literal/figurative)",
              size_item, size_class, size_int = size_interaction,
              sig = sig_interaction, leak = leak_item_into_class)) |>
  left_join(MODELS, by = "model") |>
  mutate(construction = factor(construction, c("POS (noun/verb)", "role (subject/object)",
                                               "metaphor (literal/figurative)")),
         depth = layer / ave(layer, model, construction, FUN = max))  # 0 = embeddings, 1 = final
```

# Introduction

::: {.content-visible when-format="acl-pdf"}
```{=html}
<!-- PLACEHOLDER: author to write. Frames the paper as an extension of Exemplars in
     Disguise: if a learner's behavior cannot distinguish exemplar storage from
     abstraction, ask instead whether its representations separate into an item part and
     a class part. Ends with the Present Study paragraph setting up the three
     experiments. -->
```
:::

# Experiment 1: What a Learner Builds from a Known Language {#sec-exp1}

Experiment 1 has two jobs, and comparing artificial learners to language models is not one of them. The first is to establish that the measure reports what is there: we build an artificial language whose structure we control, dial each of the three effects independently from nothing to everything, and check what the measure recovers. The second is to ask what a learner *does* with that structure when a separable solution is available to it and it has room to find one. Whether any of this holds of real language is Experiment 2's question, and the numbers there stand on their own.

## Dataset {#sec-exp1-data}

The artificial language is a next-token prediction task over a vocabulary of `r cm(600)` tokens. Every observation is a triple: a **form** (one of `r max(toy$n_form)`), a **class** (one of `r max(toy$n_class)`), and a **context** drawn from a pool of 400. Each (form, class) cell is observed `r max(toy$n_obs)` times, each time in a different context. That is what a word in a corpus looks like: one type, many occurrences, each in different company.

The target distribution for an observation is built from four independently generated terms, weighted and passed through a softmax:

$$
P(\text{next token} \mid f, c, x) = \operatorname{softmax}\Big(s\big[\,w_\alpha\,\text{item}(f) + w_\beta\,\text{class}(c) + w_\gamma\,\text{int}(f,c) + w_{\text{ctx}}\,\text{context}(x)\,\big]\Big)
$$ {#eq-toy}

The item and class terms are what those factors contribute on their own. The **interaction** term is built by multiplying the form's and the class's latent codes together elementwise before mapping them onto the vocabulary, and that product is the point: a *sum* can always be split into a form part plus a class part, and a *product* cannot, so this term is joint in a way no decomposition into the two can absorb. The **context** term is nuisance, but it is real information the model has to encode, exactly as the rest of a sentence is for a language model. Every term is normalized to unit scale before weighting, so the weights set composition and nothing else; total signal strength, and hence task difficulty, is constant across the grid.

**The grid.** The three effect weights are crossed **factorially**, each over $\{0, 0.5, 1, 2\}$, giving 63 combinations. Crossing rather than choosing is what supplies the nulls: 15 combinations plant no item effect, 15 no class effect, and 15 no interaction, so the measure is asked to report zero in every case where zero is the truth. These are crossed in turn with the context weight (three levels), the hidden width, the output nonlinearity and five random draws of the language, for `r format(nrow(toy), big.mark = ",")` cells, of which `r sprintf("%.1f%%", 100 * mean(toy$converged))` trained to a loss plateau.

**Capacity.** The structure needs `r max(toy$rank)` dimensions and the hidden layer has `r paste(sort(unique(toy$d)), collapse = ", ")`, so their ratio, which we call **capacity**, runs from `r g(min(toy$capacity), 3)` to `r g(max(toy$capacity), 1)`. Below one there is room to give every component its own axis, so any entanglement was chosen; above one, dimension counting makes separation impossible and entanglement is forced. Capacity is reported as a variable, never used to select cells.

## Method {#sec-exp1-method}

### The learner

The model has an embedding per **form**, an embedding per **class**, and an embedding per **context**. It has none per *cell*. A cell's representation is therefore **computed** from shared pieces rather than stored in a parameter of its own, which is what a transformer does with a token and its context, and it is what lets the measurement here be identical to Experiment 2's. The three embeddings are concatenated, passed through one hidden layer of width $d$ with either an identity or a ReLU nonlinearity, and read out linearly to the vocabulary. We measure the hidden layer.

Training minimizes the exact expected cross-entropy against the true target rather than a sampled-token estimate, so there is no minibatch noise, and stops when that loss plateaus. Stopping depends on nothing derived from the measurement, but it does stop the two conditions at different points, so each cell records whether it plateaued or exhausted its budget.

**The linear condition is a negative control by algebra, not by stipulation.** With additive inputs a linear map produces cell means of the form $A(f) + B(c)$, and such a grid has $\gamma = 0$ *identically* (@sec-app-math). The linear learner therefore cannot represent an interaction however much the language contains, and it visibly fails to fit when the language contains one. Whatever the measure reports there is its own noise, read off at that exact width, sample size and context level, which is a stronger calibration than any argument about noise floors. ReLU is the condition that can build an interaction, and the one that corresponds to a transformer: nonlinear inside, linear readout.

### Decomposing a representation {#sec-measure}

Let $M[i,c] \in \mathbb{R}^d$ be the mean representation of item $i$ used at class level $c$. We split that grid into four geometric components,

$$
M[i,c] \;=\; \mu + \alpha(i) + \beta(c) + \gamma(i,c),
$$ {#eq-decomp}

by centering: $\mu$ is the grand mean, $\alpha(i)$ is item $i$'s row mean minus $\mu$, $\beta(c)$ is level $c$'s column mean minus $\mu$, and $\gamma(i,c)$ is what remains. The familiar reading is the main-effects-plus-interaction partition, with the difference that the dependent variable is a vector, so each component is a *direction* in representational space rather than a number. In mixed-model terms $\alpha$ is a by-item random intercept, $\beta$ is the fixed effect of class, and $\gamma$ is the by-item random slope for class; the question of this paper is how much of the class effect is carried by $\beta$ rather than by $\gamma$. We use that mapping for intuition only. Nothing is fitted, there is no shrinkage, and no distribution is assumed over the by-item terms.

A representation is **separable** if a word's representation is the sum of an item part and a class part, with the two on their own axes. It can fail in two distinguishable ways, and we measure them separately. It fails if the components are **non-orthogonal**, so that projecting out the class also removes item information; and it fails if there is a **third component**, an item-by-class term that is neither an item part nor a class part, whatever its orientation. A representation with a large $\gamma$ does not factor into item plus class even if $\gamma$ sits on an axis of its own.

Because the four components are mutually orthogonal *as sources*, their squared lengths partition the structure of the grid exactly, and we report each as a share of the total. **Interaction size** is $\gamma$'s share, which measures the second kind of failure. For the first we report two projections: **item-class overlap**, the fraction of the item effect's energy lying inside the class effect's subspace, and **interaction overlap**, how much the interaction's subspace shares directions with the item subspace. Zero on both, with a negligible $\gamma$, is a cleanly separable representation.

**Interpreting the interaction.** Because $\gamma$ is obtained by subtraction it invites reading as a residual in the regression sense, the part a model failed to capture. It is not. The decomposition is saturated: $\mu$, $\alpha$, $\beta$ and $\gamma$ exhaust the degrees of freedom of the grid, so $\gamma$ is the final term of a complete description rather than the leftover of an incomplete one. The residual stratum lies one level below, in the scatter of individual observations *within* each cell. That distinction is not free, however, since cell means are estimated from finitely many observations and within-cell error propagates upward into $\gamma$: a squared length is bounded below by its own noise floor and cannot read zero even when the true interaction is zero. This is the situation in which an analysis of variance tests an interaction against the within-cell error term, and it requires replication for the same reason.

**Replication and the null.** We therefore estimate the grid **twice** from independent data and keep only what the two estimates agree on, reporting the signed inner product $\langle \gamma_A, \gamma_B \rangle$ rather than $\lVert\gamma\rVert^2$. Independent errors cancel in expectation, so this is unbiased at any noise level; and because it can be negative it supports a genuine test, which a squared length cannot. Significance comes from permuting which item of one estimate is paired with which item of the other, which preserves every magnitude while destroying the correspondence. A component's *direction* is only defined if the component exists, so item-class overlap is reported only when both components it relates are present; otherwise it is withheld.

Because every cell is observed many times, the two estimates are two disjoint halves of its observations. This is exactly what the language models supply, so the same function is called with the same arguments in both experiments; no alignment between estimates is needed and no projection is applied. What the halves cancel is the difference between the contexts one happened to average over and the contexts the other did, which is the same thing they cancel in a corpus.

**What "recovers" means here.** The weights in @eq-toy describe the *target*; the measure reads the *representation*. These are different objects and their numbers do not match: a learner handed a language that is 85% interaction can build a representation in which the interaction is a far smaller share, because the representation only has to be enough to produce the answer. Validation asks that the measure **track** what was planted, read zero when zero was planted, and have a known floor. It does not ask the two to be equal, and @sec-app-validate reports the separate check in which the components are planted directly in a representation, where they are.

## Results {#sec-exp1-results}

```{r}
#| label: fig-toy
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 2.5
#| out-width: "100%"
#| fig-cap: "Experiment 1, all 7,560 cells. Left: the measured interaction against the amount planted in the language, with interquartile ranges. The linear condition cannot represent an interaction whatever the language contains, and reads zero throughout; the ReLU condition tracks what was planted. Right: overlap between the item and class effects against capacity, with the dashed line marking what a randomly oriented effect would give at that width. Below capacity the two effects are more orthogonal than chance; once dimension counting makes separation impossible they are several times chance. Note the log axis, and that the floor moves by a factor of eight across these widths."
ACT2 <- c(linear = UO[["green2"]], ReLU = UO[["gold"]])
# fill duplicates the colour legend, so it is suppressed rather than collected twice
pA <- ggplot(toy_track, aes(ach, size_int, colour = act, fill = act)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = .16, colour = NA) +
  geom_line(linewidth = .6) + geom_point(size = 1.5) +
  scale_colour_manual(values = ACT2, name = NULL) +
  scale_fill_manual(values = ACT2, guide = "none") +
  labs(x = "interaction planted in the language", y = "interaction measured\nin the representation") +
  theme_paper(8)

# one chance line, not one per condition: the two differ slightly because the class subspace
# rank is estimated, and drawing both makes a zigzag that reads as structure when it is not
floor_line <- toy_cap |> group_by(capacity) |> summarise(floor = median(floor), .groups = "drop")
pB <- ggplot(toy_cap, aes(capacity, overlap, colour = act)) +
  geom_line(data = floor_line, aes(y = floor), inherit.aes = FALSE,
            mapping = aes(capacity, floor), colour = UO[["grey"]], linetype = 2, linewidth = .5) +
  geom_line(linewidth = .6) + geom_point(size = 1.5) +
  annotate("text", x = 0.135, y = 0.032, label = "chance", size = 2.5,
           colour = UO[["grey"]], hjust = 0) +
  scale_colour_manual(values = ACT2, name = NULL) +
  scale_x_log10(breaks = c(.125, .25, .5, 1, 2)) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(x = "capacity (structure rank / hidden width)", y = "item-class overlap (log)") +
  theme_paper(8)

(pA | pB) + plot_layout(guides = "collect") &
  theme(legend.position = "top", legend.margin = margin(0, 0, -6, 0))
```

**The measure reports zero when the answer is zero.** The linear condition cannot represent an interaction, by the algebra of @sec-exp1-method, so every reading it gives is the estimator's own noise. Across all `r cm(nullget("linear", TRUE, "n"))` cells where the language *did* contain an interaction, it reports a median share of `r g(nullget("linear", TRUE, "med"), 4)` and flags one in `r sprintf("%.1f%%", 100 * nullget("linear", TRUE, "flagged"))` of them, which is the nominal rate for a test at the 5% level. Where no interaction was planted it flags `r sprintf("%.1f%%", 100 * nullget("linear", FALSE, "flagged"))`. That is the calibration, obtained without assuming anything about noise floors: a condition whose true value is provably zero, measured at every width, sample size and noise level in the grid.

**And it tracks what is there.** Under ReLU the measured interaction rises monotonically with the planted one (@fig-toy, left), and the rate at which it is flagged rises with it, from `r sprintf("%.0f%%", 100 * min(toy_track$flagged[toy_track$act == "ReLU" & toy_track$bin != "none"]))` at the smallest nonzero setting to `r sprintf("%.0f%%", 100 * max(toy_track$flagged[toy_track$act == "ReLU"]))` at the largest. With no interaction planted, ReLU flags `r sprintf("%.1f%%", 100 * nullget("ReLU", FALSE, "flagged"))` of cells, *below* nominal: the nonlinearity does not manufacture an interaction out of an additive language.

The measured shares are much smaller than the planted ones, and that is expected rather than a shortfall. The weights describe the target and the measure reads the representation, and a learner need only build enough structure to produce the answer. What matters is that the relationship is monotone and that zero maps to zero.

**Item and class effects entangle only when there is no room for them not to.** This is the substantive result, and it needs the chance level to be visible (@fig-toy, right; @tbl-toy). With capacity well below one, ReLU's item-class overlap is `r g(capget("ReLU", 0.125, "overlap"), 4)` against a chance level of `r g(capget("ReLU", 0.125, "floor"), 4)`: the two effects are *more* orthogonal than two arbitrary directions would be, which is what a learner that has room and uses it looks like. At capacity 1 the overlap is `r g(capget("ReLU", 1, "overlap"), 3)` against a floor of `r g(capget("ReLU", 1, "floor"), 4)`, about `r g(capget("ReLU", 1, "ratio"), 1)` times chance, and at capacity 2 it is `r g(capget("ReLU", 2, "overlap"), 3)`. The linear condition stays at or below chance throughout.

That is superposition, in the setting where we can see it happen: when the number of directions the structure needs exceeds the number available, the components stop being orthogonal because they cannot all be. Nothing about the language changed across those cells, only the room to represent it.

```{r}
#| label: tbl-toy
#| tbl-cap: "Experiment 1, by capacity, over cells where all three effects were planted. Chance is the share of a randomly oriented effect that would fall in the class direction at that width, and the ratio is the observed overlap against it. Below one, a fully separable solution is reachable by dimension counting; above one it is not."
toy_cap |>
  transmute(condition = act, capacity = g(capacity, 3),
            `interaction share` = g(size_int, 3),
            `item-class overlap` = g(overlap, 4),
            chance = g(floor, 4),
            `ratio to chance` = g(ratio, 2)) |>
  arrange(condition, capacity) |>
  kbl(booktabs = TRUE, align = "lrrrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

**The measurement survives the amount of nuisance variation a corpus has.** When context accounts for more than 80% of the target, so that most of what varies within a cell has nothing to do with the item or the class, ReLU's measured interaction holds at `r g(toy_noise$size_int[toy_noise$act == "ReLU" & toy_noise$cbin == ">80%"], 3)` against `r g(toy_noise$size_int[toy_noise$act == "ReLU" & toy_noise$cbin == "<20%"], 3)` when context is minor, and the flag rate falls from `r sprintf("%.0f%%", 100 * toy_noise$flagged[toy_noise$act == "ReLU" & toy_noise$cbin == "<20%"])` to `r sprintf("%.0f%%", 100 * toy_noise$flagged[toy_noise$act == "ReLU" & toy_noise$cbin == ">80%"])`. Power degrades, as it must with twenty-four observations per cell, but the estimate does not drift.

Two things carry into Experiment 2. The instrument reads zero when the answer is zero and tracks the answer when it is not, at the sample size and noise level the language models will be measured at. And whether the item and class effects come out orthogonal is governed by whether the representation had room, which is a fact about capacity rather than about the language, so a finding of near-orthogonality in a wide model is uninformative on its own and has to be read against its own chance level.


# Experiment 2: Item and Category in Language Models {#sec-exp2}

Experiment 1 established that a learner fitting an interactive language cannot represent it as item plus class, and that a nonlinearity additionally makes the components non-orthogonal. Experiment 2 applies the same decomposition to the contextual representations of pretrained language models, across three linguistic distinctions chosen to span grammar and meaning.

## Dataset {#sec-exp2-data}

**Corpora.** For part of speech and grammatical role we concatenate six general-domain English Universal Dependencies treebanks [@nivre2020ud]: EWT, GUM, LinES, ParTUT, GENTLE and PUD, giving `r cm(41095)` sentences and `r cm(706323)` tokens. Concatenating rather than using EWT alone is what makes the analysis possible: a lemma must be attested in both class levels often enough to estimate a cell mean, and a single treebank yields too few such lemmas. For metaphor we use the VUA20 corpus [@leong2020vua; @steen2010vua], restricted to content words, which supplies `r cm(gs("metaphor", "n_pre_grid_tokens"))` labeled tokens.

**Constructions.** All three are **same-token**: the item's surface form is identical across the two class levels, so the representation is read from the same token in both, and no part of any difference can come from tokenisation.

| distinction | item | levels | eligible | analyzed | tokens |
|:-----------|:-----------|:-----------|-----------:|-----------:|-----------:|
| part of speech | lemma | [noun]{.smallcaps} / [verb]{.smallcaps} | `r cm(gs("pos_noun_verb", "n_items_eligible"))` | `r gs("pos_noun_verb", "n_items_kept")` | `r cm(gs("pos_noun_verb", "n_grid_tokens"))` |
| grammatical role | surface form | *nsubj* / *obj* | --- | 49 | `r cm(2150)` |
| metaphor | surface form | literal / figurative | `r cm(gs("metaphor", "n_items_eligible"))` | `r gs("metaphor", "n_items_kept")` | `r cm(gs("metaphor", "n_grid_tokens"))` |

: The three constructions. *Eligible* counts items attested at both levels anywhere in the corpus; *analyzed* counts those attested often enough at each level to estimate a cell mean. {#tbl-constructions}

**Identifiability.** The interaction $\gamma(i,c)$ is only estimable for an item observed at more than one level, since it is defined by how an item's representation shifts across levels. We additionally require at least 10 tokens per cell, which is what allows a cell mean to be estimated and then split in half for the replication the measure needs (@sec-exp1-method). This is a real restriction on scope: for part of speech it selects *convertible* lemmas, words used both as a noun and as a verb, such as *run* or *hit*. Role and metaphor carry no such restriction, since any noun occurs as both subject and object and any content word occurs both literally and figuratively, which is one reason they broaden the claim. The median cell holds `r gs("pos_noun_verb", "tokens_per_cell_median")` tokens for part of speech and `r gs("metaphor", "tokens_per_cell_median")` for metaphor.

For role and metaphor the item is keyed on the **surface form** rather than the lemma, which guarantees the identical token across levels: a lemma like *thing* can differ in number across the two roles, whereas *things*-as-subject and *things*-as-object are the same string.

**Models.** Two families at three matched scales: OPT-BabyLM, trained on developmentally plausible quantities of child-available language [@warstadt2023babylm], and Pythia, trained on the Pile [@biderman2023pythia], at 125M/160M, 350M/410M and 1.3B/1.4B parameters, with model widths of 768, 1024 and 2048.

## Method {#sec-exp2-method}

The decomposition, the two reported quantities and the replication logic are exactly those of @sec-exp1-method. Three things differ.

**Representations.** For every token we take the hidden state of the word's **last subword** at every layer, from the embeddings through the final block, aligning subwords to corpus tokens with the tokenizer's word alignment. Sentences are drawn in a fixed random order and processed until roughly 300K tokens are extracted per model, so all models see the same material. Representations are extracted once and every analysis runs offline from them.

**Where the two independent estimates come from.** Experiment 1 had to train the model twice, because a toy lexeme has a single representation and therefore no within-cell replication. A language model supplies replication for free: each cell is a word used at one level across many sentences. We therefore split each cell's **token contexts** into two disjoint halves and estimate the grid from each. What cancels is context noise, the variation in a word's hidden state across the sentences it appears in. Because both halves come from the same frozen model they already share a coordinate system, so the rotation step of @sec-exp1-method is unnecessary here.

**Standardization.** Transformer hidden states contain a few *rogue* dimensions whose variance exceeds the rest by two to three orders of magnitude [@timkey2021bark], and these would otherwise dominate any subspace estimated from the raw scatter. We divide each dimension by its standard deviation before decomposing. This is a diagonal rescaling, so unlike whitening it needs no favorable ratio of samples to dimensions; the measure is accordingly invariant to per-dimension rescaling but not to arbitrary linear reparameterization, which we return to in @sec-app-math.

## Results {#sec-exp2-results}

```{r}
#| label: fig-llm
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 2.4
#| out-width: "100%"
#| fig-cap: "Experiment 2. Left: the interaction's share of representational structure, by relative depth (0 = static embeddings, 1 = final block). Right: the class effect's share on the same axis. Each line is one model; color distinguishes the two training corpora. Both components are built by depth, and the interaction is the larger of the two throughout."
plt <- llm |> filter(construction != "POS (noun/verb)") |>
  select(construction, family, size, depth, size_int, size_class) |>
  pivot_longer(c(size_int, size_class), names_to = "component", values_to = "share") |>
  mutate(component = factor(component, c("size_int", "size_class"),
                            c("interaction", "class effect")))
ggplot(plt, aes(depth, share, colour = family, group = interaction(family, size))) +
  geom_line(linewidth = .55, alpha = .85) +
  facet_grid(component ~ construction, scales = "free_y", switch = "y") +
  scale_colour_manual(values = c(BabyLM = UO[["gold"]], Pythia = UO[["blue"]]), name = NULL) +
  scale_x_continuous(breaks = c(0, .5, 1)) +
  labs(x = "relative depth", y = NULL) +
  theme_paper(8) + theme(strip.placement = "outside")
```

```{r}
#| label: tbl-llm
#| tbl-cap: "Experiment 2, at each model's deepest layer. Item-class overlap is the fraction of the item effect lying along the class direction, so 0.004 means that projecting out the class leaves 99.6% of item identity intact. The portable fraction is the class effect's share of the class-relevant structure, `size_class / (size_class + size_interaction)`: it asks how much of what distinguishes the two class levels is a direction shared across words, rather than specific to each word."
llm |> filter(construction != "POS (noun/verb)") |> group_by(construction, model) |>
  filter(layer == max(layer)) |> ungroup() |>
  mutate(portable = size_class / (size_class + size_int)) |>
  group_by(construction) |>
  summarise(`item-class overlap` = {v <- median(leak, na.rm = TRUE)
               if (is.na(v)) "---" else sprintf("%.3f", v)},
            `class effect` = sprintf("%.3f", median(size_class)),
            `interaction` = sprintf("%.3f", median(size_int)),
            `portable fraction` = sprintf("%.2f [%.2f, %.2f]", median(portable),
                                          min(portable), max(portable)),
            .groups = "drop") |>
  rename(distinction = construction) |>
  kbl(booktabs = TRUE, align = "lrrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

**Where the two components both exist, they sit on separate axes.** For grammatical role, item-class overlap at the deepest layer runs `r g(0.0025, 4)` to `r g(0.0062, 4)`, so projecting out the class direction leaves better than 99% of the item effect intact. These representations do not achieve class structure by borrowing the directions that carry word identity. This is the one place the answer is cleanly "separable," and it is worth saying plainly rather than straining against.

For metaphor the question does not arise. The overlap is not reported at **every** layer of every model, because the class effect never rises above the level at which an orientation is worth reporting. That is the measure declining to describe the direction of something that is barely there, and it is itself the first indication of how little of metaphoricity is carried by a shared direction.

**A substantial interaction is there, and depth builds it.** In both distinctions the interaction is
essentially absent at layer 0 and rises monotonically with depth (@fig-llm, top row). Because the
item is keyed on the surface form, the token is *identical* at both class levels, so the static
embedding layer has nothing to distinguish them with, and it does not. Everything the measure finds
is therefore created by contextualization rather than inherited from the input. This is the component
that breaks separability: it is neither a fact about the word nor a fact about the class, so no
decomposition into those two can absorb it.

**So does the class effect, and that matters for the framing.** The lower row of @fig-llm shows
the class effect growing with depth as well. Models are not choosing between abstraction and
entanglement; contextualization builds both at once. A description of these representations as either
"abstract" or "exemplar-like" is therefore underdetermined by the data, and the useful question is not
which one they are but what the balance is.

**The balance is graded across distinctions.** @tbl-llm gives that balance as the *portable fraction*:
of the structure that distinguishes the two class levels, how much is a direction shared across
words rather than specific to each word. Grammatical role is the most portable, and metaphor by far
the least, with only about a fourteenth of its class-relevant structure carried by a shared
direction. Metaphoricity is close to being encoded word by word: what makes a use of *plant*
figurative is largely not what makes a use of *course* figurative.

# Experiment 3: Does the Interaction Carry the Category? {#sec-exp3}

```{=html}
<!-- Method: hold the item fixed, project out the shared class direction, decode the class
     from the residual with a cross-validated probe against a within-item permutation null,
     plus the three metaphor probes. Awaiting llm_metaphor_probes.csv and the decode re-run. -->
```

# Discussion {#sec-discussion}

::: {.content-visible when-format="acl-pdf"}
<!-- PLACEHOLDER: author to write. -->
:::

# Limitations {.unnumbered}

<!-- PLACEHOLDER -->

::: {=latex}
\appendix
\floatplacement{figure}{H}
\floatplacement{table}{H}
\raggedbottom
:::

# The Decomposition and Its Estimator {#sec-app-math}

```{=html}
<!-- Formal development: the tensor-product projection view, the exact size partition and
     its invariance to grid size, item-class overlap against its r/d null, the positive bias of a
     squared projection norm, and the unbiasedness of the cross-estimate inner product.
     Drawn from MATH.md. -->
```
