---
title: "Are Lexical Representations Separable? Item and Category in Learned Distributed Representations"
author:
  - name: "Anonymous"
    affiliations:
      - name: "Anonymous Institution"
    email: anon@example.edu
abstract: |
  PLACEHOLDER --- author to write.
bibliography: refs.bib
acl-mode: review
number-sections: true
tbl-cap-location: bottom
format:
  acl-pdf:
    keep-tex: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr); library(dplyr); library(tidyr); library(ggplot2); library(kableExtra)
library(patchwork); library(scales)

DATA <- file.path("..", "data")
rd <- function(f) suppressMessages(read_csv(file.path(DATA, f), show_col_types = FALSE))

# UO palette, as in the companion papers. Green = the linear reference, gold = the
# nonlinearity; amber and blue are held for the model families in Experiments 2-3.
UO <- c(green = "#154733", green2 = "#007030", gold = "#F0B323",
        amber = "#875E00", blue = "#005581", grey = "#6E6E6E")
ACT_COL <- c(identity = UO[["green2"]], relu = UO[["gold"]])
ACT_LAB <- c(identity = "linear", relu = "ReLU")

theme_paper <- function(base = 9) {
  theme_minimal(base_size = base) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, colour = "grey88"),
          strip.text = element_text(face = "bold", size = base),
          legend.position = "top", legend.margin = margin(0, 0, -4, 0),
          legend.key.height = unit(9, "pt"),
          axis.title = element_text(size = base))
}

# ---- Experiment 1: the toy grid, cross-denoised (two initialisations per cell) ----
# ALL 4800 cells are used. Capacity enters as a reported variable, not as a filter: whether
# the representation has room to hold every generative direction on its own axis is part of
# the result, so it is shown rather than assumed.
toy <- rd("experiment9_cross_results.csv") |>
  mutate(band = cut(cap, c(-Inf, .5, 1, Inf),
                    labels = c("room to spare\n(rank/d < 0.5)", "tight\n(0.5-1)",
                               "impossible\n(> 1)")),
         act = factor(activation, c("identity", "relu"), c("linear", "ReLU")))

# Separability fails in two distinguishable ways, and the toy separates them: the presence
# of an interaction term at all, and non-orthogonality among the components. `size_int` is
# the first; `leak_marg` (item marginal inside the category subspace) and `ov_item` (the
# interaction's subspace overlapping the item subspace) are the second.
toy_dose <- toy |>
  group_by(int_frac, act) |>
  summarise(n = n(),
            size_med = median(size_interaction),
            size_lo = quantile(size_interaction, .25),
            size_hi = quantile(size_interaction, .75),
            sig = mean(sig_interaction),
            leak_med = median(leak_item_into_class, na.rm = TRUE),
            leak_n = sum(!is.na(leak_item_into_class)),
            ov_med = median(overlap_item_int, na.rm = TRUE),
            ov_n = sum(!is.na(overlap_item_int)),
            .groups = "drop")

# Interaction geometry is only interpretable where an interaction exists AND the marginals
# do: at int_frac = 0 there is no gamma to orient and at int_frac = 1 no marginals to orient
# against. The gate already withholds those, and they are excluded from the geometry panels.
TOY_MID <- c(.25, .5, .75)
toy_cap <- toy |> filter(int_frac %in% TOY_MID) |>
  group_by(act, band) |>
  summarise(n = n(), size_med = median(size_interaction),
            leak_med = median(leak_item_into_class, na.rm = TRUE),
            leakint_med = median(leak_int_into_margins, na.rm = TRUE),
            ov_med = median(overlap_item_int, na.rm = TRUE), .groups = "drop")

toy_null <- toy |> filter(int_frac == 0) |>
  group_by(act, band) |>
  summarise(n = n(), sig_rate = mean(sig_interaction),
            size_med = median(size_interaction), .groups = "drop")

g <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
pick <- function(tb, a, ifrac, col) tb[[col]][tb$act == a & tb$int_frac == ifrac]
capget <- function(tb, a, b, col) tb[[col]][tb$act == a & grepl(b, tb$band)]
```

# Introduction

::: {.content-visible when-format="acl-pdf"}
<!-- PLACEHOLDER: author to write. Frames the paper as an extension of Exemplars in
     Disguise: if a learner's behaviour cannot distinguish exemplar storage from
     abstraction, ask instead whether its representations separate into an item part and
     a category part. Ends with the Present Study paragraph setting up the three
     experiments. -->
:::

# Experiment 1: Item and Category in a Hybrid Learner {#sec-exp1}

Before asking whether a language model's representations separate lexical identity from
category, we need a setting where the right answer is known. Experiment 1 trains a learner
on an artificial language whose generative structure we control, so that we can vary how
much of the language is item-by-category specific and check what the measure recovers.

The learner is deliberately a hybrid of the two accounts at issue. Every word-in-category
pairing receives its own private embedding, with no parameter sharing between them, which
is exemplar storage; those embeddings are then read out through a single network shared by
every pairing, which is abstraction. This is also the architecture of the models in
Experiment 2, where private token embeddings feed shared transformer blocks. The question
is what such a learner does with structure that *could* be represented separably.

## Dataset {#sec-exp1-data}

The artificial language is a next-token prediction task over a vocabulary of
`r max(toy$n_lex) / 16` word types. Each **lexeme** is a (form, category) pair: 250 forms
crossed with 16 categories, giving 4,000 lexemes, each with its own target distribution
over the vocabulary. Forms are the items and categories are the classes, so the design is
the same item-by-category layout used throughout the paper.

Each form $i$ and category $c$ carries a latent code, and the target distribution is
generated from those codes through independently drawn loadings onto the vocabulary:

$$
\text{logits}(i,c) \;=\; s\Big[\sqrt{1-f}\;\overline{m}(i,c) \;+\; \sqrt{f}\;\overline{x}(i,c)\Big]
$$ {#eq-toy}

where $\overline{m}$ is an **additive** term, the sum of a form contribution and a category
contribution, and $\overline{x}$ is a **bilinear** term formed from the elementwise product
of the two codes, so it is irreducibly joint. Both are normalised to unit per-element
standard deviation before mixing, and the $\sqrt{1-f}, \sqrt{f}$ weighting holds the total
signal scale fixed, so the dial $f$ changes the *composition* of the language without
changing how hard it is to learn.

That dial, $f \in \{0, .25, .5, .75, 1\}$, is the manipulation. At $f = 0$ a word's
behaviour is exactly its form contribution plus its category contribution, with **no
interaction whatever**. At $f = 1$ behaviour is purely the form-by-category conjunction and
the marginals vanish. Intermediate values mix the two.

Two properties of the construction matter for what can be concluded. First, the three sets
of loadings are drawn independently, so the additive and the interactive structure occupy
approximately orthogonal subspaces of the output space: **a fully separable representation
exists** for every value of $f$. Second, the learner's hidden layer is at least as wide as
the generative rank in the cells we analyse, so it has room to place each component on its
own axis. Any entanglement the measure reports is therefore something the learner did, not
something the task or the architecture forced.

We cross the dial with the generative ranks of the form code ($r_i \in \{1,2,4,8,16,32\}$),
the category code ($r_c \in \{1,2,4,8\}$), the hidden width ($d \in \{16,32,64,128\}$), the
output nonlinearity (linear or ReLU) and five random draws of the language, for
`r format(nrow(toy), big.mark = ",")` cells in total, all of which are analysed. The ratio of
generative rank to hidden width, which we call **capacity**, ranges across the grid from
`r g(min(toy$cap), 2)` to `r g(max(toy$cap), 1)`. Below one a fully separable solution is
reachable by dimension counting and above one it is not, so capacity is reported as a
variable rather than used to select cells.

## Method {#sec-exp1-method}

### The learner

Each lexeme has a free $d$-dimensional embedding, initialised at random and trained without
any sharing across lexemes: nothing ties *dog*-as-noun to *dog*-as-verb. Those embeddings
pass through one hidden layer of width $d$ with either an identity or a ReLU nonlinearity,
and then a linear map to the vocabulary. We measure the hidden layer.

The identity condition is the reference case. A linear map cannot destroy linear
recoverability, so whatever the linear learner does to the geometry is a rotation and
shear of separable structure rather than a fusion of it. Differences between the two
conditions are attributable to the nonlinearity.

Training minimises the exact expected cross-entropy against the true target distribution
rather than a sampled-token estimate, so a cell is deterministic given its language and
initialisation, and a seed varies only the language and the starting weights. Cells are
trained to a loss plateau.

### Decomposing a representation {#sec-measure}

Let $M[i,c] \in \mathbb{R}^d$ be the mean representation of item $i$ used at category level
$c$. We split that grid into four geometric components,

$$
M[i,c] \;=\; \mu + \alpha(i) + \beta(c) + \gamma(i,c),
$$ {#eq-decomp}

by centring: $\mu$ is the grand mean, $\alpha(i)$ is item $i$'s row mean minus $\mu$,
$\beta(c)$ is level $c$'s column mean minus $\mu$, and $\gamma(i,c)$ is what remains. The
familiar reading is the main-effects-plus-interaction partition, with the difference that
the dependent variable is a vector, so each component is a *direction* in representational
space rather than a number. In mixed-model terms $\alpha$ is a by-item random intercept,
$\beta$ is the fixed effect of category, and $\gamma$ is the by-item random slope for
category; the question of this paper is how much of the category effect is carried by
$\beta$ rather than by $\gamma$. We use that mapping for intuition only. Nothing is fitted,
there is no shrinkage, and no distribution is assumed over the by-item terms.

A representation is **separable** if a word's representation is the sum of an item part and
a category part, with the two on their own axes. It can fail in two distinguishable ways,
and we measure them separately. It fails if the components are **non-orthogonal**, so that
projecting out the category also removes item information; and it fails if there is a
**third component**, an item-by-category term that is neither an item part nor a category
part, whatever its orientation. A representation with a large $\gamma$ does not factor into
item plus category even if $\gamma$ sits on an axis of its own.

Because the four components are mutually orthogonal *as sources*, their squared lengths
partition the structure of the grid exactly, and we report each as a share of the total.
**Interaction size** is $\gamma$'s share, which measures the second kind of failure. For the
first we report two projections: **marginal leakage**, the fraction of the item component's
energy lying inside the category component's subspace, and **interaction overlap**, how much
the interaction's subspace shares directions with the item subspace. Zero on both, with a
negligible $\gamma$, is a cleanly separable representation.

**Interpreting the interaction.** Because $\gamma$ is obtained by subtraction it invites
reading as a residual in the regression sense, the part a model failed to capture. It is
not. The decomposition is saturated: $\mu$, $\alpha$, $\beta$ and $\gamma$ exhaust the
degrees of freedom of the grid, so $\gamma$ is the final term of a complete description
rather than the leftover of an incomplete one. The residual stratum lies one level below,
in the scatter of individual observations *within* each cell. That distinction is not free,
however, since cell means are estimated from finitely many observations and within-cell
error propagates upward into $\gamma$: a squared length is bounded below by its own noise
floor and cannot read zero even when the true interaction is zero. This is the situation in
which an analysis of variance tests an interaction against the within-cell error term, and
it requires replication for the same reason.

**Replication and the null.** We therefore estimate the grid **twice** from independent
data and keep only what the two estimates agree on, reporting the signed inner product
$\langle \gamma_A, \gamma_B \rangle$ rather than $\lVert\gamma\rVert^2$. Independent errors
cancel in expectation, so this is unbiased at any noise level; and because it can be
negative it supports a genuine test, which a squared length cannot. Significance comes from
permuting which item of one estimate is paired with which item of the other, which preserves
every magnitude while destroying the correspondence. A component's *direction* is only
defined if the component exists, so leakage is reported only when both components it relates
are present; otherwise it is withheld.

In Experiment 1 the two estimates come from **two initialisations trained on the same
language**. This is necessary rather than incidental: the toy has one representation per
cell and hence no within-cell replication, so interaction and error are not identifiable, in
exactly the way they are not identifiable in an unreplicated two-way design. Training twice
supplies the missing replication, and what it cancels is the initialisation-dependent
content that gradient descent leaves in the directions the loss does not constrain. Because
two independently trained networks occupy different coordinate systems, one is rotated onto
the other before comparison, with the rotation fitted on held-out items so that an item's
own noise never shapes its own alignment. @sec-app-math gives the algebra.

## Results {#sec-exp1-results}

```{r}
#| label: fig-toy
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 2.3
#| out-width: "100%"
#| fig-cap: "Experiment 1, all 4,800 cells. Left: the interaction's share of representational structure. Middle: leakage of the item component into the category subspace, on a log axis (it spans four orders of magnitude). Right: how much the interaction's subspace overlaps the item subspace. Points are medians. Both readouts build a substantial interaction, and it tracks the generative dial; only the ReLU readout also makes the components non-orthogonal. The two outer panels are shown for the settings where the quantities are defined: an orientation requires the components it relates to exist."
mid <- toy_dose |> filter(int_frac %in% TOY_MID)
pan <- function(dat, y, ylab, logy = FALSE) {
  p <- ggplot(dat, aes(int_frac, .data[[y]], colour = act)) +
    geom_line(linewidth = .6) + geom_point(size = 1.5) +
    scale_colour_manual(values = c(linear = UO[["green2"]], ReLU = UO[["gold"]]), name = NULL) +
    scale_x_continuous(breaks = c(0, .5, 1)) +
    labs(x = NULL, y = ylab) + theme_paper(8)
  if (logy) p + scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) else p
}
((pan(toy_dose, "size_med", "interaction share\nof structure") |
  pan(filter(toy_dose, int_frac > 0), "leak_med", "item leaks into\ncategory (log)", TRUE) |
  pan(mid, "ov_med", "interaction overlaps\nthe item subspace")) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top", legend.margin = margin(0, 0, -6, 0)))
```

**The measure tracks what the language contains.** With a purely additive language the
estimator reports a median interaction share of
`r g(toy_dose$size_med[toy_dose$act == "linear" & toy_dose$int_frac == 0], 4)` under a linear
readout; with a purely interactive one it reports
`r g(pick(toy_dose, "linear", 1, "size_med"), 3)`. The instrument exercises nearly its full
range on a task whose answer is known at both ends (@fig-toy, left).

**Even the linear learner builds an interaction.** At every intermediate setting the linear
readout carries an item-by-category component of
`r g(pick(toy_dose, "linear", .25, "size_med"), 2)` to
`r g(pick(toy_dose, "linear", .75, "size_med"), 2)` of representational structure, flagged as
present in **100%** of cells. This is not optional. A linear readout maps the hidden layer to
the logits, and a linear map of a purely additive representation can only produce additive
logits; so any learner that actually fits an interactive language must carry a non-additive
representation. The interaction is forced by the task, not chosen by the architecture.

**The linear learner nonetheless keeps everything orthogonal.** Its item and category
components stay on essentially separate axes, with leakage from
`r g(pick(toy_dose, "linear", .25, "leak_med"), 5)` to
`r g(pick(toy_dose, "linear", .75, "leak_med"), 4)`, and its interaction occupies its own
axis, overlapping the item subspace at
`r g(pick(toy_dose, "linear", .25, "ov_med"), 3)` and
`r g(pick(toy_dose, "linear", .5, "ov_med"), 3)` at the two lower settings. So the linear
representation fails separability in exactly one of the two senses: it contains a third
component that is neither item nor category, while the components it does contain are
mutually orthogonal.

**The nonlinearity compounds it, on both counts.** A ReLU readout carries an interaction of
similar or slightly larger size (`r g(pick(toy_dose, "ReLU", .75, "size_med"), 2)` at the
highest intermediate setting), but now the components are also non-orthogonal. Marginal
leakage runs `r g(pick(toy_dose, "ReLU", .25, "leak_med"), 3)` to
`r g(pick(toy_dose, "ReLU", .75, "leak_med"), 3)`, two orders of magnitude above the linear
readout at matched settings, and the interaction overlaps the item subspace at
`r g(pick(toy_dose, "ReLU", .25, "ov_med"), 2)` to
`r g(pick(toy_dose, "ReLU", .75, "ov_med"), 2)` where the linear readout sits near zero.
Under a nonlinearity the interaction stops being a separate module and is superposed onto the
code that carries item and category identity.

**Capacity governs whether the entanglement was avoidable** (@tbl-toy). Where the hidden
layer has room for every generative direction, the linear readout's interaction overlaps the
item subspace at `r g(capget(toy_cap, "linear", "room", "ov_med"), 3)`: separation was
achievable and it was achieved. Where dimension counting makes separation impossible, the
same linear readout reaches `r g(capget(toy_cap, "linear", "impossible", "ov_med"), 2)`. The
ReLU readout is entangled at `r g(capget(toy_cap, "ReLU", "room", "ov_med"), 2)` even with
room to spare, so its entanglement is learned rather than forced. This is also why the
additive control must be read by capacity: with an additive language and room to spare the
linear readout flags an interaction in
`r sprintf("%.0f%%", 100 * capget(toy_null, "linear", "room", "sig_rate"))` of cells, but
where separation is impossible it flags one in
`r sprintf("%.0f%%", 100 * capget(toy_null, "linear", "impossible", "sig_rate"))`, correctly,
since a representation with nowhere to put the components cannot keep them apart.

```{r}
#| label: tbl-toy
#| tbl-cap: "Experiment 1, by capacity. Medians over the intermediate settings of the dial, where both the interaction and the marginals exist. Capacity is the generative rank over the hidden width: below one, a fully separable solution is reachable. Leakage and overlap are zero for a separable representation."
toy_cap |>
  transmute(readout = act,
            capacity = gsub("\n", " ", band),
            `interaction share` = g(size_med, 3),
            `item into category` = g(leak_med, 4),
            `interaction into margins` = g(leakint_med, 3),
            `interaction / item overlap` = g(ov_med, 3)) |>
  arrange(readout, capacity) |>
  kbl(booktabs = TRUE, align = "llrrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

Two conclusions carry into Experiment 2. Separability fails for a **principled** reason and
not an accidental one: no learner that fits an interactive language can represent it as item
plus category, because the readout would then be unable to produce the interaction the
language requires. And the *further* failure, the components ceasing to be orthogonal, is
specifically what a nonlinearity adds. Real language models are nonlinear and real language
is interactive, so both mechanisms are in play before we look.

# Experiment 2: Item and Category in Language Models {#sec-exp2}

<!-- Dataset: six UD English treebanks concatenated, plus VUA20 for metaphor; the four
     constructions and the identifiability criteria; corpus and balanced-grid Ns from
     data/methods_grid_stats.csv. Method: differences from Experiment 1 only, chiefly that
     the two independent estimates are two halves of each cell's token contexts rather than
     two initialisations, so no rotation is needed. Awaiting methods_grid_stats.csv. -->

# Experiment 3: Does the Interaction Carry the Category? {#sec-exp3}

<!-- Method: hold the item fixed, project out the shared category axis, decode the category
     from the residual with a cross-validated probe against a within-item permutation null.
     Awaiting the re-run of decode_from_interaction.py for the bootstrap intervals. -->

# Discussion {#sec-discussion}

::: {.content-visible when-format="acl-pdf"}
<!-- PLACEHOLDER: author to write. -->
:::

# Limitations {.unnumbered}

<!-- PLACEHOLDER -->

::: {=latex}
\appendix
\floatplacement{figure}{H}
\floatplacement{table}{H}
\raggedbottom
:::

# The Decomposition and Its Estimator {#sec-app-math}

<!-- Formal development: the tensor-product projection view, the exact size partition and
     its invariance to grid size, leakage against its r/d null, the positive bias of a
     squared projection norm, and the unbiasedness of the cross-estimate inner product.
     Drawn from MATH.md. -->
