---
title: "Are Lexical Representations Separable? Item and Category in Learned Distributed Representations"
author:
  - name: "Anonymous"
    affiliations:
      - name: "Anonymous Institution"
    email: anon@example.edu
abstract: |
  PLACEHOLDER --- author to write.
bibliography: refs.bib
acl-mode: review
number-sections: true
tbl-cap-location: bottom
format:
  acl-pdf:
    keep-tex: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
# Force kable to emit raw LaTeX rather than markdown pipe tables. Without it pandoc builds the
# longtable itself and puts \bottomrule in \endlastfoot, which longtable typesets after the body,
# so the rule lands below the caption whenever tbl-cap-location is bottom.
if (knitr::is_latex_output()) options(knitr.table.format = "latex")
library(readr); library(dplyr); library(tidyr); library(ggplot2); library(kableExtra)
library(patchwork); library(scales)

DATA <- file.path("..", "data")
# Reads a result CSV, falling back to the gzipped copy. validate_measure is 252,000 rows and is
# tracked compressed, with the plain file gitignored, so a fresh clone has only the .gz; readr
# decompresses transparently. Handled here rather than at the call site so any file can be
# compressed later without the paper needing to know.
rd <- function(f) {
  p <- file.path(DATA, f)
  if (!file.exists(p) && file.exists(paste0(p, ".gz"))) p <- paste0(p, ".gz")
  suppressMessages(read_csv(p, show_col_types = FALSE))
}

# Fill in columns a CSV predates, as NA, so a file measured before a field existed still loads
# and the affected numbers read as missing rather than bringing the whole render down.
ensure <- function(df, ...) {
  for (nm in c(...)) if (!nm %in% names(df)) df[[nm]] <- NA_real_
  df
}

# Green and gold are the University of Oregon colours, as in the companion papers; grey,
# amber and blue are supporting tones, not part of that palette. Green = the linear reference, gold = the
# nonlinearity; amber and blue are held for the model families in Experiments 2-3.
UO <- c(green = "#154733", green2 = "#007030", gold = "#F0B323",
        amber = "#875E00", blue = "#005581", grey = "#6E6E6E")
ACT_COL <- c(identity = UO[["green2"]], relu = UO[["gold"]])
ACT_LAB <- c(identity = "linear", relu = "ReLU")

theme_paper <- function(base = 9) {
  theme_minimal(base_size = base) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, color = "grey88"),
          strip.text = element_text(face = "bold", size = base),
          legend.position = "top", legend.margin = margin(0, 0, -4, 0),
          legend.key.height = unit(9, "pt"),
          axis.title = element_text(size = base))
}

# ---- Experiment 1: the artificial-language grid ----
# 7,560 runs, measured with unified_split -- the SAME function and arguments Experiment 2
# uses. The two independent estimates are two even-sized splits of the observations for each item in each class, so there is
# no alignment step and no projection. Capacity is a reported variable, never a filter.
toy <- rd("artificial_language_grid.csv") |>
  ensure("obs_size_interaction", "size_interaction_excludes_zero",
         "obs_size_class", "size_class_excludes_zero",
         "size_item_excludes_zero") |>
  mutate(act = factor(activation, c("identity", "relu"), c("linear", "ReLU")),
         # An orthogonality reading means nothing on its own, so each overlap is carried
         # alongside the upper tail of its own null: the largest value an arbitrary orientation
         # produces in 97.5% of draws, from rotating the target subspace 200 times.
         null_hi = leak_item_into_class_null_hi,
         null_lo = leak_item_into_class_null_lo,
         null_med = leak_item_into_class_null_med,
         null_int = leak_int_into_margins_null_hi,
         null_int_lo = leak_int_into_margins_null_lo,
          null_int_med = leak_int_into_margins_null_med,
          # `ez` is the test the results are quoted with: the re-split 95% interval excludes
          # zero. `flagged` below keeps the older permutation test for continuity only.
          ez = size_interaction_excludes_zero)

# Where no interaction was planted the linear arm is the floor, and by algebra it should be
# one: with additive inputs a linear map gives population means of the form A(f)+B(c), whose
# interaction is zero however much the language contains. Where an interaction WAS planted it
# is not at zero, because a trained linear model still carries whatever its contexts bring to
# each cell -- see the Results. Both readings are the measure's at this exact width, sample
# size and context level.
toy_null <- toy |> group_by(act, planted = w_int > 0) |>
  summarize(n = n(), med = median(size_interaction),
            p95 = quantile(size_interaction, .95),
            flagged = mean(sig_interaction),
            flagged_ez = mean(ez, na.rm = TRUE), .groups = "drop")

# Does the measured interaction track the planted one? Binned, because the achieved share is
# continuous (neither the softmax nor the normalization is linear in the weights).
toy_track <- toy |> filter(w_item > 0, w_class > 0) |>
  mutate(bin = cut(ach_int, c(-Inf, .005, .05, .15, .30, .55, Inf),
                   labels = c("none", ".005-.05", ".05-.15", ".15-.30", ".30-.55", ">.55"))) |>
  group_by(act, bin) |>
  summarize(n = n(), ach = median(ach_int), size_int = median(size_interaction),
            lo = quantile(size_interaction, .25), hi = quantile(size_interaction, .75),
            flagged = mean(sig_interaction),
            flagged_ez = mean(ez, na.rm = TRUE), .groups = "drop")

# ---- what the learner builds, by width --------------------------------------------------------
# A run contributes only if the measure found structure for the three shares to divide: at least
# one of the three components has a re-split interval excluding zero. Where none does, the shares
# divide noise, whichever component lands slightly positive takes the whole of it, and the value
# is arbitrary rather than small. Every run this removes is at the narrowest width, and how many
# there are is reported in the Results rather than passed over.
toy_relu <- toy |> filter(act == "ReLU", w_item > 0, w_class > 0) |>
  mutate(resolved = size_item_excludes_zero | size_class_excludes_zero |
                   size_interaction_excludes_zero)
toy_unbuilt <- toy_relu |> group_by(d) |>
  summarize(runs = n(), none = sum(!resolved), pct = none / n(), .groups = "drop")

# Long form, one row per component, so the three can be drawn on one pair of axes. Intervals are
# the 2.5th and 97.5th percentiles ACROSS RUNS: what the bar shows is how differently two runs of
# the same language at the same width come out, which is itself one of the findings.
toy_long <- toy_relu |> filter(resolved) |>
  select(d, ach_int, item = size_item, class = size_class, interaction = size_interaction) |>
  pivot_longer(c(item, class, interaction), names_to = "component", values_to = "share") |>
  mutate(component = factor(component, c("item", "class", "interaction")))

# One question asked three times: when the language holds a given amount of a component, how much
# of it does the learner build? Each component is matched against its OWN planted share, so the
# diagonal is a meaningful reference and the three panels are the same question about item, class
# and interaction rather than three different ones.
# The frame is `df`, not `d`: tibble() evaluates its arguments in order, so a column named `d`
# would shadow the frame for every argument after it.
.pair <- function(df, planted, measured, name) {
  # The planted share is taken over the three components only, leaving context out of the
  # denominator, because that is what the measured share divides too: context is averaged away in
  # forming the item-by-class means and cannot appear in them. Left unmatched the two axes divide
  # different things and the diagonal means nothing, since a faithful representation would sit
  # above it by however much context the language held. Taken this way both axes are the
  # component's share of the same three things and the diagonal is a faithful representation.
  # Dividing by the three rather than by one minus the context is also the safer arithmetic: the
  # four achieved shares sum to one only to within a percent, and this way the three always sum
  # to exactly one.
  tot <- df$ach_item + df$ach_class + df$ach_int
  ok <- !is.na(tot) & tot > 1e-3
  tibble(d = df$d[ok], component = name,
         planted = df[[planted]][ok] / tot[ok],
         measured = df[[measured]][ok])
}
.res <- toy_relu |> filter(resolved)
toy_learn <- bind_rows(
  .pair(.res, "ach_item", "size_item", "item"),
  .pair(.res, "ach_class", "size_class", "class"),
  .pair(.res, "ach_int", "size_interaction", "interaction")) |>
  mutate(component = factor(component, c("item", "class", "interaction")),
         bin = cut(planted, c(-Inf, .02, .06, .12, .22, .40, Inf))) |>
  filter(!is.na(planted), !is.na(measured))
# The binned summary stays for the prose; the figure draws the runs and a fit through them.
toy_learn_bin <- toy_learn |> group_by(component, d, bin) |>
  summarize(n = n(), x = median(planted), med = median(measured),
            lo = quantile(measured, .025), hi = quantile(measured, .975), .groups = "drop") |>
  filter(n >= 10)

# The linear arm, prepared the same way, for the comparison the prose makes. It is not drawn in
# @fig-toy, which would double the lines for one sentence.
.lin <- toy |> filter(act == "linear", w_item > 0, w_class > 0) |>
  mutate(resolved = size_item_excludes_zero | size_class_excludes_zero |
                      size_interaction_excludes_zero) |>
  filter(resolved)
toy_learn_lin <- bind_rows(
  .pair(.lin, "ach_item", "size_item", "item"),
  .pair(.lin, "ach_class", "size_class", "class"),
  .pair(.lin, "ach_int", "size_interaction", "interaction")) |>
  mutate(component = factor(component, c("item", "class", "interaction"))) |>
  filter(!is.na(planted), !is.na(measured))

# Colour carries width here, since the components are separated by panel. A single ramp from the
# University of Oregon's green, because width is ordered and hues are not.
WIDC <- c("8" = "#BFCFA8", "16" = "#8AA45F", "32" = "#4C6E36", "128" = UO[["green"]])

# The fitted slope of represented share against the language's share, which is the quantity the
# prose quotes: how much of an increase in the language reaches the representation. Below one
# means the component is represented at less than its rate in the language.
slope <- function(dd, comp, dat = toy_learn) {
  x <- dat[dat$d == dd & dat$component == comp, ]
  unname(coef(stats::lm(measured ~ planted, data = x))[2])
}
slope_lin <- function(dd) slope(dd, "interaction", toy_learn_lin)

# The linear arm's composition by width, for the contrast the prose draws.
toy_comp_lin <- toy_learn_lin |> group_by(d, component) |>
  summarize(med = median(measured), .groups = "drop")
cmplin <- function(dd, comp) toy_comp_lin$med[toy_comp_lin$d == dd & toy_comp_lin$component == comp]

# Reproducibility at a FIXED language. Within a cell of weights, context weight and width the
# only thing that varies is the seed, so the spread across those runs is run-to-run variability.
# Spread pooled over languages is not: it is dominated by the languages differing, which is the
# horizontal axis of @fig-toy showing up on the vertical one.
toy_seed <- toy_relu |> filter(resolved) |>
  group_by(w_item, w_class, w_int, w_ctx, d) |>
  filter(dplyr::n() >= 3) |>
  summarize(item = max(size_item) - min(size_item),
            class = max(size_class) - min(size_class),
            interaction = max(size_interaction) - min(size_interaction), .groups = "drop") |>
  group_by(d) |>
  summarize(item = median(item), class = median(class),
            interaction = median(interaction), .groups = "drop")
seedspan <- function(dd, comp) toy_seed[[comp]][toy_seed$d == dd]

# Reach into the per-component view: the measured share at the lowest or highest planted band.
lrn <- function(dd, comp, top = TRUE) {
  x <- toy_learn_bin[toy_learn_bin$d == dd & toy_learn_bin$component == comp, ]
  x <- x[order(x$x), ]
  x$med[if (top) nrow(x) else 1]
}

# Pooled composition by width, for the prose.
toy_comp <- toy_long |> group_by(d, component) |>
  summarize(n = n(), med = median(share),
            lo = quantile(share, .025), hi = quantile(share, .975), .groups = "drop")

# Pooled over languages, for the prose and for the width trend on its own.
toy_width <- toy_long |> group_by(d, component) |>
  summarize(n = n(), med = median(share), lo = quantile(share, .025),
            hi = quantile(share, .975), .groups = "drop")
cmp <- function(dd, comp, col = "med") {
  toy_width[[col]][toy_width$d == dd & toy_width$component == comp]
}
cmpw <- function(dd, comp) cmp(dd, comp, "hi") - cmp(dd, comp, "lo")
unbuilt <- function(dd, col = "pct") toy_unbuilt[[col]][toy_unbuilt$d == dd]

# How the other properties of the language move the composition. Reported in the appendix rather
# than folded into the figure, so nothing is collapsed without saying so.
toy_bylang <- bind_rows(
  toy_relu |> filter(resolved) |>
    mutate(property = "context in the language",
           level = cut(ach_ctx, c(-Inf, .3, .7, Inf),
                       labels = c("under 30%", "30 to 70%", "over 70%"))),
  toy_relu |> filter(resolved) |>
    mutate(property = "item against class",
           level = cut(w_item / pmax(w_class, 1e-9), c(-Inf, .95, 1.05, Inf),
                       labels = c("item < class", "item = class", "item > class")))) |>
  group_by(property, level) |>
  summarize(runs = n(), item = median(size_item), class = median(size_class),
            interaction = median(size_interaction), .groups = "drop")

# Capacity: below 1 every component can have its own axis, above 1 they cannot.
# One overlap for a table: the value, how many times its own null's median it is, and how many of
# the measurements fall outside that null's 95% interval. The multiple is what makes overlaps
# comparable across widths, since the null moves by orders of magnitude with d; the count is what
# says the effect holds rather than resting on one measurement. Both tables use this, so they
# report the same three things and differ only in what a row is and how many measurements it holds.
# Per-model renderings, for tables whose rows are one model rather than six. These are the
# quantities as measured, with no median taken across models: a value with its own re-split
# interval, and an overlap with its own null's median. Vectorized, so they work inside mutate.
ci <- function(m, lo, hi, digits = 3) {
  f <- paste0("%.", digits, "f")
  out <- sprintf(paste0(f, " [", f, ", ", f, "]"), m, lo, hi)
  out[is.na(m)] <- "---"
  out
}
ov1 <- function(v, med, digits = 3) {
  out <- sprintf(paste0("%.", digits, "f (%.1f×)"), v, v / med)
  out[is.na(v)] <- "---"
  out
}

ov <- function(v, med, lo, hi, digits = 3) {
  m <- median(v, na.rm = TRUE)
  if (is.na(m)) return("---")
  ratio <- m / median(med, na.rm = TRUE)
  out <- sum(!is.na(v) & (v > hi | v < lo))
  sprintf(paste0("%.", digits, "f  (%.1f×, %d/%d)"), m, ratio, out, sum(!is.na(v)))
}

# One size for a table: the median share, and how many measurements clear zero -- their re-split
# 95% interval excludes zero. All three sizes are cross-split dot products with a null of exactly
# zero, so there is no "x null median" multiple to report and no raw null band to read against;
# `ez` is the per-measurement excludes-zero flag.
# A size as the share, with an interval when one is supplied, and how many of the group's
# measurements have a re-split interval clear of zero.
#
# What belongs in the brackets depends on what the row pools. Where a row is six models, the
# re-split interval is the wrong interval to print: it is 35 to 110 times narrower than the
# spread across those models, so it would claim the quantity is pinned to a thousandth while the
# models themselves disagree in the second decimal. The across-model range is the honest spread
# at that level. The re-split
# interval still does its work through the count, which is what says each model is clear of zero.
sz <- function(share, ez, lo = NULL, hi = NULL, digits = 3) {
  m <- median(share, na.rm = TRUE)
  if (is.na(m)) return("---")
  f <- paste0("%.", digits, "f")
  base <- if (is.null(lo)) sprintf(f, m) else
    sprintf(paste0(f, " [", f, ", ", f, "]"), m, lo, hi)
  n <- sum(!is.na(ez))
  if (n == 0) return(base)                                   # intervals not measured yet
  k <- sum(ez %in% c(TRUE, "TRUE"), na.rm = TRUE)
  sprintf("%s (%d/%d)", base, k, n)
}

# "4 of 6", or "NA" before the intervals are measured, for use in prose. For a size, "clearing"
# means its re-split 95% interval excludes zero.
outof <- function(constr, stem) {
  v <- llm_deep[[paste0(stem, "_excludes_zero")]][llm_deep$construction == constr]
  if (!length(v) || all(is.na(v))) return("NA")
  sprintf("%d of %d", sum(v %in% c(TRUE, "TRUE")), sum(!is.na(v)))
}

toy_cap <- toy |> filter(w_int > 0, w_item > 0, w_class > 0) |>
  # Same inclusion rule as @fig-toy: a run contributes only if at least one component's re-split
  # interval excludes zero, since otherwise the three shares divide noise.
  filter(size_item_excludes_zero | size_class_excludes_zero | size_interaction_excludes_zero) |>
  group_by(act, capacity) |>
  # Formatted here rather than in the table, because `ov` needs the runs themselves to count how
  # many fall outside their null; a summarized row no longer has them.
  summarize(n = n(), size_int = median(size_interaction),
            # median with the 95% range over the runs in the row. A row pools hundreds of runs
            # that differ in seed and in what was planted, and the median alone said nothing
            # about how much they differ.
            sz_int = sz(size_interaction, size_interaction_excludes_zero,
                        quantile(size_interaction, .025), quantile(size_interaction, .975),
                        digits = 3),
            ov_item = ov(leak_item_into_class, null_med, null_lo, null_hi, 4),
            ov_int = ov(leak_int_into_margins, null_int_med, null_int_lo, null_int, 3),
            overlap = median(leak_item_into_class, na.rm = TRUE),
            floor = median(null_hi, na.rm = TRUE),
            ratio = median(leak_item_into_class / null_hi, na.rm = TRUE),
            .groups = "drop")

# Context is the nuisance dial: how much of the target is about something other than the item,
# the class and their conjunction. The high end is the regime Experiment 2 sits in.
toy_noise <- toy |> filter(w_int > 0, w_item > 0, w_class > 0) |>
  mutate(cbin = cut(ach_ctx, c(-Inf, .2, .5, .8, Inf),
                    labels = c("<20%", "20-50%", "50-80%", ">80%"))) |>
  group_by(act, cbin) |>
  summarize(n = n(), size_int = median(size_interaction),
            flagged = mean(sig_interaction),
            flagged_ez = mean(ez, na.rm = TRUE), .groups = "drop")

g <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
cm <- function(x) format(x, big.mark = ",")

# Short display names for the three constructions. The data keeps the long ones, which the prose
# and the inline expressions index by.
SHORT <- c("POS (noun/verb)"               = "part of speech",
           "role (subject/object)"         = "grammatical role",
           "metaphor (literal/figurative)" = "metaphor")
shortname <- function(x) unname(SHORT[as.character(x)])
nullget <- function(a, pl, col) toy_null[[col]][toy_null$act == a & toy_null$planted == pl]
capget <- function(a, cap, col) toy_cap[[col]][toy_cap$act == a & toy_cap$capacity == cap]

# ---- Experiment 2: language model representations ----
# Per-class token medians, from the per-item stimulus list rather than the pooled summary. A
# single pooled median cannot say whether a form occurs that often in EACH class or that often in
# total; both are reported instead, in the order of the contrast. The counts are corpus-level and
# identical across all six models, so one model gives the figure.
stim <- rd("stimuli_items.csv")
# The span of those per-class medians, for prose, where a slash pair does not read as a sentence.
stimrange <- function() {
  v <- unlist(lapply(c("pos_noun_verb_form", "role_nsubj_obj", "metaphor"), function(k) {
    g <- stim[stim$construction == k, ]
    g <- g[g$model == g$model[1], ]
    c(median(g$n_class_a), median(g$n_class_b))
  }))
  sprintf("%.0f to %.0f", min(v), max(v))
}
stimmed <- function(key) {
  g <- stim[stim$construction == key, ]
  g <- g[g$model == g$model[1], ]
  sprintf("%.0f / %.0f", median(g$n_class_a), median(g$n_class_b))
}

grid_stats <- rd("methods_grid_stats.csv")
gs <- function(constr, col) {
  v <- grid_stats[[col]][grid_stats$construction == constr]
  # median(numeric(0)) is NA, so a construction missing from the stats file used to reach the
  # rendered PDF as a silent "NA". Refuse to render instead.
  if (!length(v)) stop("methods_grid_stats.csv has no rows for construction '", constr,
                       "'; regenerate it with scripts/llm/dataset_stats.py (--vua-dir is ",
                       "required for metaphor)", call. = FALSE)
  if (is.numeric(v)) round(median(v)) else v[1]
}

# The same, but falling back to the last typed value rather than stopping the render, for
# construction keys the stats file only grows on its next regeneration (the role and
# form-keyed POS rows and the corpus counts). Once the stats file carries the key, the data
# wins and the fallback is dead weight.
gsf <- function(constr, col, fallback = NA_real_) {
  v <- grid_stats[[col]][grid_stats$construction == constr]
  v <- v[!is.na(v)]
  if (!length(v)) return(fallback)
  if (is.numeric(v)) round(median(v)) else v[1]
}

# Deepest-layer medians, so the prose cannot drift from the table: every figure quoted in the
# text is read from here rather than typed in. `ld` is the median across models.
llm_deep <- NULL   # filled once `llm` exists, below
ld <- function(constr, col) median(llm_deep[[col]][llm_deep$construction == constr], na.rm = TRUE)
# Same, at the embedding layer, and a range of either across the three constructions. Both read
# NA through rather than erroring, so a column the CSVs predate degrades to a blank in the text.
l0 <- function(constr, col) median(llm[[col]][llm$construction == constr & llm$layer == 0],
                                   na.rm = TRUE)
# A size as a share of the whole representation rather than of the grid: the grid's share of the
# total, times the component's share of the item-by-class structure.
tot <- function(constr, col) ld(constr, col) * ld(constr, "between_adj")

# Range of a column across every category and model at one depth (embeddings, or deepest),
# for prose that quotes a spread over all twelve grids rather than one construction's medians.
allrange <- function(col, deep = TRUE, digits = 4) {
  d <- if (deep) llm_deep else llm[llm$layer == 0, ]
  v <- d[[col]]; v <- v[!is.na(v)]
  sprintf("%s to %s", g(min(v), digits), g(max(v), digits))
}

rng <- function(f, col, digits = 2) {
  v <- vapply(levels(llm_deep$construction), function(k) f(k, col), numeric(1))
  if (all(is.na(v))) return("NA")
  paste0(g(min(v, na.rm = TRUE), digits), " to ", g(max(v, na.rm = TRUE), digits))
}

# The range of a column across the six models at one construction's deepest layer, for the
# few places the prose quotes a spread rather than a median. Same NA-through convention.
# deep = FALSE reads the embedding layer instead, which is where the keying comparison lives.
ldrng <- function(constr, col, digits = 4, deep = TRUE) {
  d <- if (deep) llm_deep else llm[llm$layer == 0, ]
  v <- d[[col]][d$construction == constr]
  if (all(is.na(v))) return("NA")
  sprintf("%s to %s", g(min(v, na.rm = TRUE), digits), g(max(v, na.rm = TRUE), digits))
}

# `scale` pairs the two families at the three matched widths, so a figure can separate the six
# models with three linetypes and two colors rather than leaving three of them on one color.
MODELS <- tibble::tribble(
  ~model,                                       ~family,  ~size,  ~scale,
  "znhoughton/opt-babylm-125m-20eps-seed964",   "BabyLM", "125M", "125M/160M",
  "znhoughton/opt-babylm-350m-20eps-seed964",   "BabyLM", "350M", "350M/410M",
  "znhoughton/opt-babylm-1.3B-20eps-seed964",   "BabyLM", "1.3B", "1.3B/1.4B",
  "EleutherAI/pythia-160m",                     "Pythia", "160M", "125M/160M",
  "EleutherAI/pythia-410m",                     "Pythia", "410M", "350M/410M",
  "EleutherAI/pythia-1.4b",                     "Pythia", "1.4B", "1.3B/1.4B") |>
  dplyr::mutate(scale = factor(scale, c("125M/160M", "350M/410M", "1.3B/1.4B")))

# The three same-token constructions carried in the body. Each file reports the same measure;
# the POS file prefixes its columns std_ because it also stores an unstandardized read.
# POS is read from the SURFACE-FORM-keyed run. Keyed on the lemma instead, an item pools every
# inflected form (run/runs as a noun against run/runs/ran/running as a verb), which are
# different tokens, and the static embedding layer separates them: the layer-0 interaction is
# 0.083-0.172 keyed on the lemma against 0.0010-0.0028 keyed on the form, a 50-100x difference
# that is inflectional rather than representational. The file also holds random-init rows, which
# the paper makes no claim about -- these are endpoint measurements -- so they are filtered out.
# These carry the orientation nulls and the between-grid share. They also hold the random and
# position-ablated inits, so all three sources need the pretrained filter.
llm <- bind_rows(
  rd("llm_unified_form.csv") |>
    ensure("std_size_item_ci_med", "std_size_class_ci_med", "std_size_interaction_ci_med", "std_between_share_adj", "std_obs_size_class", "std_size_class_excludes_zero", "std_obs_size_interaction", "std_size_interaction_excludes_zero", "std_obs_size_item", "std_size_item_excludes_zero") |>
    filter(init == "pretrained") |>
    transmute(model, layer, d, n_items = std_n_items, construction = "POS (noun/verb)",
              size_item = std_size_item, size_class = std_size_class,
              size_int = std_size_interaction,
              leak = std_leak_item_into_class, leak_int = std_leak_int_into_margins,
              leak_null = std_leak_item_into_class_null_hi,
              leak_null_lo = std_leak_item_into_class_null_lo,
              leak_null_med = std_leak_item_into_class_null_med,
              leak_int_null = std_leak_int_into_margins_null_hi,
              leak_int_null_lo = std_leak_int_into_margins_null_lo,
              leak_int_null_med = std_leak_int_into_margins_null_med,
              between = std_between_share, between_adj = std_between_share_adj,
              obs_size_class = std_obs_size_class,
              size_class_excludes_zero = std_size_class_excludes_zero,
              obs_size_interaction = std_obs_size_interaction,
              size_interaction_excludes_zero = std_size_interaction_excludes_zero,
              obs_size_item = std_obs_size_item,
              size_item_excludes_zero = std_size_item_excludes_zero,
              size_item_ci_lo = std_size_item_ci_lo, size_item_ci_med = std_size_item_ci_med,
              size_item_ci_hi = std_size_item_ci_hi,
              size_class_ci_lo = std_size_class_ci_lo, size_class_ci_med = std_size_class_ci_med,
              size_class_ci_hi = std_size_class_ci_hi,
              size_interaction_ci_lo = std_size_interaction_ci_lo,
              size_interaction_ci_med = std_size_interaction_ci_med,
              size_interaction_ci_hi = std_size_interaction_ci_hi),
  rd("llm_role.csv") |>
    ensure("size_item_ci_med", "size_class_ci_med", "size_interaction_ci_med", "between_share_adj", "obs_size_class", "size_class_excludes_zero", "obs_size_interaction", "size_interaction_excludes_zero", "obs_size_item", "size_item_excludes_zero") |>
    filter(init == "pretrained") |>
    transmute(model, layer, d, n_items, construction = "role (subject/object)",
              size_item, size_class, size_int = size_interaction,
              leak = leak_item_into_class,
              leak_int = leak_int_into_margins,
              leak_null = leak_item_into_class_null_hi,
              leak_null_lo = leak_item_into_class_null_lo,
              leak_null_med = leak_item_into_class_null_med,
              leak_int_null = leak_int_into_margins_null_hi,
              leak_int_null_lo = leak_int_into_margins_null_lo,
              leak_int_null_med = leak_int_into_margins_null_med,
              between = between_share, between_adj = between_share_adj,
              obs_size_class = obs_size_class,
              size_class_excludes_zero = size_class_excludes_zero,
              obs_size_interaction = obs_size_interaction,
              size_interaction_excludes_zero = size_interaction_excludes_zero,
              obs_size_item = obs_size_item,
              size_item_excludes_zero = size_item_excludes_zero,
              size_item_ci_lo, size_item_ci_med, size_item_ci_hi,
              size_class_ci_lo, size_class_ci_med, size_class_ci_hi,
              size_interaction_ci_lo, size_interaction_ci_med, size_interaction_ci_hi),
  rd("llm_metaphor.csv") |>
    ensure("size_item_ci_med", "size_class_ci_med", "size_interaction_ci_med", "between_share_adj", "obs_size_class", "size_class_excludes_zero", "obs_size_interaction", "size_interaction_excludes_zero", "obs_size_item", "size_item_excludes_zero") |>
    filter(init == "pretrained") |>
    transmute(model, layer, d, n_items, construction = "metaphor (literal/figurative)",
              size_item, size_class, size_int = size_interaction,
              leak = leak_item_into_class,
              leak_int = leak_int_into_margins,
              leak_null = leak_item_into_class_null_hi,
              leak_null_lo = leak_item_into_class_null_lo,
              leak_null_med = leak_item_into_class_null_med,
              leak_int_null = leak_int_into_margins_null_hi,
              leak_int_null_lo = leak_int_into_margins_null_lo,
              leak_int_null_med = leak_int_into_margins_null_med,
              between = between_share, between_adj = between_share_adj,
              obs_size_class = obs_size_class,
              size_class_excludes_zero = size_class_excludes_zero,
              obs_size_interaction = obs_size_interaction,
              size_interaction_excludes_zero = size_interaction_excludes_zero,
              obs_size_item = obs_size_item,
              size_item_excludes_zero = size_item_excludes_zero,
              size_item_ci_lo, size_item_ci_med, size_item_ci_hi,
              size_class_ci_lo, size_class_ci_med, size_class_ci_hi,
              size_interaction_ci_lo, size_interaction_ci_med, size_interaction_ci_hi)) |>
  left_join(MODELS, by = "model") |>
  mutate(construction = factor(construction, c("POS (noun/verb)", "role (subject/object)",
                                               "metaphor (literal/figurative)")),
         depth = layer / ave(layer, model, construction, FUN = max),  # 0 = embeddings, 1 = final
         # Sizes are shares of the re-split medians, not of the permutation-denoised single
         # split. The cross-split product already has a null of exactly zero, so the permutation
         # baseline was subtracting an estimate of nothing, and it was subtracted from the item
         # and interaction but not from the class, which left the three on different footings.
         # The medians carry no permutation at all and are on the same scale as the intervals
         # quoted beside them, so a share and its interval are finally the same quantity.
         sz_denom = size_item_ci_med + size_class_ci_med + size_interaction_ci_med,
         size_item = size_item_ci_med / sz_denom,
         size_class = size_class_ci_med / sz_denom,
         size_int = size_interaction_ci_med / sz_denom,
         size_item_lo = size_item_ci_lo / sz_denom,
         size_item_hi = size_item_ci_hi / sz_denom,
         size_class_lo = size_class_ci_lo / sz_denom,
         size_class_hi = size_class_ci_hi / sz_denom,
         size_int_lo = size_interaction_ci_lo / sz_denom,
         size_int_hi = size_interaction_ci_hi / sz_denom)

llm_deep <- llm |> group_by(construction, model) |> filter(layer == max(layer)) |> ungroup()

# A construction measured on fewer than six models would silently report a median over whatever
# happened to finish, so say so loudly instead.
local({
  n <- table(droplevels(llm_deep$construction))
  if (any(n < 6)) warning("under-covered constructions: ",
                          paste(names(n)[n < 6], n[n < 6], sep = "=", collapse = ", "),
                          call. = FALSE)
})
```

# Introduction

One of the most fundamental questions in linguistics is to what extent a lexical representation is decomposable into smaller, more abstract representations. For example, while in theory the phrase *trick or treat* is composable from its individual pieces, its meaning-in-context is more akin to a greeting than it is to the individual words. On the other hand, the ability to extend a word or phrase to novel contexts requires some amount of abstraction beyond the exact context it was encountered in.

## Abstract Knowledge

It is well-documented that linguistic knowledge comprises, to some degree, abstract knowledge. For example, children can reliably produce the correct plural of a novel noun, such as producing *wugs* given the novel noun *wug*.

Beyond productivity, there is also evidence from priming studies showing that constructions that share an abstract feature, such as shared syntactic, phonological, or semantic structure, can prime other constructions with the same features.

Even something as deceptively simple as word order preferences, such as binomial ordering preferences, are shaped by abstract preferences.

## Holistic Storage

On the other side of the spectrum, there is also a great deal of evidence that many linguistic constructions are memorized. For example, ....

## Present study: A Hybrid account

The above results demonstrate a great deal of abstract knowledge along with stored knowledge. Typically the centrist position is a bit of a cop-out, but here we argue for it: a linguistic representation contains both abstract components and a memorized, context-specific component. This allows the representation to be extended to new contexts while also enabling its shape to be context-specific, accounting for all the results shown above.

We show this in the present study in a combination of an artificial language study combined with a study of LLMs. We show that representations of various linguistic constructions contain abstract components along with context-specific components. For example, we show that the representation of *walk* (as a verb) is not decomposable strictly into *walk* (in general, regardless of PoS) and a verb representation. Instead, there is a sizable portion of the representation that is unique to *walk* as a verb.

Our main contributions are as follows:

-   We show that lexical representations in LLMs contain both abstract and item-specific components.

-   We introduce a novel measurement to decompose multi-dimensional representations into abstract and item-specific components.

-   We validate this measure using a controlled artificial language design and show that it holds for LLMs as well.

# Experiment 1: Artificial Languages {#sec-exp1}

In Experiment 1 we test our measurement on an artificial language whose features we control. This allows us to validate our measurement while simultaneously examining how different features of the language affect a model's representations.

## Dataset {#sec-exp1-data}

In order to examine the separability of linguistic representations, we constructed an artificial language which comprises a probability distribution over a vocabulary of `r cm(600)` tokens. Every observation is a combination of a **form** (one of `r max(toy$n_form)`), a **class** (one of `r max(toy$n_class)`), and a **context** (drawn from a pool of 400). The motivation for this setup is to give each word a naturalistic environment: the word occurs in a context that favors a specific class-membership (e.g., *walk* as a verb occurs in a different context than *walk* as a noun). Each form is observed `r max(toy$n_obs)` times in each class, every time in a different context, mirroring the patterns of human language. Consider *walk* used as a verb in one of those contexts. Four things determine the distribution over the next token. *Walk* itself contributes the same amount regardless of the class it is used in, and the verb class also contributes the same amount for every item. The third contribution is what is specific to *walk*-as-a-verb over and above those two: every form and every class has its own random vector, and this term multiplies *walk*'s by the verb's, entry by entry. The context contributes the rest, which is noise as far as our question goes, though the model still has to encode it in order to accurately predict the distribution. The four are weighted, summed and passed through a softmax:

$$
\begin{aligned}
&P(\text{next token} \mid f, c, x) = \\
&\quad \operatorname{softmax}\Big( s\big[\, w_\alpha\,\text{item}(f)
  + w_\beta\,\text{class}(c) \\
&\qquad + w_\gamma\,\text{int}(f,c)
  + w_{\text{ctx}}\,\text{context}(x) \,\big] \Big)
\end{aligned}
$$ {#eq-toy}

Each term is isotropically scaled to unit variance before it is weighted, so changing the weights changes the mix of the three effects without changing how much signal there is in total.

The weights $w_\alpha$, $w_\beta$, and $w_\gamma$ in @eq-toy control how much of the true distribution each of the three effects contributes, and we cross them factorially over $\{0, 0.5, 1, 2\}$: zero removes an effect from the language altogether, and the other three values make an effect half as strong as another, equal to it, or twice it. Similarly, $w_{\text{ctx}}$ governs how much the context contributes, so we cross those languages with three context weights: $\{0.5, 2.0, 6.0\}$, ranging from context being a minor part of the true distribution to the dominant one. We train a model on each of the resulting languages at hidden widths of `r paste(sort(unique(toy$d)), collapse = ", ")`, with either a linear or a ReLU activation. The whole grid is then run with five random seeds, each redrawing the language and reinitializing the model, giving `r format(nrow(toy), big.mark = ",")` runs in all, of which `r sprintf("%.1f%%", 100 * mean(toy$converged))` trained to a loss plateau.

Entanglement is guaranteed under a specific subset of the languages. The structure needs `r max(toy$rank)` dimensions, so its ratio to the hidden width, which we call **crowding**, runs from `r g(min(toy$capacity), 3)` to `r g(max(toy$capacity), 1)`. Below one there is room to give every component its own axis, so any entanglement was chosen; above one there are more components than dimensions to put them in, so they cannot all have their own axis and entanglement is forced [@elhage2022superposition].

## Method {#sec-exp1-method}

All code and analyses in this paper can be found here: <https://github.com/znhoughton/representation-separability>.

### The learner

The learner model is a multi-layer perceptron (MLP) with a single hidden layer of width $d$ and either a linear or a ReLU activation function. The model takes as its input three pieces of information: form, class, and context. Each has its own embedding, and the three are concatenated and passed through the model which then outputs a probability distribution over the vocabulary (@fig-training). The training objective is thus to minimize cross-entropy between the model's predicted distribution over the vocabulary and the true distribution.

![What the learner sees. Every form is observed 24 times at each class, each time in a different context; that variation across contexts is what the measurement later splits in two.](fig-training.pdf){#fig-training width="100%" fig-env="figure*"}

### Decomposing a representation {#sec-measure}

We measure separability by decomposing a word-in-context representation into a lexical component, a class component, and an idiosyncratic component. The representation is the MLP's hidden layer, which for a given form and class varies across the contexts it is observed in, so let $M[i,c] \in \mathbb{R}^d$ be the mean of those hidden states for item $i$ used at class $c$. We decompose those item-by-class means by centering, so that:

$$
M[i,c] \;=\; \mu + \alpha(i) + \beta(c) + \gamma(i,c),
$$ {#eq-decomp}

where $\mu$ is the grand mean; $\alpha(i)$, item $i$'s mean across classes minus $\mu$, carries the lexical information; $\beta(c)$, the mean across items at class $c$ minus $\mu$, carries the class information; and $\gamma(i,c)$ carries the idiosyncratic information, what item $i$ at class $c$ holds that neither marginal accounts for (@fig-measurement). Because every item is observed in every class with equal weight, neither marginal absorbs part of the other. The four terms add back to $M[i,c]$ exactly, and no two of them vary together across the item-by-class means, so the squared lengths of the three that vary sum to the total and we report each as a share of it.

Each of these averages is taken over a sample of observations that differ only in context, so each carries some of the context term of @eq-toy as noise. We remove it by dividing the observations behind each average into two even-sized splits, running the decomposition on each split separately, and estimating each component's size as the dot product of its two split estimates, which is large when the splits agree about the component and near zero when they do not. It is the contexts that are divided and not the words, so every word appears in both splits, and the dot product compares two estimates of the same word computed from different sentences. As a result, noise that differs between the splits is independent and cancels.

For each representation we report two measurements. The first measure is the fraction of the representation accounted for by lexical, class, and idiosyncratic information: each component's squared length, summed over its vectors and divided by the total, so that the three fractions sum to one. The second measure quantifies the extent to which those components occupy the same directions. Specifically, we project the item vectors onto the class direction and record the fraction of their squared length that survives, and we project the interaction onto the smallest subspace containing both the item and the class vectors (i.e., the span of the two subspaces). A representation is separable if a word's representation is the sum of an item part and a class part, each on its own axis. Separability can fail in two ways: the parts may be non-orthogonal, so that projecting out the class also removes item information, or there may be a third component that is neither, whatever its orientation. The question is accordingly not whether class information is present, but how it is carried: on a single direction shared across words, or on a separate direction for each of them.

To quantify whether a size is meaningful, we compare each against zero. Size is estimated by taking the dot product of two independent split estimates of the same component, so noise that differs between the splits cancels out and its expected value is zero when the component is absent, which is what makes zero the null. We repeat the measurement over two hundred re-samplings to estimate how much the size varies with which observations fall in which split, and we call a size meaningful when its confidence interval excludes zero. For representations with a component planted at exactly zero, this flags an effect about 5% of the time (Appendix \ref{sec-appendix-validate}), the false-positive rate expected of a 95% confidence interval.

In order to quantify whether the overlap is meaningful, we measure the overlap an effect has with a random subspace. We do so by holding the effect exactly as measured and projecting it onto a random subspace of the same rank, two hundred times, which leaves the effect untouched and varies only the subspace. These two hundred projections give us a distribution of overlaps that an unrelated subspace would yield. An overlap lower than this baseline indicates meaningful orthogonality; an overlap larger than it indicates meaningful non-orthogonality. We present each overlap as a multiple of the median of this distribution, together with how many measurements fall outside its 95% interval. This gives both the size of the effect and how consistently it holds across measurements.

![The four item-by-class means formed by two items in two classes. Under an item part plus a class part, three corners fix the fourth and the parallelogram closes. The arrow is $\gamma(i,c)$, the displacement between the fourth mean and the point that additive account places it at, with length $\lVert \gamma(i,c) \rVert$.](fig-measurement.pdf){#fig-measurement width="78%"}

It's important to note that the weights in @eq-toy describe the distribution the model is trained to match, while the measure reads the representation it forms, so the two sets of numbers are not on a common scale: a learner handed a language that is 85% interaction can represent it with the interaction taking a far smaller share. Appendix \ref{sec-appendix-validate} reports a separate check in which components were planted directly into a representation and recovered almost exactly, which is what licenses reading the results below as facts about the learner rather than about the measure. A run contributes to the results only if at least one of the three components has a re-split interval excluding zero; where none does, the three shares divide noise and whichever lands slightly positive takes all of it.

## Results {#sec-exp1-results}

Results are shown in @fig-toy; the effects of crowding and of the language's other properties are in Appendix \ref{sec-appendix-capacity}.

```{r}
#| label: fig-toy
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 2.5
#| out-width: "100%"
#| fig-cap: "What the learner represents of each component, against how much of that component the language held, under ReLU. Each faint point is one run. Horizontally it sits at that component's proportion of the structure in the language, with context set aside; vertically at its proportion of the structure in the representation. Both are proportions of the same three components, so the two are on a common scale. Lines are least-squares fits through the runs at each hidden width, shaded with a 95% confidence band on the fit; how far the runs themselves scatter is shown by the points, not by that band. The dashed diagonal is where a representation would sit that reproduced the language's composition exactly: above it a component is over-represented relative to the language, below it under-represented. Runs in which none of the three components had a re-split 95% interval excluding zero are not plotted, since the three shares then divide noise; all such runs are at width 8, and their number is given in the text."
ggplot(toy_learn, aes(planted, measured, color = factor(d), fill = factor(d))) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, linewidth = .3, color = UO[["grey"]]) +
  geom_point(size = .32, alpha = .22, stroke = 0) +
  geom_smooth(method = "lm", formula = y ~ x, linewidth = .5, alpha = .16) +
  facet_wrap(~ component, nrow = 1) +
  scale_color_manual(values = WIDC, name = "hidden width") +
  scale_fill_manual(values = WIDC, guide = "none") +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, .5, 1)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "proportion of structure in the language",
       y = "proportion of structure\nin the representation") +
  theme_paper(8) +
  theme(legend.position = "top", legend.margin = margin(0, 0, -6, 0),
        panel.grid.minor = element_blank())
```

The interaction effect in the linear condition was zero only where the language was additive. On the other hand, across the `r cm(nullget("linear", TRUE, "n"))` runs whose language contained an interaction, the linear model's interaction interval excludes zero in `r sprintf("%.0f%%", 100 * nullget("linear", TRUE, "flagged_ez"))` of them, against `r sprintf("%.0f%%", 100 * nullget("ReLU", TRUE, "flagged_ez"))` for ReLU. In contrast, where no interaction was planted the rates were `r sprintf("%.0f%%", 100 * nullget("linear", FALSE, "flagged_ez"))` and `r sprintf("%.0f%%", 100 * nullget("ReLU", FALSE, "flagged_ez"))`. The linear model has no parameter depending on the form and the class together beyond a linear combination of them, yet represents an interaction anyway. One possible explanation for this is that the model can still learn context-specific representations for different items, by learning that a specific item in a specific class occurs with a specific context (and then subsequently representing that information within the context weights). In the models with a ReLU activation function, surprisingly we see an interaction effect represented in the model even in cases where the language is additive. One potential explanation is that the model only ever encounters words in specific contexts; the additive structure of the language is visible across the corpus rather than in any single observation, and the model does not seem to abstract away the interaction completely.

For models with larger dimensionality, the added room seems to be allocated toward representing the interaction. From a hidden width of 16 to a width of 128, the interaction's share of the item-by-class structure rises from `r g(cmp(16, "interaction"), 3)` to `r g(cmp(128, "interaction"), 3)`, while the item effect falls from `r g(cmp(16, "item"), 2)` to `r g(cmp(128, "item"), 2)` and the class effect holds steady at around `r g(cmp(128, "class"), 2)`. In the linear model the same width is spent elsewhere: the item effect rises from `r g(cmplin(16, "item"), 2)` to `r g(cmplin(128, "item"), 2)` while the interaction stays at `r g(cmplin(128, "interaction"), 3)`.

Additionally, at lower dimensions the representation seems to be more variable. This is not particularly surprising since each model run has to decide which information to represent with only a small number of dimensions. As a result, the model makes different decisions from run to runielding vastly different representations of the same language depending on what the model represents in its limited dimensionality. For example, at a width of 8 two runs of the same language differ by `r g(seedspan(8, "item"), 2)` in the item effect and `r g(seedspan(8, "class"), 2)` in the class effect, roughly half the available range; by a width of 32 that has fallen to `r g(seedspan(32, "item"), 2)` and `r g(seedspan(32, "class"), 2)`, and it falls further at 128.

Finally, crowding also governs the amount of overlap between the item and class effects (Appendix \ref{sec-appendix-capacity}). With crowding well below one, ReLU's item-class overlap is `r g(capget("ReLU", 0.125, "overlap"), 4)` against a chance level of `r g(capget("ReLU", 0.125, "floor"), 4)`; at a crowding of 1 it is `r g(capget("ReLU", 1, "overlap"), 3)`, roughly `r g(capget("ReLU", 1, "ratio"), 1)` times chance, and at a crowding of 2 it is `r g(capget("ReLU", 2, "overlap"), 3)`. The linear condition stays at or below chance throughout.

# Experiment 2: Language Models {#sec-exp2}

Experiment 1 established that if the language a learner model encounters is not purely additive, the model will represent the language in a non-separable manner, the extent to which being governed by the size and crowding of the model. Experiment 2 applies the same decomposition to the contextual representations of pretrained language models, across three linguistic categories: part of speech, grammatical role, and metaphor.

## Dataset {#sec-exp2-data}

In order to examine the separability of linguistic representations in models, we examined three linguistic categories: part of speech, grammatical role, and metaphor. For part of speech and grammatical role, we concatenate six general-domain English Universal Dependencies treebanks [@nivre2020ud]: EWT [@silveira2014ewt], GUM [@zeldes2017gum], LinES [@ahrenberg2007lines], ParTUT [@sanguinetti2015partut], GENTLE [@aoyama2023gentle], and PUD [@zeman2017pud], giving `r cm(gsf("ud_corpus", "n_sentences", 41095))` sentences and `r cm(gsf("ud_corpus", "n_pre_grid_tokens", 706323))` tokens. For metaphor we use the VUA20 corpus [@leong2020vua; @steen2010vua], restricted to nouns, verbs, adjectives, and adverbs in VUA's own annotation, which supplies `r cm(gs("metaphor", "n_pre_grid_tokens"))` labeled tokens. We measure two model families at three matched scales: OPT-BabyLM, trained on developmentally plausible quantities of child-available language [@warstadt2023babylm; @houghton2026holisticstorageverb+up], and Pythia, trained on the Pile [@biderman2023pythia], at 125M/160M, 350M/410M, and 1.3B/1.4B parameters, with widths of 768, 1024, and 2048. Appendix \ref{sec-appendix-models} gives parameter counts and training-token estimates.

## Stimuli {#sec-exp2-stimuli}

Our stimuli comprise three linguistic categories: part of speech ([noun]{.smallcaps} versus [verb]{.smallcaps}), grammatical role (*nsubj* versus *obj*, restricted to nouns), and metaphor (literal versus figurative). For each category we select items that occur in both of its classes, such as *walk*, which occurs as both a noun and a verb. An item is a single inflected word form, not a lemma, so different inflections of one lemma are separate items: *walk* and *walking* are both items, each included only if it is itself attested in both classes. Pooling them under one lemma would collect *walk*, *walks*, *walked*, and *walking* into a single item, and the two classes would then be read at different tokens, which the embedding layer could then separate before any contextualization has occurred. Capitalization is ignored when occurrences are grouped, but not in the text the models read, so a sentence-initial *Walk* counts toward the item *walk* and is tokenized as it appears there.[^1]

[^1]: OPT-BabyLM's tokenizer is uncased, so this choice has no affect for that family of models. On the other hand, Pythia's tokenizer is not uncased, however lowercasing the input would be the larger distortion, since a sentence-initial word is capitalized in written English which the model was trained on.

An item is included only if it occurs at least ten times in each class; forms occurring in only one are excluded, which makes part of speech the most restrictive of the three. This yields `r max(llm$n_items[llm$construction == "POS (noun/verb)"])` items for part of speech, `r max(llm$n_items[llm$construction == "role (subject/object)"])` for grammatical role, and `r cm(gs("metaphor", "n_items_kept"))` for metaphor, with median counts per class running from `r stimrange()` (@tbl-constructions), inside the range we validated for the measurement (Appendix \ref{sec-appendix-validate}).

```{r}
#| label: tbl-constructions
#| tbl-cap: "Stimuli for the three linguistic categories, with the two classes of each in parentheses. *Item types* is how many word forms occur in both of them at least ten times; forms occurring in only one are excluded. *Median tokens* is how often a typical form occurs in each of its two classes, in the order they are listed."
tibble::tribble(
  ~class,                                  ~`item types`,
  "part of speech (noun / verb)",          max(llm$n_items[llm$construction == "POS (noun/verb)"]),
  "grammatical role (subject / object)",   max(llm$n_items[llm$construction == "role (subject/object)"]),
  "metaphor (literal / figurative)",       gs("metaphor", "n_items_kept")) |>
  mutate(`median tokens` = c(stimmed("pos_noun_verb_form"),
                             stimmed("role_nsubj_obj"),
                             stimmed("metaphor"))) |>
  kbl(booktabs = TRUE, align = "lrl", linesep = "", escape = FALSE) |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

## Method {#sec-exp2-method}

The decomposition, the two reported quantities, and the replication logic are identical to Experiment 1. For every token we take the hidden state of the word's **last subword** at every layer of each model (using the model's own tokenizer).[^2]

[^2]: Almost every item is a single subword, so this choice has little bearing on the results: zero items are split into multiple subword tokens under Pythia's tokenizer, and only eight are split into multiple subword tokens under OPT-BabyLM's tokenizer.

Transformer hidden states contain a few *rogue* dimensions whose variance exceeds the rest by two to three orders of magnitude [@timkey2021bark], and these would otherwise dominate any subspace estimated from the raw scatter, so we divide each dimension by its standard deviation before decomposing. The measure is accordingly invariant to per-dimension rescaling but not to arbitrary linear reparameterization, which we return to in Appendix \ref{sec-appendix-math}.

## Results {#sec-exp2-results}

The results for the final layer are reported in @tbl-llm; the layer-by-layer results are shown in @fig-llm.

The item effect is the largest component in all three categories, running from `r ldrng("POS (noun/verb)", "size_item", 3)` for part of speech, `r ldrng("role (subject/object)", "size_item", 3)` for grammatical role, and `r ldrng("metaphor (literal/figurative)", "size_item", 3)` for metaphor. What differs between them is how the remainder divides. For grammatical role the class effect, at `r ldrng("role (subject/object)", "size_class", 3)`, is comparable to the interaction at `r ldrng("role (subject/object)", "size_int", 3)`; for part of speech the interaction, at `r ldrng("POS (noun/verb)", "size_int", 3)`, runs well above the class effect at `r ldrng("POS (noun/verb)", "size_class", 3)`; and for metaphor the class effect, at `r ldrng("metaphor (literal/figurative)", "size_class", 3)`, is more than an order of magnitude below the interaction at `r ldrng("metaphor (literal/figurative)", "size_int", 3)`, though it still clears zero in `r outof("metaphor (literal/figurative)", "size_class")`.

Both the class effect and the interaction grow stronger through the layers of the model. The interaction is essentially absent at layer 0 in all three categories, between `r allrange("size_int", deep = FALSE)` across every category and model, and rises to between `r allrange("size_int")` (@fig-llm, top row). This is not surprising since at the first layer, the model has yet to integrate enough context into the representation to commit to the class of the representation.

Turning to how those components are oriented, the class direction carries almost none of the item effect. At the deepest layer, item-class overlap runs `r ldrng("POS (noun/verb)", "leak")` for part of speech and `r ldrng("role (subject/object)", "leak")` for grammatical role, so projecting out the class direction leaves more than 99% of the item effect intact; metaphor's class effect is too small to fit a subspace to, so there is no orientation to read for it. While both overlaps are technically less orthogonal than a random subspace, the overlap itself is tiny, so the two are near-orthogonal rather than exactly orthogonal.

In contrast to the main effects, the interaction is not orthogonal to class or item. Across the six models, `r ldrng("POS (noun/verb)", "leak_int", 3)` of the interaction for part of speech is in the span of the two main effects, for grammatical role `r ldrng("role (subject/object)", "leak_int", 3)` is, and for metaphor `r ldrng("metaphor (literal/figurative)", "leak_int", 3)` is, against nulls whose upper ends run `r ldrng("POS (noun/verb)", "leak_int_null", 3)`, `r ldrng("role (subject/object)", "leak_int_null", 3)`, and `r ldrng("metaphor (literal/figurative)", "leak_int_null", 3)` respectively. The smallest of these overlaps exceeds the largest of the nulls, so the pattern holds in every model. In other words, the small departure from orthogonality between item and class becomes a large one for the interaction.

```{r}
#| label: fig-llm
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 3.1
#| out-width: "100%"
#| fig-cap: "The three components through the network, for every model and linguistic category. Relative depth runs from the static embedding layer (0) to the final block (1). Rows are the three components' shares of the item-by-class structure, which sum to one within each model and layer. Each line is one model: color distinguishes the training corpus, linetype the three matched widths. The shaded band is that model's re-split 95% interval."
# Built by hand rather than pivoted, because each component carries its own re-split interval.
band <- function(d, val, lo, hi, label) {
  tibble(construction = d$construction, family = d$family, scale = d$scale, depth = d$depth,
         component = label, share = d[[val]],
         lo = if (is.null(lo)) NA_real_ else d[[lo]],
         hi = if (is.null(hi)) NA_real_ else d[[hi]])
}
plt <- llm
plt <- bind_rows(
  band(plt, "size_item", "size_item_lo", "size_item_hi", "item effect"),
  band(plt, "size_class", "size_class_lo", "size_class_hi", "class effect"),
  band(plt, "size_int", "size_int_lo", "size_int_hi", "interaction")) |>
  mutate(component = factor(component, c("item effect", "class effect", "interaction")))
# Color separates the two corpora, linetype the three matched widths, so each of the six lines
# is identifiable rather than three of them sharing one color. The shaded band is each model's
# re-split 95% interval; it is narrow enough to read as a thickening of the line, which is the
# point: the spread that matters here is between models, not within a measurement.
ggplot(plt, aes(depth, share, color = family, linetype = scale,
                group = interaction(family, scale))) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = family), color = NA, alpha = .3,
              show.legend = FALSE, na.rm = TRUE) +
  geom_line(linewidth = .55, alpha = .85) +
  facet_grid(component ~ construction, scales = "free_y", switch = "y") +
  scale_color_manual(values = c(BabyLM = UO[["gold"]], Pythia = UO[["blue"]]), name = NULL) +
  scale_fill_manual(values = c(BabyLM = UO[["gold"]], Pythia = UO[["blue"]]), guide = "none") +
  scale_linetype_manual(values = c("125M/160M" = "dotted", "350M/410M" = "dashed",
                                   "1.3B/1.4B" = "solid"), name = NULL) +
  scale_x_continuous(breaks = c(0, .5, 1)) +
  labs(x = "relative depth", y = NULL) +
  theme_paper(8) + theme(strip.placement = "outside")
```

# Experiment 3: Decoding the Interaction {#sec-exp3}

Experiment 2 shows that the interaction is large and that it replicates across independent splits of a word's occurrences, but neither shows what information is contained within the interaction. An item-by-class term captures any reliable difference between a word's uses in the two classes, and some of that difference may be correlated with the context, since metaphorical uses occur in different sentences from literal ones and subjects appear in different constructions from objects.

## Method {#sec-exp3-method}

We test whether the interaction encodes class information by decoding. If $\gamma$ encodes the class, the class should be recoverable from the part of a token's representation that is neither the item effect nor the shared class direction, and we remove both explicitly. All decoding happens *within* a single word, so we ask whether *plant*'s figurative uses can be told from *plant*'s literal ones; the item effect is constant within a word and so cannot contribute. We then estimate $\beta$ as the difference between the two class means pooled over all items, normalize it, and remove that component from every token, so that whatever separates the classes afterwards is not the direction they share.

What remains is the interaction: the part of a word of a specific class that is neither a fact about the word nor a fact about the class. We decode it with a cross-validated logistic probe, five folds, with classes balanced by subsampling within each word so that chance is 0.50, and average the resulting accuracies over words, with a word contributing only if it has at least eight tokens in each class. Some words rest on few tokens, so the mean carries a percentile bootstrap interval over words. The whole procedure is then repeated with the class labels permutated within each word, which preserves every word's token count and class balance while destroying the labeling, and that null is where a probe reading its own cross-validation folds would land. We run this at the deepest layer of the largest model in each family, for all three categories.

## Results {#sec-exp3-results}

Results are reported in @tbl-decode.

```{r}
#| label: tbl-decode
#| tbl-cap: "Decoding the class from the interaction, at the deepest layer of the largest model in each family. The item is held fixed and the shared class direction is projected out, so neither main effect can contribute. *Accuracy* is the mean over them; *null* repeats the decode with the class labels permutated within each form. Brackets are percentile bootstrap intervals over forms, and chance is 0.50."
dec <- bind_rows(
  rd("llm_decode_pos_form.csv") |> mutate(construction = "POS (noun/verb)"),
  rd("llm_decode_interaction.csv") |> filter(construction != "pos") |>
    mutate(construction = recode(construction,
                                 role = "role (subject/object)",
                                 metaphor = "metaphor (literal/figurative)"))) |>
  left_join(MODELS, by = "model") |>
  mutate(construction = factor(construction, c("POS (noun/verb)", "role (subject/object)",
                                               "metaphor (literal/figurative)")))

dec |>
  arrange(construction, family) |>
  transmute(class = shortname(construction), model = paste(family, size),
            accuracy = sprintf("%.2f [%.2f, %.2f]", class_peritem,
                            class_peritem_lo, class_peritem_hi),
            null = sprintf("%.2f [%.2f, %.2f]", class_peritem_null,
                           class_peritem_null_lo, class_peritem_null_hi)) |>
  kbl(booktabs = TRUE, align = "llll", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 7)
```

For both families of LLMs across all three linguistic categories, the class is recoverable from the interaction at well above chance while the permutated null sits on chance, reaching `r g(dec$class_peritem[dec$construction == "POS (noun/verb)" & dec$family == "Pythia"], 2)` for part of speech, `r g(dec$class_peritem[dec$construction == "role (subject/object)" & dec$family == "Pythia"], 2)` for grammatical role, and `r g(dec$class_peritem[dec$construction == "metaphor (literal/figurative)" & dec$family == "Pythia"], 2)` for metaphor, with bootstrap intervals excluding the nulls in all six cases. The null landing on chance is what makes those accuracies interpretable, since it is the same probe, the same folds, and the same word-level averaging with only the labeling destroyed.

Further, the effect is graded the same way the interaction's size is: part of speech decodes best, role next, metaphor least, matching the order in which the interaction's size falls across the three categories (@tbl-llm). In other words, the geometric and the functional measurement agree on the ranking.

Metaphor is the case where this matters most. Its shared class direction is too small to fit an orientation to, so a method reading only that direction would take metaphoricity to be barely represented, yet it decodes from the interaction at `r g(dec$class_peritem[dec$construction == "metaphor (literal/figurative)" & dec$family == "Pythia"], 2)` against a null of `r g(dec$class_peritem_null[dec$construction == "metaphor (literal/figurative)" & dec$family == "Pythia"], 2)`. This suggests that most of what distinguishes its uses lives in the interaction rather than in a direction shared across words.

The results demonstrate that $\gamma$ carries class information rather than generic correlated context, since generic context would not survive holding the word fixed and removing the shared axis. It does not show that $\gamma$ is *only* the label: context correlated with the class within a word survives both removals, so the interaction remains an upper bound.

# Discussion {#sec-discussion}

::: {.content-visible when-format="acl-pdf"}
```{=html}
<!-- PLACEHOLDER: author to write.

     Two arguments were cut from the Experiment 1 results, where they were doing
     interpretive work the results section should not. Kept here rather than lost:

     WHY A LINEAR MODEL CARRIES AN INTERACTION IT CANNOT COMPUTE. The learner's
     representations are not context-invariant. Which contexts *walk* occurs in is partly a
     fact about *walk*, partly about noun-hood, and partly about the two together, and the
     part that varies by word and class jointly cannot be split back into an item part and a
     class part. The form and class vectors are shared across every pairing, so in a linear
     model the contexts a form is observed with are the only place a property of the
     form-as-class can live; its hidden layer is a sum of its inputs, so it preserves whatever
     the contexts do differently in the two classes, carrying the property without ever
     representing it. What ReLU adds is another place to put it: a hidden unit responds to the
     word and the class together, so the model can compute the property from them rather than
     take it in from context.

     ENTANGLEMENT AS SUPERPOSITION. Item and class effects entangle only when there is no room
     for them not to. This is superposition in a setting where we can watch it happen: when the
     number of directions the structure needs exceeds the number available, the components stop
     being orthogonal because they cannot all be, and nothing about the language changed across
     those runs.

     ABSTRACTION VERSUS ENTANGLEMENT IS THE WRONG QUESTION. The class effect and the interaction
     both grow with depth, so models are not choosing between abstraction and entanglement;
     contextualization produces both at once. A description of these representations as either
     abstract or exemplar-like is underdetermined by the data, and the useful question is not
     which one they are but what the balance is.

     WHY PART OF SPEECH IS LESS PORTABLE THAN ROLE. Part of speech has the largest class effect
     of the three in absolute terms, but also much the largest interaction, so proportionally
     less of what distinguishes its classes is carried by a shared direction than for grammatical role. That
     fraction is a ratio, and part of speech loses on the denominator rather than the numerator.

     METAPHOR IS ENCODED WORD BY WORD. Metaphoricity is close to being encoded word by word:
     what makes a use of *plant* figurative is largely not what makes a use of *course*
     figurative. A shared direction carries barely a tenth of what distinguishes its classes.

     WHAT THE DECODE ADDS. The geometric measurement says the interaction is large and does not
     have axes of its own; the decode says it carries the class label specifically. Together
     these say something a shared-direction method cannot reach: for metaphor, a method reading
     only the shared class direction would conclude the distinction is barely represented, when
     in fact it is represented almost entirely word by word.
-->
```
:::

# Limitations {.unnumbered}

Everything here is English, and rests on three linguistic categories measured in two model families at three scales each. The graded ordering we report is an ordering across *these* categories.

The measure is invariant to per-dimension rescaling and to rotation, but not to arbitrary linear reparameterization, so overlap between the item and class effects is relative to the model's own basis. We take that to be the basis of interest, since LayerNorm and the feed-forward nonlinearities act coordinate-wise and a downstream output layer operates on the unwhitened space; a reader who wants a fully gauge-free quantity will not find one here. The component sizes, their growth with depth and whether their intervals exclude zero do not depend on that choice.

```{=latex}
\bibliography{refs}
% The references go here, before the appendix, as ACL requires. The template emits its own
% \bibliography at the end of the document; left active it would give BibTeX a second \bibdata
% command and it would error out, so it is made a no-op from here on.
\renewcommand{\bibliography}[1]{}
\clearpage
```
\onecolumn

\appendix

```{=latex}
% Appendix layout, following the same reasoning as the companion papers.
%
% \raggedbottom matters most. acl.sty sets \flushbottom, which forces each page to exact
% height by stretching the glue between paragraphs and around headings. On an appendix page
% that cannot fill naturally that stretch becomes a gap in the middle of the page with text
% stranded either side. \raggedbottom lets the slack collect at the foot instead. Main-text
% pages keep \flushbottom and the two-column setting.
%
% Appendix tables are longtables, so they typeset inline and break across pages rather than
% floating away from the prose that introduces them. Figures stay floats, pinned with [H].
\raggedbottom
\floatplacement{figure}{H}
\floatplacement{table}{H}
```
# Validating the Measure {#sec-appendix-validate}

Experiment 1 plants structure in a *language* and measures a trained model's representation. However, one limitation with that approach is that it cannot tell a failure in the *measurement* from the learner model simply representing the language differently than the structure planted in the language. This appendix removes the ambiguity by planting $\mu$, $\alpha$, $\beta$, and $\gamma$ **directly in a representation**, with no model and no training, and passing the result to the same function the language models are measured with. Ground truth now sits at the level the measure operates on.

```{r}
#| label: setup-validate
#| include: false
val <- rd("validate_measure.csv")
v_rec <- val |> filter(n_obs == 24, d >= 768) |> group_by(planted_int) |>
  summarize(n = n(), measured = median(size_interaction),
            lo = quantile(size_interaction, .025), hi = quantile(size_interaction, .975),
            flagged = mean(size_interaction_excludes_zero), .groups = "drop")
v_null <- bind_rows(
  val |> filter(planted_int == 0) |> group_by(n_obs) |>
    summarize(effect = "interaction", med = median(size_interaction),
              p95 = quantile(size_interaction, .95),
              flagged = mean(size_interaction_excludes_zero), .groups = "drop"),
  val |> filter(planted_class == 0) |> group_by(n_obs) |>
    summarize(effect = "class effect", med = median(size_class),
              p95 = quantile(size_class, .95),
              flagged = mean(size_class_excludes_zero), .groups = "drop"),
  val |> filter(planted_item == 0) |> group_by(n_obs) |>
    summarize(effect = "item effect", med = median(size_item),
              p95 = quantile(size_item, .95),
              flagged = mean(size_item_excludes_zero), .groups = "drop"))
# False-positive rate of the re-split test BY WIDTH: how often a truly-zero effect's interval
# excludes zero. This is what says whether a small measured effect at a given width is a real
# detection or the test's own floor, and it tells the reader where the test is calibrated.
v_fp_d <- bind_rows(
  val |> filter(planted_int == 0, n_obs == 24) |> group_by(d) |>
    summarize(effect = "interaction", fp = mean(size_interaction_excludes_zero), .groups = "drop"),
  val |> filter(planted_class == 0, n_obs == 24) |> group_by(d) |>
    summarize(effect = "class", fp = mean(size_class_excludes_zero), .groups = "drop"),
  val |> filter(planted_item == 0, n_obs == 24) |> group_by(d) |>
    summarize(effect = "item", fp = mean(size_item_excludes_zero), .groups = "drop"))
fpget <- function(eff, dd) v_fp_d$fp[v_fp_d$effect == eff & v_fp_d$d == dd]
v_ov <- val |> filter(!is.na(leak_item_into_class), n_obs == 24, planted_overlap == 0) |>
  group_by(d) |> summarize(measured = median(leak_item_into_class),
                           lo = quantile(leak_item_into_class, .025),
                           hi = quantile(leak_item_into_class, .975),
                           chance = median(k_class / d), .groups = "drop")
# The power side: a KNOWN fraction of the class direction planted inside the item subspace. The
# measured overlap should rise above chance and be flagged outside its own random-orientation null.
v_ov_power <- val |> filter(!is.na(leak_item_into_class), n_obs == 24, planted_overlap > 0) |>
  group_by(d) |> summarize(measured = median(leak_item_into_class),
                           lo = quantile(leak_item_into_class, .025),
                           hi = quantile(leak_item_into_class, .975),
                           chance = median(k_class / d),
                           detected = mean(leak_item_into_class_outside %in% c(TRUE, "TRUE")),
                           .groups = "drop")
vget <- function(eff, obs, col) v_null[[col]][v_null$effect == eff & v_null$n_obs == obs]
```

The three components are drawn at random and then centered so each satisfies the constraints its role implies: $\alpha$ and $\beta$ sum to zero over their own index, $\gamma$ over both. Without that, a planted $\gamma$ would partly be an item or class main effect and the planted shares would be wrong before the measure saw them. Each is rescaled to hit a target share of the total squared length across the grid, under the same weighting the measure uses. Each mean is then observed $n$ times under isotropic noise, whose scale is set as a multiple of the mean squared length. The grid crosses the planted structure (the same factorial Experiment 1 runs, mapped into share space, plus the three compositions Experiment 2 measures), item counts from 27 to 180, two and four classes, widths from 8 to 2048, 10, 24, and 100 observations per item and class, three noise levels, planted-orthogonal and planted-non-orthogonal effects, and five seeds: `r cm(nrow(val))` runs.

At 24 observations per item and class, inside the 10-to-100 range in which the language-model item sets sit, the measured interaction matches the planted one across the whole range (Table \ref{tbl-val-recovery}), and is flagged in essentially every run where anything was planted and about one in twenty where nothing was.

```{r}
#| label: tbl-val-recovery
#| tbl-cap: "Recovery at 24 observations per item and class in 768 or more dimensions, pooled over noise levels, item counts, and widths. Brackets are the 2.5th and 97.5th percentiles across runs."
v_rec |> filter(planted_int == 0 | planted_int > 0.02) |> head(12) |>
  transmute(`planted` = g(planted_int, 3),
            runs = n,
            `measured` = sprintf("%.4f [%.4f, %.4f]", measured, lo, hi),
            `flagged` = sprintf("%.0f%%", 100 * flagged)) |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "rrrr", linesep = "") |>
  kable_styling(font_size = 9)
```

All three effects read essentially zero when nothing is planted, at every sample size, and each is flagged at about the nominal 5% (Table \ref{tbl-val-null}): at 24 observations per item and class a true zero class effect reads `r g(vget("class effect", 24, "med"), 4)`. This calibration is what licenses reading the small metaphor class effect in @sec-exp2-results as real rather than as the floor: since a true zero reads essentially zero at this sample size, an interval that excludes zero separates even an effect as small as metaphor's from zero rather than measuring noise.

```{r}
#| label: tbl-val-null
#| tbl-cap: "What the measure reports when an effect is planted at exactly zero, by observations per item and class, pooled over every other axis of the validation grid."
v_null |>
  transmute(effect, observations = n_obs,
            median = g(med, 5), `95th pct` = g(p95, 4),
            flagged = ifelse(is.na(flagged), "---", sprintf("%.1f%%", 100 * flagged))) |>
  arrange(effect, observations) |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "lrrrr", linesep = "") |>
  kable_styling(font_size = 9)
```

The rate at which a genuine zero is nonetheless flagged varies with width and effect rather than holding at the nominal 5% (Table \ref{tbl-val-fpd}). The class effect is the most liberal, and most so at the smallest width, where a true zero is flagged `r sprintf("%.0f%%", 100 * fpget("class", 8))` of the time; the item effect is the most conservative, falling to `r sprintf("%.0f%%", 100 * fpget("item", 2048))` at the widest. No setting exceeds ten percent, and pooled across widths the class and interaction are a little more liberal than the item effect. This is the floor the Experiment 1 results at small width are read against, so a lone flag at $d = 8$ is weak on its own. The language-model results sit at $d \geq 768$, where the rate is mildly liberal at worst, and each rests on the effect clearing zero in all six models, which even an eight-percent per-model rate would not produce by chance.

```{r}
#| label: tbl-val-fpd
#| tbl-cap: "How often a component planted at exactly zero is flagged, by width and by effect."
v_fp_d |>
  tidyr::pivot_wider(names_from = effect, values_from = fp) |>
  transmute(`width d` = d,
            item = sprintf("%.0f%%", 100 * item),
            class = sprintf("%.0f%%", 100 * class),
            interaction = sprintf("%.0f%%", 100 * interaction)) |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "rrrr", linesep = "") |>
  kable_styling(font_size = 9)
```

With the item and class effects planted orthogonal, the measured overlap tracks $r/d$ closely and sits a little below it (Table \ref{tbl-val-overlap}). This is the number that Experiment 2's overlaps are read against: at $d = 2048$ a genuinely orthogonal pair reads `r g(v_ov$measured[v_ov$d == 2048], 5)`, and the values reported in @sec-exp2-results are `r sprintf("%.1f to %.1f times", min(llm_deep$leak, na.rm = TRUE) / v_ov$measured[v_ov$d == 2048], max(llm_deep$leak, na.rm = TRUE) / v_ov$measured[v_ov$d == 2048])` that.

```{r}
#| label: tbl-val-overlap
#| tbl-cap: "Measured item-class overlap when the two effects are planted orthogonal, at 24 observations per item and class. *Measured* is the median over the runs at that width with the 2.5th and 97.5th percentiles in brackets. The *chance* column is the analytic level of Appendix \\ref{sec-appendix-math}."
v_ov |>
  transmute(`width d` = d, measured = ci(measured, lo, hi, 5),
            `analytic chance` = g(chance, 5)) |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "rrr", linesep = "") |>
  kable_styling(font_size = 9)
```

The orthogonal case shows the overlap does not cry wolf; the planted-non-orthogonal case shows it does not miss. With a known fraction of the class direction placed inside the item subspace, the measured overlap rises well above the chance level and falls outside its own null in the large majority of runs (Table \ref{tbl-val-power}): at $d = 2048$ a planted-non-orthogonal pair reads `r g(v_ov_power$measured[v_ov_power$d == 2048], 4)` against a chance of `r g(v_ov_power$chance[v_ov_power$d == 2048], 5)`, detected in `r sprintf("%.0f%%", 100 * v_ov_power$detected[v_ov_power$d == 2048])` of runs. So a reported overlap is one the measure can place at chance when the effects are orthogonal and lift above it when the non-orthogonality is real, which is what makes the Experiment 2 overlaps interpretable in both directions.

```{r}
#| label: tbl-val-power
#| tbl-cap: "Measured item-class overlap when a fraction of the class direction is planted INSIDE the item subspace (non-orthogonal), at 24 observations per item and class. *Measured* is the median over the runs at that width with the 2.5th and 97.5th percentiles in brackets; *chance* is the analytic random-orientation level; *detected* is the fraction of runs whose overlap falls outside that null's 95% interval."
v_ov_power |>
  transmute(`width d` = d, measured = ci(measured, lo, hi, 4), chance = g(chance, 5),
            detected = sprintf("%.0f%%", 100 * detected)) |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "rrrr", linesep = "") |>
  kable_styling(font_size = 9)
```

{{< pagebreak >}}

# Crowding and Orthogonality {#sec-appendix-capacity}

Whether the item and class effects come out orthogonal depends on whether the representation had room for them to be, which is a fact about crowding rather than about the language. The languages are held fixed across these runs; only the hidden width changes. @fig-capacity shows the overlap against crowding and @tbl-toy the interaction size and both overlaps at each crowding value.

```{r}
#| label: fig-capacity
#| fig-width: 3.2
#| fig-height: 2.5
#| out-width: "70%"
#| fig-cap: "Item-class overlap against crowding, by activation, over the 7,560 artificial-language runs. Crowding is the rank the structure needs divided by the hidden width. The dashed line is the median upper end of the null for an arbitrarily oriented effect. Both axes are logarithmic."
ACT2 <- c(linear = UO[["green2"]], ReLU = UO[["gold"]])
# one chance line, not one per condition: the two differ slightly because the class subspace
# rank is estimated, and drawing both makes a zigzag that reads as structure when it is not
floor_line <- toy_cap |> group_by(capacity) |> summarize(floor = median(floor), .groups = "drop")
ggplot(toy_cap, aes(capacity, overlap, color = act)) +
  geom_line(data = floor_line, mapping = aes(capacity, floor), inherit.aes = FALSE,
            color = UO[["grey"]], linetype = 2, linewidth = .5) +
  geom_line(linewidth = .6) + geom_point(size = 1.5) +
  annotate("text", x = 0.135, y = 0.032, label = "chance", size = 2.5,
           color = UO[["grey"]], hjust = 0) +
  scale_color_manual(values = ACT2, name = NULL) +
  scale_x_log10(breaks = c(.125, .25, .5, 1, 2)) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(x = "crowding (structure rank / hidden width)", y = "item-class overlap (log)") +
  theme_paper(8)
```

```{r}
#| label: tbl-toy
#| tbl-cap: "Interaction size and the two overlaps by activation and crowding, over the 7,560 artificial-language runs. Crowding is the rank the structure needs divided by the hidden width; each crowding value corresponds to one width, from 128 down to 8. *Interaction* is the median share of the representation over the runs in the row, their interquartile range in brackets, and how many have a re-split 95% interval excluding zero. The overlap columns give the median fraction surviving the projection, then in parentheses how many times its own null median that is and how many runs fall outside that null. The linear arm's interaction stays below the 2% floor at which orientations are reported, so its last column is empty."
toy_cap |>
  transmute(condition = act, crowding = g(capacity, 3), runs = n,
            interaction = sz_int,
            `item-class overlap` = ov_item,
            # the linear arm's interaction stays under the 2% floor at which orientations are
            # reported, so there is no fitted direction to place
            `interaction overlap` = ifelse(act == "linear", "---", ov_int)) |>
  arrange(condition, crowding) |>
  kbl(booktabs = TRUE, align = "lrrrrr", linesep = "", escape = FALSE) |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 7)
```

Two further properties of the language move the composition, and are given here rather than folded into @fig-toy (Table \ref{tbl-lang}). Context competes with the interaction for room: the more of the language that is context, the smaller the interaction's share of the item-by-class structure. And the balance the language strikes between the item and the class is reflected in the representation, the measured item share rising with the planted one.

```{r}
#| label: tbl-lang
#| tbl-cap: "How two other properties of the artificial language move the composition of the item-by-class structure, under ReLU, pooled over widths. Rows within a property partition the same runs. Values are medians of the three shares over the runs in the row."
toy_bylang |>
  transmute(property, level, runs,
            item = g(item, 3), class = g(class, 3), interaction = g(interaction, 3)) |>
  kbl(booktabs = TRUE, align = "llrrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8) |>
  collapse_rows(1, latex_hline = "major", valign = "top")
```

{{< pagebreak >}}

# Model Details {#sec-appendix-models}

```{r}
#| label: tbl-models
#| tbl-cap: "The six models, paired by width. Training-token counts are approximate; the OPT-BabyLM models train for 20 epochs over a 100M-word corpus, so their effective token exposure is roughly twenty times the corpus size. *Layers* counts the transformer blocks, and the measure is applied to each block's output plus the embedding layer."
tibble::tribble(
  ~family,  ~model,          ~params, ~width, ~layers, ~`training data`,     ~`tokens`,
  "BabyLM", "opt-babylm-125m", "125M",   768,      12, "BabyLM (child-available)", "~2B",
  "BabyLM", "opt-babylm-350m", "350M",  1024,      24, "BabyLM (child-available)", "~2B",
  "BabyLM", "opt-babylm-1.3B", "1.3B",  2048,      24, "BabyLM (child-available)", "~2B",
  "Pythia", "pythia-160m",     "160M",   768,      12, "the Pile",                 "~300B",
  "Pythia", "pythia-410m",     "410M",  1024,      24, "the Pile",                 "~300B",
  "Pythia", "pythia-1.4b",     "1.4B",  2048,      24, "the Pile",                 "~300B") |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "llrrrlr", linesep = "") |>
  kable_styling(font_size = 9) |>
  collapse_rows(columns = 1, latex_hline = "major")
```

The six models are listed in @tbl-models. The two families are matched on width rather than on parameter count, which is what the measure is sensitive to: the chance level for overlap is the class subspace rank over the width (Appendix \ref{sec-appendix-math}), so a comparison across models of different width has to be read against different chance levels. Pairing at 768, 1024, and 2048 removes that from the family comparison.

They differ in two ways that matter elsewhere in the paper. The training corpora differ by more than two orders of magnitude in size, which is the manipulation of interest. And they position tokens differently: Pythia uses rotary embeddings, applied inside attention, while OPT learns absolute position embeddings added at the input. Appendix \ref{sec-appendix-positions} uses that difference to rule out absolute position as the source of the interaction.

{{< pagebreak >}}

# The Decomposition {#sec-appendix-math}

## The item-by-class means {#sec-appendix-math-grid}

![Each word in each class is observed in many contexts, and those observations are averaged within the pairing to give one of the item-by-class means. Every one of them is a vector in $\mathbb{R}^d$, and every item appears in every class.](fig-grid.pdf){#fig-grid width="100%"}

A word used in a given class is observed in many contexts, and the representation differs a little in each. Averaging those observations within the pairing gives one of the item-by-class means: for items $i \in \{1,\dots,I\}$ crossed with classes $c \in \{1,\dots,C\}$, $M[i,c] \in \mathbb{R}^d$ is the mean representation of item $i$ in class $c$ (@fig-grid). An item observed in only one class is dropped, as is an item with fewer than 10 tokens in either class, so the set is complete and every mean rests on a comparable number of observations. Because every item appears at every class, the means arrange into $I$ by $C$ positions, and the rest of this appendix calls that arrangement the grid when what matters is the positions rather than the vectors sitting at them. The variation across contexts that the averaging sets aside is not discarded; it is what the two splits of @sec-measure are drawn from.

## The decomposition as a projection {#sec-appendix-math-proj}

![Centering, drawn for a single hidden dimension read across the whole grid. The all-ones direction is where every position takes the same value, so it carries no differences and no variance, and $\mu$ is the projection onto it. Removing it leaves a hyperplane of dimension $N-1$, and every remaining component is a displacement inside it. The dashed copy is that displacement moved to the origin, where it lies in the hyperplane.](fig-projection.pdf){#fig-projection width="82%"}

Two spaces are in play. The representations themselves live in $\mathbb{R}^d$. The decomposition happens in the grid index space: fix one hidden dimension, read it across all $IC$ positions of the grid, and it is a vector in $\mathbb{R}^{IC}$ (@fig-projection). Everything below is that operation carried out once per hidden dimension, with the results stacked back up into the vectors $\alpha_i$, $\beta_c$, and $\gamma_{ic}$.

In that space the all-ones direction is the one in which every position of the grid takes the same value. Moving along it adds the same amount everywhere and creates no difference between any two positions, so it is the direction along which nothing distinguishes anything, and $\mu$ is the projection onto it. Removing it leaves the orthogonal complement, and the three remaining components are displacements inside that complement rather than raw values.

Each component is fixed by two choices, one per axis: along the item axis either average or center, and along the class axis the same. Averaging keeps what is common and discards what makes positions differ along that axis; centering does the reverse. The four combinations are the four components of @eq-decomp:

|               | varies over items | varies over classes |
|:--------------|:-----------------:|:-------------------:|
| $\mu$         |        no         |         no          |
| $\alpha_i$    |        yes        |         no          |
| $\beta_c$     |        no         |         yes         |
| $\gamma_{ic}$ |        yes        |         yes         |

So $\alpha_i$ is what holds of an item in both classes, $\beta_c$ is what holds of a class for every item, and $\gamma_{ic}$ is what varies along both at once, which is why it cannot be a fact about either one alone. Since averaging and centering along an axis add back to what was there, the four reconstruct $M[i,c]$ exactly, and they are mutually orthogonal.

Centering an axis costs one dimension, the all-ones direction along it. The four components therefore occupy $1$, $I{-}1$, $C{-}1$, and $(I{-}1)(C{-}1)$ of the grid's $IC$ positions, which is why two classes leave $\beta$ a single direction whatever $I$ is. Nothing is left over as error: $\gamma$ is not a residual in the sense of a model that failed to fit, but the part of the representation belonging to the pairing of a particular item with a particular class.

Two consequences are used in the body. First, $\sum_i\alpha_i = 0$, $\sum_c\beta_c = 0$, and $\sum_i\gamma_{ic} = \sum_c\gamma_{ic} = 0$ follow from the centering rather than being imposed as constraints. Second, an **additive** set of means, one of the form $M[i,c] = A(i) + B(c)$, has $\gamma = 0$ *identically*: writing $\bar A$ and $\bar B$ for the averages, $\mu = \bar A + \bar B$, $\alpha_i = A(i) - \bar A$, $\beta_c = B(c) - \bar B$, and the remainder cancels exactly. This is the algebra behind the linear condition of Experiment 1: its population means, the pairing means taken over the same context distribution everywhere, are additive and its interaction is zero. What it is measured on is the means of its observed sample, which can also hold a trace of each pairing's own contexts (Experiment 1, Results), so the floor the linear arm provides is read off the planted-zero languages, not deduced.

## Sizes {#sec-appendix-math-sizes}

![How many positions of the grid each component occupies. $\alpha_i$ does not depend on the class, so it occupies all $C$ positions of its row; $\beta_c$ does not depend on the item, so it occupies all $I$ positions of its column; $\gamma_{ic}$ takes its own value at every position and is never repeated.](fig-replication.pdf){#fig-replication width="100%"}

Because the three components are mutually orthogonal, their squared lengths add rather than overlapping, and the total squared length across the grid partitions exactly:

$$
\sum_{i,c}\lVert M[i,c]-\mu\rVert^{2}
= C\sum_{i}\lVert\alpha_i\rVert^{2}
+ I\sum_{c}\lVert\beta_c\rVert^{2}
+ \sum_{i,c}\lVert\gamma_{ic}\rVert^{2},
$$

$C$ and $I$ are not a choice of scaling. They are how many positions of the grid each component occupies (@fig-replication): $\alpha_i$ does not depend on the class, so it sits at all $C$ positions of its row, and $\beta_c$ does not depend on the item, so it sits at all $I$ positions of its column. Every one of the $I$ items is displaced by the same $\beta_c$, so the displacement present in the grid is $I$ copies of it. The interaction is simply summed because it takes its own value at every position and is never repeated. Dividing all three by $IC$ would report averages per position instead of totals and leave every share unchanged; what the occupancy counts rule out is weighting the three differently from one another.

Sizes are reported as shares of the total. All three terms scale linearly in $I$, so the shares do not depend on how many items there are and are comparable across categories of different size.

The partition above is of the item-by-class structure, which has already averaged the contexts away, so it leaves open how much of the representation that structure is. Splitting each observation into its pairing mean and its deviation from it, and using the fact that deviations from a mean sum to zero, gives a second exact partition one level up:

$$
\sum_{x}\lVert x - \bar x\rVert^{2}
= \underbrace{\sum_{i,c}\sum_{x \in (i,c)} \lVert x - \hat M[i,c]\rVert^{2}}_{\text{within a pairing}}
+ \underbrace{\sum_{i,c} n_{ic}\lVert \hat M[i,c] - \bar x\rVert^{2}}_{\text{between pairings}} .
$$

The raw between fraction overstates the item-by-class structure, because $\hat M[i,c]$ is itself an average of $n_{ic}$ observations and carries $d\sigma^2/n_{ic}$ of context noise that lands in the between term. The bias falls as tokens per item rise, so it differs across categories and would not be comparable between them. We therefore report the variance-component form: the within mean square estimates the context variance, the between mean square estimates that plus $n_0$ times the true spread of the pairing means with $n_0$ the effective group size of the unbalanced design, and the difference gives the spread alone. Planting no structure at all, the raw fraction reads $0.081$ at 12 observations per pairing and $0.005$ at 200, while the corrected one reads zero at both.

Because the corrected quantities are variance components rather than weighted sums, they compose with the sizes: each size times that structure's share gives that component's share of the representation, and the three plus the context term account for it entirely.

## The overlaps and their chance level {#sec-appendix-math-overlap}

The sizes say how much of the representation each component accounts for and nothing about direction. Two components can both be large whether or not they occupy the same directions, and separability is the second question: a word's representation is an item part plus a class part only if the two can be varied independently, which requires them to lie on distinct axes. That question lives in $\mathbb{R}^d$, not in the grid index space the sizes were computed in, and unlike the sizes it does not partition, because each overlap is a relation between two particular components rather than a share of one whole.

Each overlap projects one component into the subspace another occupies:

| projected | into the subspace of         |
|:----------|:-----------------------------|
| $\alpha$  | $\beta$                      |
| $\beta$   | $\alpha$                     |
| $\gamma$  | $\alpha$ and $\beta$ jointly |

The first two are not redundant with each other. One direction and thirty-odd are very different targets to land in by chance, and the chance level below is what puts them on a comparable footing.

Subspaces are orthonormal bases from the leading right singular vectors, with rank set by the participation ratio $r = \operatorname{round}\!\big((\sum_k\lambda_k)^2/\sum_k\lambda_k^2\big)$ of the scatter eigenvalues, capped at the algebraic maxima above. The participation ratio counts directions discounted by how much each carries: it returns 1 when all the spread lies along one direction and $k$ when it is even across $k$ of them. With two classes $\beta$ spans one direction, so $r_{\text{class}} = 1$ throughout Experiment 2.

Writing $S$ for the $d \times r$ basis and $V$ for the component being projected, the overlap is $\lVert V S\rVert_F^2 / \lVert V\rVert_F^2$. $V$ is complete, every vector of the component at full dimension; only $S$ is rank-truncated. So the rank enters through the target alone, which is why it fixes the chance level without altering any measured overlap.

The overlap is a squared cosine. For one vector and one direction, $\langle v, u\rangle^2/\lVert v\rVert^2 = \cos^2\theta$ exactly, so the ratio is a length-weighted mean of $\cos^2\theta_i$ over the vectors of $V$, with $\theta_i$ the angle between vector $i$ and the nearest direction in $S$. Weighting by squared length keeps items with small, poorly determined effects from moving the number. The square root inverts to an angle, which is the more legible scale near the floor: an overlap of $0.19$ is $64^\circ$ and one of $0.0056$ is $86^\circ$, nearly perpendicular rather than merely small.

Squaring rather than summing the signed dot products is not cosmetic. Because $\sum_i \alpha_i = 0$, the signed version is $\langle \sum_i \alpha_i, \beta\rangle = 0$ identically, whatever the geometry: items leaning one way along the class axis cancel those leaning the other. Squaring is also what makes the two directions of a component count equally, which is what the question requires, since projecting the class direction out costs an item the same either way.

The chance level is not zero. For a randomly oriented $V$,

$$
\mathbb{E}\big[\lVert V S\rVert_F^2\big] = \operatorname{tr}(V^{\top}V)/d \cdot r
\;\Longrightarrow\;
\mathbb{E}[\text{overlap}] = r/d ,
$$

whatever $V$'s rank or shape. At $d = 2048$ that is $0.0005$ for a one-direction class subspace, and Appendix \ref{sec-appendix-validate} recovers it empirically, a planted-orthogonal pair landing a little below its own $r/d$ at every width. Any overlap has to be read against that number, which moves by two orders of magnitude across the widths used in this paper.

The measure is exactly invariant to per-dimension rescaling, which the standardization absorbs, and to orthogonal rotation of already-standardized data. It is **not** invariant to general linear reparameterization: we do not whiten, because whitening would make a linear representation read separable in any basis and would discard the basis the model actually computes in. Overlap values are therefore basis-relative, while the sizes and their growth with depth do not depend on the choice.

{{< pagebreak >}}

# Deepest-Layer Measurements {#sec-appendix-llm}

Table \ref{tbl-llm} gives the deepest-layer values behind the curves of @fig-llm, together with the null each overlap is read against.

```{r}
#| label: tbl-llm
#| tbl-cap: "Deepest-layer measurements, one row per model. The two overlaps are the fraction of the item effect lying in the class subspace and the fraction of the interaction lying in the span of the two main effects, each followed in parentheses by how many times its own rotation null's median it is. That null holds the effect fixed and rotates the target subspace at random 200 times. The class effect and the interaction are shares of the item-by-class structure, with each model's re-split 95% interval in brackets."
llm |> group_by(construction, model) |>
  filter(layer == max(layer)) |> ungroup() |>
  arrange(construction, family, scale) |>
  # One row per model. A median across the six hid exactly what the six are being compared for,
  # and the sizes' re-split intervals are far narrower than the spread across models, so pooling
  # them reported a precision that belonged to a single measurement rather than to the set.
  # Each overlap is against its own model's null rather than r/d: the two differ, since r/d
  # assumes the subspace has its algebraic maximum rank while the projection uses the rank
  # actually fitted, roughly half of it for the marginal subspace.
  transmute(class = shortname(construction),
            model = paste(family, size),
            `item-class overlap` = ov1(leak, leak_null_med, 4),
            `interaction overlap` = ov1(leak_int, leak_int_null_med, 3),
            `class effect` = ci(size_class, size_class_lo, size_class_hi),
            `interaction` = ci(size_int, size_int_lo, size_int_hi)) |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "llrrrr", linesep = "") |>
  kable_styling(font_size = 7) |>
  collapse_rows(1, latex_hline = "major", valign = "top")
```

{{< pagebreak >}}

# Positional Embeddings Are Not the Source {#sec-appendix-positions}

One architectural explanation of the interaction has to be ruled out before it can be read as a fact about lexical representation. OPT-BabyLM adds a *learned absolute position embedding* at the input. The two classes of a linguistic category are not positionally interchangeable in a sentence, since subjects precede objects and nouns and verbs occupy different slots, so a representation that varies with position varies with the class. If that is what the decomposition is picking up, the interaction would be an artifact of where words sit rather than of how they are represented.

We re-extracted every OPT-BabyLM model with *decoder.embed_positions* zeroed and measured it identically. The ablation was verified per model rather than assumed: with the embedding zeroed, the same token at two different offsets must give identical layer-0 states, and it does, in every model (drift $0.0$ to machine precision). Pythia has no such module, so there is nothing to zero there.

The right quantity is the class effect's share of class-relevant structure, the class effect divided by the sum of the class effect and the interaction, which we call the *portable fraction*. If positions were manufacturing the interaction, removing them should leave a larger share of what distinguishes the two classes in a direction shared across words. Absolute sizes cannot be compared here: zeroing the embeddings in a model that was trained with them is a heavy intervention, and it shrinks every component at once, the class effect included.

```{r}
#| label: tbl-posabl
#| tbl-cap: "Trained OPT-BabyLM at the deepest layer, with absolute position embeddings intact and zeroed. Sizes are shares of the item-by-class structure and the portable fraction is the class effect divided by the sum of the class effect and the interaction."
rd("llm_unified_form_ablation.csv") |>
  filter(grepl("babylm", model), grepl("^pretrained", init)) |>
  group_by(model, init) |>
  slice_max(layer, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(model = sub(".*/", "", sub("-20eps-seed964", "", model)),
            positions = ifelse(grepl("noposemb$", init), "zeroed", "intact"),
            class = std_size_class, interaction = std_size_interaction,
            portable = std_size_class / (std_size_class + std_size_interaction)) |>
  arrange(match(model, c("opt-babylm-125m", "opt-babylm-350m", "opt-babylm-1.3B")),
          desc(positions)) |>
  transmute(model, positions, `class` = g(class, 3), `interaction` = g(interaction, 3),
            `portable fraction` = g(portable, 3)) |>
  kbl(longtable = FALSE, booktabs = TRUE, align = "llrrr", linesep = "") |>
  kable_styling(font_size = 9)
```

Removing the position embeddings makes these representations *less* separable, not more. The portable fraction falls in every model, from 0.439 to 0.264, from 0.468 to 0.253, and from 0.333 to 0.198 (Table \ref{tbl-posabl}). Whatever the absolute position embedding contributes, it is contributing to the part of class structure that *is* carried by a shared direction. Take it away and what remains is more word-specific, not less. The interaction is not an artifact of position.

A second line of evidence points the same way and needs no ablation at all. Pythia never had an absolute position embedding to remove, and its portable fractions are 0.253, 0.287 and 0.262, at or below every intact OPT-BabyLM value. The family without the mechanism shows the phenomenon at least as strongly as the family with it. An explanation resting on absolute position embeddings would have to account for the models that do not have them.

We tested this for part of speech, the category where the positional worry is sharpest because noun and verb uses of a form occupy systematically different sentence positions. Grammatical role carries the same concern for the same reason, and behaves the same way. Metaphor does not, since literal and figurative uses of a word are not positionally distinguished.
