---
title: "Are Lexical Representations Separable? Item and Category in Learned Distributed Representations"
author:
  - name: "Anonymous"
    affiliations:
      - name: "Anonymous Institution"
    email: anon@example.edu
abstract: |
  PLACEHOLDER --- author to write.
bibliography: refs.bib
acl-mode: review
number-sections: true
tbl-cap-location: bottom
format:
  acl-pdf:
    keep-tex: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr); library(dplyr); library(tidyr); library(ggplot2); library(kableExtra)
library(patchwork); library(scales)

DATA <- file.path("..", "data")
rd <- function(f) suppressMessages(read_csv(file.path(DATA, f), show_col_types = FALSE))

# UO palette, as in the companion papers. Green = the linear reference, gold = the
# nonlinearity; amber and blue are held for the model families in Experiments 2-3.
UO <- c(green = "#154733", green2 = "#007030", gold = "#F0B323",
        amber = "#875E00", blue = "#005581", grey = "#6E6E6E")
ACT_COL <- c(identity = UO[["green2"]], relu = UO[["gold"]])
ACT_LAB <- c(identity = "linear", relu = "ReLU")

theme_paper <- function(base = 9) {
  theme_minimal(base_size = base) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, colour = "grey88"),
          strip.text = element_text(face = "bold", size = base),
          legend.position = "top", legend.margin = margin(0, 0, -4, 0),
          legend.key.height = unit(9, "pt"),
          axis.title = element_text(size = base))
}

# ---- Experiment 1: the artificial-language grid ----
# 7,560 cells, measured with unified_split -- the SAME function and arguments Experiment 2
# uses. The two independent estimates are two halves of each cell's observations, so there is
# no alignment step and no projection. Capacity is a reported variable, never a filter.
toy <- rd("artificial_language_grid.csv") |>
  mutate(act = factor(activation, c("identity", "relu"), c("linear", "ReLU")),
         # the share of a random vector that lands in the class direction by chance. An
         # orthogonality reading means nothing without it, and it moves 16-fold across the
         # widths used here, so it is carried alongside rather than argued about.
         floor = k_class / d)

# The linear arm is a negative control by ALGEBRA, not by stipulation: with additive inputs a
# linear map gives cell means of the form A(f)+B(c), whose interaction is identically zero
# however much the language contains. Whatever it reports is the measure's own noise, read at
# that exact width, sample size and context level.
toy_null <- toy |> group_by(act, planted = w_int > 0) |>
  summarise(n = n(), med = median(size_interaction),
            p95 = quantile(size_interaction, .95),
            flagged = mean(sig_interaction), .groups = "drop")

# Does the measured interaction track the planted one? Binned, because the achieved share is
# continuous (neither the softmax nor the normalisation is linear in the weights).
toy_track <- toy |> filter(w_item > 0, w_class > 0) |>
  mutate(bin = cut(ach_int, c(-Inf, .005, .05, .15, .30, .55, Inf),
                   labels = c("none", ".005-.05", ".05-.15", ".15-.30", ".30-.55", ">.55"))) |>
  group_by(act, bin) |>
  summarise(n = n(), ach = median(ach_int), size_int = median(size_interaction),
            lo = quantile(size_interaction, .25), hi = quantile(size_interaction, .75),
            flagged = mean(sig_interaction), .groups = "drop")

# Capacity: below 1 every component can have its own axis, above 1 they cannot.
toy_cap <- toy |> filter(w_int > 0, w_item > 0, w_class > 0) |>
  group_by(act, capacity) |>
  summarise(n = n(), size_int = median(size_interaction),
            overlap = median(leak_item_into_class, na.rm = TRUE),
            floor = median(floor),
            ratio = median(leak_item_into_class / floor, na.rm = TRUE), .groups = "drop")

# Context is the nuisance dial: how much of the target is about something other than the item,
# the class and their conjunction. The high end is the regime Experiment 2 sits in.
toy_noise <- toy |> filter(w_int > 0, w_item > 0, w_class > 0) |>
  mutate(cbin = cut(ach_ctx, c(-Inf, .2, .5, .8, Inf),
                    labels = c("<20%", "20-50%", "50-80%", ">80%"))) |>
  group_by(act, cbin) |>
  summarise(n = n(), size_int = median(size_interaction),
            flagged = mean(sig_interaction), .groups = "drop")

g <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
cm <- function(x) format(x, big.mark = ",")
nullget <- function(a, pl, col) toy_null[[col]][toy_null$act == a & toy_null$planted == pl]
capget <- function(a, cap, col) toy_cap[[col]][toy_cap$act == a & toy_cap$capacity == cap]

# ---- Experiment 2: language model representations ----
grid_stats <- rd("methods_grid_stats.csv")
gs <- function(constr, col) {
  v <- grid_stats[[col]][grid_stats$construction == constr]
  if (is.numeric(v)) round(median(v)) else v[1]
}

MODELS <- tibble::tribble(
  ~model,                                       ~family,  ~size,   ~d,
  "znhoughton/opt-babylm-125m-20eps-seed964",   "BabyLM", "125M",  768,
  "znhoughton/opt-babylm-350m-20eps-seed964",   "BabyLM", "350M",  1024,
  "znhoughton/opt-babylm-1.3B-20eps-seed964",   "BabyLM", "1.3B",  2048,
  "EleutherAI/pythia-160m",                     "Pythia", "160M",  768,
  "EleutherAI/pythia-410m",                     "Pythia", "410M",  1024,
  "EleutherAI/pythia-1.4b",                     "Pythia", "1.4B",  2048)

# The three same-token constructions carried in the body. Each file reports the same measure;
# the POS file prefixes its columns std_ because it also stores an unstandardized read.
# POS is read from the SURFACE-FORM-keyed run. Keyed on the lemma instead, a cell pools every
# inflected form (run/runs as a noun against run/runs/ran/running as a verb), which are
# different tokens, and the static embedding layer separates them: the layer-0 interaction is
# 0.083-0.172 keyed on the lemma against 0.0010-0.0028 keyed on the form, a 50-100x difference
# that is inflectional rather than representational. The file also holds random-init rows, which
# the paper makes no claim about -- these are endpoint measurements -- so they are filtered out.
llm <- bind_rows(
  rd("llm_unified_form.csv") |> filter(init == "pretrained") |>
    transmute(model, layer, d, n_items, construction = "POS (noun/verb)",
              size_item = std_size_item, size_class = std_size_class,
              size_int = std_size_interaction, sig = std_sig_interaction,
              leak = std_leak_item_into_class, leak_int = std_leak_int_into_margins),
  rd("llm_role.csv") |>
    transmute(model, layer, d, n_items, construction = "role (subject/object)",
              size_item, size_class, size_int = size_interaction,
              sig = sig_interaction, leak = leak_item_into_class,
              leak_int = leak_int_into_margins),
  rd("llm_metaphor.csv") |>
    transmute(model, layer, d, n_items, construction = "metaphor (literal/figurative)",
              size_item, size_class, size_int = size_interaction,
              sig = sig_interaction, leak = leak_item_into_class,
              leak_int = leak_int_into_margins)) |>
  left_join(MODELS, by = "model") |>
  mutate(construction = factor(construction, c("POS (noun/verb)", "role (subject/object)",
                                               "metaphor (literal/figurative)")),
         depth = layer / ave(layer, model, construction, FUN = max))  # 0 = embeddings, 1 = final
```

# Introduction

::: {.content-visible when-format="acl-pdf"}
```{=html}
<!-- PLACEHOLDER: author to write. Frames the paper as an extension of Exemplars in
     Disguise: if a learner's behavior cannot distinguish exemplar storage from
     abstraction, ask instead whether its representations separate into an item part and
     a class part. Ends with the Present Study paragraph setting up the three
     experiments. -->
```
:::

# Experiment 1: What a Learner Builds from a Known Language {#sec-exp1}

Experiment 1 has two jobs, and comparing artificial learners to language models is not one of them. The first is to establish that the measure reports what is there: we build an artificial language whose structure we control, dial each of the three effects independently from nothing to everything, and check what the measure recovers. The second is to ask what a learner *does* with that structure when a separable solution is available to it and it has room to find one. Whether any of this holds of real language is Experiment 2's question, and the numbers there stand on their own.

## Dataset {#sec-exp1-data}

The artificial language is a next-token prediction task over a vocabulary of `r cm(600)` tokens. Every observation is a triple: a **form** (one of `r max(toy$n_form)`), a **class** (one of `r max(toy$n_class)`), and a **context** drawn from a pool of 400. Each (form, class) cell is observed `r max(toy$n_obs)` times, each time in a different context. That is what a word in a corpus looks like: one type, many occurrences, each in different company.

The target distribution for an observation is built from four independently generated terms, weighted and passed through a softmax:

$$
P(\text{next token} \mid f, c, x) = \operatorname{softmax}\Big(s\big[\,w_\alpha\,\text{item}(f) + w_\beta\,\text{class}(c) + w_\gamma\,\text{int}(f,c) + w_{\text{ctx}}\,\text{context}(x)\,\big]\Big)
$$ {#eq-toy}

The item and class terms are what those factors contribute on their own. The **interaction** term is built by multiplying the form's and the class's latent codes together elementwise before mapping them onto the vocabulary, and that product is the point: a *sum* can always be split into a form part plus a class part, and a *product* cannot, so this term is joint in a way no decomposition into the two can absorb. The **context** term is nuisance, but it is real information the model has to encode, exactly as the rest of a sentence is for a language model. Every term is normalized to unit scale before weighting, so the weights set composition and nothing else; total signal strength, and hence task difficulty, is constant across the grid.

**The grid.** The three effect weights are crossed **factorially**, each over $\{0, 0.5, 1, 2\}$, giving 63 combinations. Crossing rather than choosing is what supplies the nulls: 15 combinations plant no item effect, 15 no class effect, and 15 no interaction, so the measure is asked to report zero in every case where zero is the truth. These are crossed in turn with the context weight (three levels), the hidden width, the output nonlinearity and five random draws of the language, for `r format(nrow(toy), big.mark = ",")` cells, of which `r sprintf("%.1f%%", 100 * mean(toy$converged))` trained to a loss plateau.

**Capacity.** The structure needs `r max(toy$rank)` dimensions and the hidden layer has `r paste(sort(unique(toy$d)), collapse = ", ")`, so their ratio, which we call **capacity**, runs from `r g(min(toy$capacity), 3)` to `r g(max(toy$capacity), 1)`. Below one there is room to give every component its own axis, so any entanglement was chosen; above one, dimension counting makes separation impossible and entanglement is forced. Capacity is reported as a variable, never used to select cells.

## Method {#sec-exp1-method}

### The learner

The model has an embedding per **form**, an embedding per **class**, and an embedding per **context** (@fig-training). It has none per *cell*. A cell's representation is therefore **computed** from shared pieces rather than stored in a parameter of its own, which is what a transformer does with a token and its context, and it is what lets the measurement here be identical to Experiment 2's. The three embeddings are concatenated, passed through one hidden layer of width $d$ with either an identity or a ReLU nonlinearity, and read out linearly to the vocabulary. We measure the hidden layer.

Training minimizes the exact expected cross-entropy against the true target rather than a sampled-token estimate, so there is no minibatch noise, and stops when that loss plateaus. Stopping depends on nothing derived from the measurement, but it does stop the two conditions at different points, so each cell records whether it plateaued or exhausted its budget.

**The linear condition is a negative control by algebra, not by stipulation.** With additive inputs a linear map produces cell means of the form $A(f) + B(c)$, and such a grid has $\gamma = 0$ *identically* (Appendix \ref{sec-appendix-math}). The linear learner therefore cannot represent an interaction however much the language contains, and it visibly fails to fit when the language contains one. Whatever the measure reports there is its own noise, read off at that exact width, sample size and context level, which is a stronger calibration than any argument about noise floors. ReLU is the condition that can build an interaction, and the one that corresponds to a transformer: nonlinear inside, linear readout.

![What the learner sees and what it is trained to predict. Every (form, class) cell is observed 24 times, each in a different context; that within-cell variation is what the measurement later splits in half. $\odot$ is the elementwise product: a sum splits into an item part plus a class part, and a product does not.](fig-training.pdf){#fig-training width=100%}

### Decomposing a representation {#sec-measure}

Let $M[i,c] \in \mathbb{R}^d$ be the mean representation of item $i$ used at class level $c$. We split that grid into four geometric components,

$$
M[i,c] \;=\; \mu + \alpha(i) + \beta(c) + \gamma(i,c),
$$ {#eq-decomp}

by centering: $\mu$ is the grand mean, $\alpha(i)$ is item $i$'s row mean minus $\mu$, $\beta(c)$ is level $c$'s column mean minus $\mu$, and $\gamma(i,c)$ is what remains. The familiar reading is the main-effects-plus-interaction partition, with the difference that the dependent variable is a vector, so each component is a *direction* in representational space rather than a number. In mixed-model terms $\alpha$ is a by-item random intercept, $\beta$ is the fixed effect of class, and $\gamma$ is the by-item random slope for class; the question of this paper is how much of the class effect is carried by $\beta$ rather than by $\gamma$. We use that mapping for intuition only. Nothing is fitted, there is no shrinkage, and no distribution is assumed over the by-item terms.

A representation is **separable** if a word's representation is the sum of an item part and a class part, with the two on their own axes. It can fail in two distinguishable ways, and we measure them separately. It fails if the components are **non-orthogonal**, so that projecting out the class also removes item information; and it fails if there is a **third component**, an item-by-class term that is neither an item part nor a class part, whatever its orientation. A representation with a large $\gamma$ does not factor into item plus class even if $\gamma$ sits on an axis of its own.

Because the four components are mutually orthogonal *as sources*, their squared lengths partition the structure of the grid exactly, and we report each as a share of the total. **Interaction size** is $\gamma$'s share, which measures the second kind of failure. For the first we report two projections: **item-class overlap**, the fraction of the item effect's energy lying inside the class effect's subspace, and **interaction overlap**, the fraction of the interaction's energy lying inside the subspace spanned by the two main effects together. The second is worth stating carefully, because the components are orthogonal *as sources* by construction while the directions they occupy in representational space need not be: $\gamma$ can reuse the very axes that carry $\alpha$ and $\beta$. Zero on both, with a negligible $\gamma$, is a cleanly separable representation.

**Interpreting the interaction.** Because $\gamma$ is obtained by subtraction it invites reading as a residual in the regression sense, the part a model failed to capture. It is not. The decomposition is saturated: $\mu$, $\alpha$, $\beta$ and $\gamma$ exhaust the degrees of freedom of the grid, so $\gamma$ is the final term of a complete description rather than the leftover of an incomplete one. The residual stratum lies one level below, in the scatter of individual observations *within* each cell. That distinction is not free, however, since cell means are estimated from finitely many observations and within-cell error propagates upward into $\gamma$: a squared length is bounded below by its own noise floor and cannot read zero even when the true interaction is zero. This is the situation in which an analysis of variance tests an interaction against the within-cell error term, and it requires replication for the same reason.

**Replication and the null.** We therefore estimate the grid **twice** from independent data and keep only what the two estimates agree on, reporting the signed inner product $\langle \gamma_A, \gamma_B \rangle$ rather than $\lVert\gamma\rVert^2$. Independent errors cancel in expectation, so this is unbiased at any noise level; and because it can be negative it supports a genuine test, which a squared length cannot. Significance comes from permuting which item of one estimate is paired with which item of the other, which preserves every magnitude while destroying the correspondence. A component's *direction* is only defined if the component exists, so item-class overlap is reported only when both components it relates are present; otherwise it is withheld.

@fig-measurement puts the four parts together: what the decomposition is geometrically, and how each reported quantity is computed from it. Because every cell is observed many times, the two estimates are two disjoint halves of its observations. This is exactly what the language models supply, so the same function is called with the same arguments in both experiments; no alignment between estimates is needed and no projection is applied. What the halves cancel is the difference between the contexts one happened to average over and the contexts the other did, which is the same thing they cancel in a corpus.

![If a word's representation were an item part plus a class part, three corners of a $2 \times 2$ grid of cell means would fix the fourth and the parallelogram would close. It does not. $\gamma$ is how far the actual cell sits from where that additive account puts it, and it is not error: the four components exhaust the grid exactly, so $\gamma$ is the part that belongs to the pairing of a particular word with a particular class. The arrow's length is $\lVert \gamma(i,c) \rVert$, and with two items and two classes it is the whole interaction, since $\gamma$ then has a single degree of freedom and the four cells share one vector up to sign. What we report is each component's squared length as a share of the structure across cells.](fig-measurement.pdf){#fig-measurement width=72%}

**What "recovers" means here.** The weights in @eq-toy describe the *target*; the measure reads the *representation*. These are different objects and their numbers do not match: a learner handed a language that is 85% interaction can build a representation in which the interaction is a far smaller share, because the representation only has to be enough to produce the answer. Validation asks that the measure **track** what was planted, read zero when zero was planted, and have a known floor. It does not ask the two to be equal, and Appendix \ref{sec-appendix-validate} reports the separate check in which the components are planted directly in a representation, where they are.

## Results {#sec-exp1-results}

```{r}
#| label: fig-toy
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 2.5
#| out-width: "100%"
#| fig-cap: "Experiment 1, all 7,560 cells. Left: the measured interaction against the amount planted in the language, with interquartile ranges. The linear condition cannot represent an interaction whatever the language contains, and reads zero throughout; the ReLU condition tracks what was planted. Right: overlap between the item and class effects against capacity, with the dashed line marking what a randomly oriented effect would give at that width. Below capacity the two effects are more orthogonal than chance; once dimension counting makes separation impossible they are several times chance. Note the log axis, and that the floor moves by a factor of eight across these widths."
ACT2 <- c(linear = UO[["green2"]], ReLU = UO[["gold"]])
# fill duplicates the colour legend, so it is suppressed rather than collected twice
pA <- ggplot(toy_track, aes(ach, size_int, colour = act, fill = act)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = .16, colour = NA) +
  geom_line(linewidth = .6) + geom_point(size = 1.5) +
  scale_colour_manual(values = ACT2, name = NULL) +
  scale_fill_manual(values = ACT2, guide = "none") +
  labs(x = "interaction planted in the language", y = "interaction measured\nin the representation") +
  theme_paper(8)

# one chance line, not one per condition: the two differ slightly because the class subspace
# rank is estimated, and drawing both makes a zigzag that reads as structure when it is not
floor_line <- toy_cap |> group_by(capacity) |> summarise(floor = median(floor), .groups = "drop")
pB <- ggplot(toy_cap, aes(capacity, overlap, colour = act)) +
  geom_line(data = floor_line, aes(y = floor), inherit.aes = FALSE,
            mapping = aes(capacity, floor), colour = UO[["grey"]], linetype = 2, linewidth = .5) +
  geom_line(linewidth = .6) + geom_point(size = 1.5) +
  annotate("text", x = 0.135, y = 0.032, label = "chance", size = 2.5,
           colour = UO[["grey"]], hjust = 0) +
  scale_colour_manual(values = ACT2, name = NULL) +
  scale_x_log10(breaks = c(.125, .25, .5, 1, 2)) +
  scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  labs(x = "capacity (structure rank / hidden width)", y = "item-class overlap (log)") +
  theme_paper(8)

(pA | pB) + plot_layout(guides = "collect") &
  theme(legend.position = "top", legend.margin = margin(0, 0, -6, 0))
```

**The measure reports zero when the answer is zero.** The linear condition cannot represent an interaction, by the algebra of @sec-exp1-method, so every reading it gives is the estimator's own noise. Across all `r cm(nullget("linear", TRUE, "n"))` cells where the language *did* contain an interaction, it reports a median share of `r g(nullget("linear", TRUE, "med"), 4)` and flags one in `r sprintf("%.1f%%", 100 * nullget("linear", TRUE, "flagged"))` of them, which is the nominal rate for a test at the 5% level. Where no interaction was planted it flags `r sprintf("%.1f%%", 100 * nullget("linear", FALSE, "flagged"))`. That is the calibration, obtained without assuming anything about noise floors: a condition whose true value is provably zero, measured at every width, sample size and noise level in the grid.

**And it tracks what is there.** Under ReLU the measured interaction rises monotonically with the planted one (@fig-toy, left), and the rate at which it is flagged rises with it, from `r sprintf("%.0f%%", 100 * min(toy_track$flagged[toy_track$act == "ReLU" & toy_track$bin != "none"]))` at the smallest nonzero setting to `r sprintf("%.0f%%", 100 * max(toy_track$flagged[toy_track$act == "ReLU"]))` at the largest. With no interaction planted, ReLU flags `r sprintf("%.1f%%", 100 * nullget("ReLU", FALSE, "flagged"))` of cells, *below* nominal: the nonlinearity does not manufacture an interaction out of an additive language.

The measured shares are much smaller than the planted ones, and that is expected rather than a shortfall. The weights describe the target and the measure reads the representation, and a learner need only build enough structure to produce the answer. What matters is that the relationship is monotone and that zero maps to zero.

**Item and class effects entangle only when there is no room for them not to.** This is the substantive result, and it needs the chance level to be visible (@fig-toy, right; @tbl-toy). With capacity well below one, ReLU's item-class overlap is `r g(capget("ReLU", 0.125, "overlap"), 4)` against a chance level of `r g(capget("ReLU", 0.125, "floor"), 4)`: the two effects are *more* orthogonal than two arbitrary directions would be, which is what a learner that has room and uses it looks like. At capacity 1 the overlap is `r g(capget("ReLU", 1, "overlap"), 3)` against a floor of `r g(capget("ReLU", 1, "floor"), 4)`, about `r g(capget("ReLU", 1, "ratio"), 1)` times chance, and at capacity 2 it is `r g(capget("ReLU", 2, "overlap"), 3)`. The linear condition stays at or below chance throughout.

That is superposition, in the setting where we can see it happen: when the number of directions the structure needs exceeds the number available, the components stop being orthogonal because they cannot all be. Nothing about the language changed across those cells, only the room to represent it.

```{r}
#| label: tbl-toy
#| tbl-cap: "Experiment 1, by capacity, over cells where all three effects were planted. Chance is the share of a randomly oriented effect that would fall in the class direction at that width, and the ratio is the observed overlap against it. Below one, a fully separable solution is reachable by dimension counting; above one it is not."
toy_cap |>
  transmute(condition = act, capacity = g(capacity, 3),
            `interaction share` = g(size_int, 3),
            `item-class overlap` = g(overlap, 4),
            chance = g(floor, 4),
            `ratio to chance` = g(ratio, 2)) |>
  arrange(condition, capacity) |>
  kbl(booktabs = TRUE, align = "lrrrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

**The measurement survives the amount of nuisance variation a corpus has.** When context accounts for more than 80% of the target, so that most of what varies within a cell has nothing to do with the item or the class, ReLU's measured interaction holds at `r g(toy_noise$size_int[toy_noise$act == "ReLU" & toy_noise$cbin == ">80%"], 3)` against `r g(toy_noise$size_int[toy_noise$act == "ReLU" & toy_noise$cbin == "<20%"], 3)` when context is minor, and the flag rate falls from `r sprintf("%.0f%%", 100 * toy_noise$flagged[toy_noise$act == "ReLU" & toy_noise$cbin == "<20%"])` to `r sprintf("%.0f%%", 100 * toy_noise$flagged[toy_noise$act == "ReLU" & toy_noise$cbin == ">80%"])`. Power degrades, as it must with twenty-four observations per cell, but the estimate does not drift.

Two things carry into Experiment 2. The instrument reads zero when the answer is zero and tracks the answer when it is not, at the sample size and noise level the language models will be measured at. And whether the item and class effects come out orthogonal is governed by whether the representation had room, which is a fact about capacity rather than about the language, so a finding of near-orthogonality in a wide model is uninformative on its own and has to be read against its own chance level.


# Experiment 2: Item and Category in Language Models {#sec-exp2}

Experiment 1 established that a learner fitting an interactive language cannot represent it as item plus class, and that a nonlinearity additionally makes the components non-orthogonal. Experiment 2 applies the same decomposition to the contextual representations of pretrained language models, across three linguistic distinctions chosen to span grammar and meaning.

## Dataset {#sec-exp2-data}

**Corpora.** For part of speech and grammatical role we concatenate six general-domain English Universal Dependencies treebanks [@nivre2020ud]: EWT, GUM, LinES, ParTUT, GENTLE and PUD, giving `r cm(41095)` sentences and `r cm(706323)` tokens. Concatenating rather than using EWT alone is what makes the analysis possible: a lemma must be attested in both class levels often enough to estimate a cell mean, and a single treebank yields too few such lemmas. For metaphor we use the VUA20 corpus [@leong2020vua; @steen2010vua], restricted to content words, which supplies `r cm(gs("metaphor", "n_pre_grid_tokens"))` labeled tokens.

**Constructions.** All three are **same-token**: the item's surface form is identical across the two class levels, so the representation is read from the same token in both, and no part of any difference can come from tokenisation.

| distinction | item | levels | analyzed | tokens |
|:-----------|:-----------|:-----------|-----------:|-----------:|
| part of speech | surface form | [noun]{.smallcaps} / [verb]{.smallcaps} | `r max(llm$n_items[llm$construction == "POS (noun/verb)"])` | `r cm(4722)` |
| grammatical role | surface form | *nsubj* / *obj* | 49 | `r cm(2150)` |
| metaphor | surface form | literal / figurative | `r gs("metaphor", "n_items_kept")` | `r cm(gs("metaphor", "n_grid_tokens"))` |

: The three constructions. All three key the item on the **surface form**, so the token is identical at both levels of the distinction. *Analyzed* counts the forms attested often enough at each level to estimate a cell mean. {#tbl-constructions}

**Why the item is the surface form, including for part of speech.** Keying on the *lemma* admits more items (`r cm(gs("pos_noun_verb", "n_items_kept"))` rather than `r max(llm$n_items[llm$construction == "POS (noun/verb)"])`, drawn from `r cm(gs("pos_noun_verb", "n_items_eligible"))` lemmas attested as both), but a lemma pools every inflected form: *run* and *runs* as a noun against *run*, *runs*, *ran* and *running* as a verb. Those are different tokens, and the consequence shows up where it must, at the embedding layer, which has no context to work with and so can only distinguish levels that differ in their input. Keyed on the lemma the layer-0 interaction is 0.083 to 0.172 across models; keyed on the form it is 0.0010 to 0.0028. The difference is inflection rather than representation, and keying on the form removes it. We pay for that in items, and it is worth paying. Appendix \ref{sec-appendix-morph} shows what the measure does on a distinction where the property cannot be arranged at all.

**Identifiability.** The interaction $\gamma(i,c)$ is only estimable for an item observed at more than one level, since it is defined by how an item's representation shifts across levels. We additionally require at least 10 tokens per cell, which is what allows a cell mean to be estimated and then split in half for the replication the measure needs (@sec-exp1-method). This restricts the scope: for part of speech it selects forms genuinely used both ways, such as *run* or *hit*. Role and metaphor carry no such restriction, since any noun occurs as both subject and object and any content word occurs both literally and figuratively, which is one reason they broaden the claim. The median cell holds around 24 tokens across the three constructions, which is the sample size the measure is validated at in Appendix \ref{sec-appendix-validate}.

For role and metaphor the item is keyed on the **surface form** rather than the lemma, which guarantees the identical token across levels: a lemma like *thing* can differ in number across the two roles, whereas *things*-as-subject and *things*-as-object are the same string.

**Models.** Two families at three matched scales: OPT-BabyLM, trained on developmentally plausible quantities of child-available language [@warstadt2023babylm], and Pythia, trained on the Pile [@biderman2023pythia], at 125M/160M, 350M/410M and 1.3B/1.4B parameters, with model widths of 768, 1024 and 2048. Appendix \ref{sec-appendix-models} gives parameter counts and training-token estimates.

## Method {#sec-exp2-method}

The decomposition, the two reported quantities and the replication logic are exactly those of @sec-exp1-method. Three things differ.

**Representations.** For every token we take the hidden state of the word's **last subword** at every layer, from the embeddings through the final block, aligning subwords to corpus tokens with the tokenizer's word alignment. Sentences are drawn in a fixed random order and processed until roughly 300K tokens are extracted per model, so all models see the same material. Representations are extracted once and every analysis runs offline from them.

**Where the two independent estimates come from.** Each cell is a word used at one level across many sentences, so we split its **token contexts** into two disjoint halves and estimate the grid from each. What cancels is context noise, the variation in a word's hidden state across the sentences it appears in. This is the same split Experiment 1 makes over an artificial cell's observations (@sec-exp1-method), which is why the two experiments call the same function with the same arguments: both halves come from one frozen model, so no alignment between the estimates is needed.

**Standardization.** Transformer hidden states contain a few *rogue* dimensions whose variance exceeds the rest by two to three orders of magnitude [@timkey2021bark], and these would otherwise dominate any subspace estimated from the raw scatter. We divide each dimension by its standard deviation before decomposing. This is a diagonal rescaling, so unlike whitening it needs no favorable ratio of samples to dimensions; the measure is accordingly invariant to per-dimension rescaling but not to arbitrary linear reparameterization, which we return to in Appendix \ref{sec-appendix-math}.

## Results {#sec-exp2-results}

```{r}
#| label: fig-llm
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 2.4
#| out-width: "100%"
#| fig-cap: "Experiment 2. Left: the interaction's share of representational structure, by relative depth (0 = static embeddings, 1 = final block). Right: the class effect's share on the same axis. Each line is one model; color distinguishes the two training corpora. Both components are built by depth, and the interaction is the larger of the two throughout."
plt <- llm |>
  select(construction, family, size, depth, size_int, size_class) |>
  pivot_longer(c(size_int, size_class), names_to = "component", values_to = "share") |>
  mutate(component = factor(component, c("size_int", "size_class"),
                            c("interaction", "class effect")))
ggplot(plt, aes(depth, share, colour = family, group = interaction(family, size))) +
  geom_line(linewidth = .55, alpha = .85) +
  facet_grid(component ~ construction, scales = "free_y", switch = "y") +
  scale_colour_manual(values = c(BabyLM = UO[["gold"]], Pythia = UO[["blue"]]), name = NULL) +
  scale_x_continuous(breaks = c(0, .5, 1)) +
  labs(x = "relative depth", y = NULL) +
  theme_paper(8) + theme(strip.placement = "outside")
```

```{r}
#| label: tbl-llm
#| tbl-cap: "Experiment 2, at each model's deepest layer. The two overlaps are the two ways separability can fail by orientation, each with its own chance level in parentheses: item-class overlap is the fraction of the item effect lying in the class subspace (chance $= r_{\\text{class}}/d$), and interaction overlap is the fraction of the interaction lying in the span of the two main effects (chance $= r_{\\text{margin}}/d$, much higher because that subspace has rank of order the item count). The portable fraction is the class effect's share of the class-relevant structure, `size_class / (size_class + size_interaction)`: how much of what distinguishes the two class levels is a direction shared across words rather than specific to each word."
llm |> group_by(construction, model) |>
  filter(layer == max(layer)) |> ungroup() |>
  mutate(portable = size_class / (size_class + size_int),
         # chance for each overlap is the target subspace's rank over the width: the class
         # subspace has rank C-1 = 1, the marginal subspace (C-1) + (I-1).
         ch_leak = 1 / d, ch_int = (1 + n_items - 1) / d) |>
  group_by(construction) |>
  summarise(`item-class overlap` = {v <- median(leak, na.rm = TRUE)
               if (is.na(v)) "---" else sprintf("%.3f (%.3f)", v, median(ch_leak))},
            `interaction overlap` = {v <- median(leak_int, na.rm = TRUE)
               if (is.na(v)) "---" else sprintf("%.3f (%.3f)", v, median(ch_int))},
            `class effect` = sprintf("%.3f", median(size_class)),
            `interaction` = sprintf("%.3f", median(size_int)),
            `portable fraction` = sprintf("%.2f [%.2f, %.2f]", median(portable),
                                          min(portable), max(portable)),
            .groups = "drop") |>
  rename(distinction = construction) |>
  kbl(booktabs = TRUE, align = "lrrrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

**The two main effects sit on separate axes.** At the deepest layer, item-class overlap runs `r g(0.0032, 4)` to `r g(0.0067, 4)` for part of speech and `r g(0.0025, 4)` to `r g(0.0062, 4)` for grammatical role, so projecting out the class direction leaves better than 99% of the item effect intact. Those values are two to seven times what a randomly oriented effect would give at these widths, which Appendix \ref{sec-appendix-validate} measures directly rather than deriving, so they are not exactly zero. But they are small enough that these representations plainly do not achieve class structure by borrowing the directions that carry word identity. Where a shared class direction exists, it is its own direction.

**The interaction does not have axes of its own, however.** This is the second way orientation can fail, and it is the one that matters for whether a third component could simply be added to the description and separability recovered in a wider sense. It cannot: `r g(0.191, 3)` of the interaction lies in the span of the two main effects for part of speech, `r g(0.131, 3)` for role and `r g(0.227, 3)` for metaphor, against chance levels of `r g(0.076, 3)`, `r g(0.048, 3)` and `r g(0.176, 3)` (@tbl-llm). The components are orthogonal *as sources* by construction, so this is not guaranteed by the decomposition and is not a restatement of it: it says the item-by-class term reuses the very directions that carry item identity and class membership. A representation in which the third component occupied a subspace of its own would be entangled in an accounting sense but not a geometric one. These are entangled in both.

Read together, the two overlaps say something more specific than "not separable." The failure is not that item and class have been collapsed onto shared axes; those remain distinct. It is that the part of class structure which is specific to each word is written in the same directions as everything else.

For metaphor the question does not arise. The overlap is not reported at **every** layer of every model, because the class effect never rises above the level at which an orientation is worth reporting. That is the measure declining to describe the direction of something that is barely there, and it is itself the first indication of how little of metaphoricity is carried by a shared direction.

**A substantial interaction is there, and depth builds it.** In all three distinctions the interaction is essentially absent at layer 0 -- between 0.0002 and 0.0051 across every construction and model -- and rises monotonically with depth to between 0.023 and 0.246 (@fig-llm, top row). Because the item is keyed on the surface form, the token is *identical* at both class levels, so the static embedding layer has nothing to distinguish them with, and it does not. Everything the measure finds is therefore created by contextualization rather than inherited from the input. Nor is it positional: zeroing OPT-BabyLM's absolute position embeddings leaves the class structure *more* word-specific rather than less (Appendix \ref{sec-appendix-positions}). This is the component
that breaks separability: it is neither a fact about the word nor a fact about the class, so no
decomposition into those two can absorb it.

**So does the class effect, and that matters for the framing.** The lower row of @fig-llm shows
the class effect growing with depth as well. Models are not choosing between abstraction and
entanglement; contextualization builds both at once. A description of these representations as either
"abstract" or "exemplar-like" is therefore underdetermined by the data, and the useful question is not
which one they are but what the balance is.

**The balance is graded across distinctions.** @tbl-llm gives that balance as the *portable fraction*:
of the structure that distinguishes the two class levels, how much is a direction shared across
words rather than specific to each word. Grammatical role is the most portable at 0.46, part of speech next at 0.31, and metaphor by far the least at 0.07: only about a fourteenth of its class-relevant structure is carried by a shared direction. Note the ordering. Part of speech has the *largest* class effect of the three in absolute terms, but also much the largest interaction, so proportionally less of what distinguishes its levels is portable than for grammatical role. Metaphoricity is close to being encoded word by word: what makes a use of *plant*
figurative is largely not what makes a use of *course* figurative.

# Experiment 3: Does the Interaction Carry the Class? {#sec-exp3}

Experiment 2 shows the interaction is large and that it replicates across independent halves of a word's occurrences. Neither shows what it is *about*. An item-by-class term captures any reliable difference between a word's uses at the two levels, and some of that difference is context correlated with the label rather than the label itself: metaphorical uses occur in different sentences from literal ones, subjects appear in different constructions from objects. So $\gamma$ is an upper bound on class-binding, and the question is whether any of it is the class.

## Method {#sec-exp3-method}

We test it by decoding. If $\gamma$ encodes the class, then the class should be recoverable from the part of a token's representation that is neither the item effect nor the shared class direction. Both are removed explicitly:

**Hold the item fixed.** All decoding happens *within* a single word: we ask whether *plant*'s figurative uses can be told from *plant*'s literal ones, never whether *plant* can be told from *account*. The item effect $\alpha$ is constant within a word, so it cannot contribute.

**Project out the shared class direction.** We estimate $\beta$ as the difference between the two class means pooled over all items, normalize it, and remove that component from every token. Whatever separates the classes afterwards is not the direction they share.

What remains is the interaction, $\gamma$: the part of a word-at-a-level that is neither a fact about the word nor a fact about the level. We decode it with a cross-validated logistic probe, five folds, classes balanced by subsampling within each word so that chance is 0.50, and average the resulting accuracies over words. Words contribute only if they have at least eight tokens at each level.

Two guards. Accuracies are averaged **unweighted** over words and some words rest on few tokens, so the mean carries a percentile bootstrap interval over words rather than being reported alone. And the whole procedure is repeated with the class labels **permuted within each word**, which preserves every word's token count and class balance while destroying the labelling; that null is where a probe reading its own cross-validation folds would land.

This is run at the deepest layer of the largest model in each family, for all three constructions.

## Results {#sec-exp3-results}

```{r}
#| label: tbl-decode
#| tbl-cap: "Class decoded from the interaction, at the deepest layer of the largest model in each family. The item is held fixed and the shared class direction is projected out, so neither the item effect nor the class effect can contribute. Brackets are percentile bootstrap intervals over words; the null permutes class labels within each word. Chance is 0.50."
dec <- bind_rows(
  rd("llm_decode_pos_form.csv") |> mutate(construction = "POS (noun/verb)"),
  rd("llm_decode_interaction.csv") |> filter(construction != "pos") |>
    mutate(construction = recode(construction,
                                 role = "role (subject/object)",
                                 metaphor = "metaphor (literal/figurative)"))) |>
  left_join(MODELS, by = "model") |>
  mutate(construction = factor(construction, c("POS (noun/verb)", "role (subject/object)",
                                               "metaphor (literal/figurative)")))

dec |>
  transmute(distinction = construction, family,
            words = n_peritem,
            `class from the interaction` =
              sprintf("%.3f [%.3f, %.3f]", class_peritem, class_peritem_lo, class_peritem_hi),
            `permuted null` =
              sprintf("%.3f [%.3f, %.3f]", class_peritem_null,
                      class_peritem_null_lo, class_peritem_null_hi)) |>
  arrange(distinction, family) |>
  kbl(booktabs = TRUE, align = "llrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

**The interaction is about the class.** In every construction and both families the class is recoverable from the interaction at well above chance, and the permuted null sits on chance (@tbl-decode). For part of speech it reaches `r g(dec$class_peritem[dec$construction == "POS (noun/verb)" & dec$family == "Pythia"], 2)`, for grammatical role `r g(dec$class_peritem[dec$construction == "role (subject/object)" & dec$family == "Pythia"], 2)`, and for metaphor `r g(dec$class_peritem[dec$construction == "metaphor (literal/figurative)" & dec$family == "Pythia"], 2)`. The bootstrap intervals exclude the nulls by a wide margin in all six cases.

That the null lands on chance is the honesty check rather than a formality. It is the same probe, the same folds, the same word-level averaging and the same small per-word samples, with only the labelling destroyed. Any accuracy the procedure could manufacture from its own structure would appear there, and none does.

**The effect is graded the same way the interaction's size is.** Part of speech decodes best, role next, metaphor least, which is the ordering of the portable fraction in @tbl-llm read backwards: the distinctions whose class-relevant structure is least portable are also the ones whose interaction carries the class least strongly. Two measurements, one geometric and one functional, agreeing on the same ranking.

Metaphor is the case where this matters most. Its class effect never clears the threshold at which an orientation is worth reporting (@sec-exp2-results), so on the geometric evidence alone one could conclude that metaphoricity is not represented at all. It decodes from the interaction at `r g(dec$class_peritem[dec$construction == "metaphor (literal/figurative)" & dec$family == "Pythia"], 2)` against a null of `r g(dec$class_peritem_null[dec$construction == "metaphor (literal/figurative)" & dec$family == "Pythia"], 2)`. It is represented; it is simply not represented as a direction shared across words. A method that looks only for a shared direction would have missed it.

**What this does and does not establish.** It shows $\gamma$ carries the specific label rather than being generic correlated context, since generic context would not survive holding the word fixed and removing the shared axis. It does not show that $\gamma$ is *only* the label. Context correlated with the class within a word survives both removals, so the interaction remains an upper bound on class-binding, and this result raises the floor rather than closing the gap.

# Discussion {#sec-discussion}

::: {.content-visible when-format="acl-pdf"}
```{=html}
<!-- PLACEHOLDER: author to write. -->
```
:::

# Limitations {.unnumbered}

Everything here is English, and rests on three distinctions measured in two model families at three scales each. The graded pattern we report is a pattern across *these* distinctions, and whether the same ordering holds for aspect, animacy or definiteness is untested.

The one property of the measure that constrains interpretation is that it is invariant to per-dimension rescaling and to rotation but not to arbitrary linear reparameterization, because we do not whiten. Overlap between the item and class effects is therefore relative to the model's own basis. We take that to be the basis of interest, since LayerNorm and the feed-forward nonlinearities act coordinate-wise and a downstream readout operates on the unwhitened space, but a reader who wants a fully gauge-free quantity will not find one here. The component sizes, their growth with depth and their significance do not depend on that choice.

Two things we would like to have done. These are endpoint measurements, so they say nothing about the order in which the two components are learned; that question needs training checkpoints and we leave it to future work. And the interaction captures any systematic within-word difference between the levels, so a distinction that is realized through the contexts a word appears in is measured together with those contexts. For a contextual model we take that to be what representing the distinction consists in rather than a confound, but separating them would need a construction in which the two can be varied independently.

```{=latex}
\bibliography{refs}
\clearpage
```

\onecolumn

\appendix

```{=latex}
% Appendix layout, following the same reasoning as the companion papers.
%
% \raggedbottom matters most. acl.sty sets \flushbottom, which forces each page to exact
% height by stretching the glue between paragraphs and around headings. On an appendix page
% that cannot fill naturally that stretch becomes a gap in the middle of the page with text
% stranded either side. \raggedbottom lets the slack collect at the foot instead. Main-text
% pages keep \flushbottom and the two-column setting.
%
% Appendix tables are longtables, so they typeset inline and break across pages rather than
% floating away from the prose that introduces them. Figures stay floats, pinned with [H].
\raggedbottom
\floatplacement{figure}{H}
```

# The Decomposition and Its Estimator {#sec-appendix-math}

## The decomposition as a projection {#sec-appendix-math-proj}

For items $i \in \{1,\dots,I\}$ crossed with class levels $c \in \{1,\dots,C\}$, let $M[i,c] \in \mathbb{R}^d$ be the mean representation of item $i$ at level $c$. Regard the grid as an element of $\mathbb{R}^{I}\otimes\mathbb{R}^{C}\otimes\mathbb{R}^{d}$ and define averaging and centering projectors on the first two factors:

$$
P_I = \tfrac{1}{I}\mathbf{1}_I\mathbf{1}_I^{\top}, \quad Q_I = \mathrm{Id}_I - P_I,
\qquad
P_C = \tfrac{1}{C}\mathbf{1}_C\mathbf{1}_C^{\top}, \quad Q_C = \mathrm{Id}_C - P_C .
$$

The four components of @eq-decomp are the four tensor-product projections of $M$:

$$
\mu = (P_I\!\otimes\! P_C)M, \quad
\alpha = (Q_I\!\otimes\! P_C)M, \quad
\beta = (P_I\!\otimes\! Q_C)M, \quad
\gamma = (Q_I\!\otimes\! Q_C)M ,
$$

which sum to $M$ because $(P_I{+}Q_I)\otimes(P_C{+}Q_C) = \mathrm{Id}$. They are mutually orthogonal idempotents, so this is an orthogonal direct sum with dimensions $1$, $I{-}1$, $C{-}1$ and $(I{-}1)(C{-}1)$. This is a geometric construction rather than a statistical model: there is no error stratum, no $F$-ratio and no distributional assumption.

Two consequences are used in the body. First, the side constraints $\sum_i\alpha_i = 0$, $\sum_c\beta_c = 0$ and $\sum_i\gamma_{ic} = \sum_c\gamma_{ic} = 0$ follow from the projectors rather than being imposed. Second, an **additive** grid, one of the form $M[i,c] = A(i) + B(c)$, has $\gamma = 0$ *identically*: writing $\bar A$ and $\bar B$ for the averages, $\mu = \bar A + \bar B$, $\alpha_i = A(i) - \bar A$, $\beta_c = B(c) - \bar B$, and the remainder cancels exactly. This is what makes the linear condition of Experiment 1 a negative control rather than an assumption.

## Sizes {#sec-appendix-math-sizes}

Orthogonality gives an exact partition of the between-cell energy,

$$
\sum_{i,c}\lVert M[i,c]-\mu\rVert^{2}
= \underbrace{C\sum_{i}\lVert\alpha_i\rVert^{2}}_{s_{\text{item}}}
+ \underbrace{I\sum_{c}\lVert\beta_c\rVert^{2}}_{s_{\text{class}}}
+ \underbrace{\sum_{i,c}\lVert\gamma_{ic}\rVert^{2}}_{s_{\text{int}}},
$$

with $C$ and $I$ appearing as replication counts: $\alpha_i$ is shared across $C$ cells and $\beta_c$ across $I$. Sizes are reported as shares of the total. All three terms scale linearly in $I$, so the shares do not depend on how many items the grid contains and are comparable across constructions of different size.

## Item-class overlap and its chance level {#sec-appendix-math-overlap}

Subspaces are orthonormal bases from the leading right singular vectors, with rank set by the participation ratio $r = \operatorname{round}\!\big((\sum_k\lambda_k)^2/\sum_k\lambda_k^2\big)$ of the scatter eigenvalues, capped at the algebraic maxima above. With two class levels $\beta$ spans one direction, so $k_{\text{class}} = 1$ throughout Experiment 2.

Overlap is the energy fraction inside a subspace, $\lVert V S\rVert_F^2 / \lVert V\rVert_F^2$. Its chance level is not zero: for a randomly oriented $V$,

$$
\mathbb{E}\big[\lVert V S\rVert_F^2\big] = \operatorname{tr}(V^{\top}V)/d \cdot r
\;\Longrightarrow\;
\mathbb{E}[\text{overlap}] = r/d ,
$$

whatever $V$'s rank or shape. At $d = 2048$ that is $0.0005$, and Appendix \ref{sec-appendix-validate} measures it directly at $0.0010$. Any overlap has to be read against that number, which moves by two orders of magnitude across the widths used in this paper.

## Why the interaction is estimated as a cross-product {#sec-appendix-math-cross}

Cell means are estimated from finitely many observations, $\hat M = M + E$ with $\mathbb{E}[E]=0$. The decomposition is linear, so $\hat\gamma = \gamma + (Q_I\otimes Q_C)E$ and

$$
\mathbb{E}\big[\lVert\hat\gamma\rVert^2\big] = \lVert\gamma\rVert^2 + (I{-}1)(C{-}1)\,d\,\sigma^2 .
$$

A squared length is therefore bounded below by its own noise floor and cannot read zero even when the true interaction is zero. Permutation does not repair this: a squared projection norm is a magnitude, its expectation under the null is that same floor, and the permuted replicates carry the floor too, so real interaction and junk of equal size give the same statistic.

Given two independent estimates $\hat\gamma_A = \gamma + G_A$ and $\hat\gamma_B = \gamma + G_B$ with $G_A \perp\!\!\!\perp G_B$,

$$
\mathbb{E}\langle\hat\gamma_A,\hat\gamma_B\rangle
= \lVert\gamma\rVert^2 + \langle\gamma,\mathbb{E}G_B\rangle + \langle\mathbb{E}G_A,\gamma\rangle
+ \langle\mathbb{E}G_A,\mathbb{E}G_B\rangle
= \lVert\gamma\rVert^2 ,
$$

exactly unbiased at any noise level, and able to come out negative, which is what makes $>0$ a test rather than a threshold on a positively biased quantity. Significance permutes the item index of one estimate against the other, preserving every magnitude while destroying the correspondence, over 200 draws at the 5% level.

The same construction is applied to the item effect. The **class effect is not cross-denoised**: it is reported as $I\sum_c\lVert\hat\beta_c\rVert^2$ from the full grid, on the grounds that $\hat\beta_c$ averages over all $I$ items and is well estimated. That addresses variance rather than bias, and Appendix \ref{sec-appendix-validate} measures the floor it leaves.

## Invariance {#sec-appendix-math-invariance}

The measure is exactly invariant to per-dimension rescaling, which the standardization absorbs, and the functionals are invariant to orthogonal rotation of already-standardized data. It is **not** invariant to general linear reparameterization: we do not whiten. Whitening would make a linear representation read separable in any basis, but it requires estimating a $d \times d$ covariance and it discards the basis the model actually computes in. LayerNorm and the feed-forward nonlinearities act coordinate-wise, and a downstream readout operates on the unwhitened space, so whether item and class occupy separate axes *in the model's own basis* is the question of interest rather than a compromise. The consequence to state plainly is that overlap values are basis-relative; the sizes, their growth with depth and their cross-denoised significance are the parts that do not depend on the choice.

# Validating the Measure {#sec-appendix-validate}

Experiment 1 plants structure in a *language* and measures a trained model's representation. A discrepancy there is ambiguous between the measure missing something and the model not mirroring the language, and the second is known to happen: the linear condition cannot represent an interaction at all. This appendix removes the ambiguity by planting $\mu$, $\alpha$, $\beta$ and $\gamma$ **directly in a representation**, with no model and no training, and passing the result to the same function the language models are measured with. Ground truth now sits at the level the measure operates on.

```{r}
#| label: setup-validate
#| include: false
val <- rd("validate_measure.csv")
v_rec <- val |> filter(n_obs == 24, d >= 768) |> group_by(planted_int) |>
  summarise(n = n(), measured = median(size_interaction),
            lo = quantile(size_interaction, .05), hi = quantile(size_interaction, .95),
            flagged = mean(sig_interaction), .groups = "drop")
v_null <- bind_rows(
  val |> filter(planted_int == 0) |> group_by(n_obs) |>
    summarise(effect = "interaction", med = median(size_interaction),
              p95 = quantile(size_interaction, .95),
              flagged = mean(sig_interaction), .groups = "drop"),
  val |> filter(planted_class == 0) |> group_by(n_obs) |>
    summarise(effect = "class effect", med = median(size_class),
              p95 = quantile(size_class, .95), flagged = NA_real_, .groups = "drop"),
  val |> filter(planted_item == 0) |> group_by(n_obs) |>
    summarise(effect = "item effect", med = median(size_item),
              p95 = quantile(size_item, .95), flagged = NA_real_, .groups = "drop"))
v_ov <- val |> filter(!is.na(leak_item_into_class), n_obs == 24, planted_overlap == 0) |>
  group_by(d) |> summarise(measured = median(leak_item_into_class),
                           chance = median(k_class / d), .groups = "drop")
vget <- function(eff, obs, col) v_null[[col]][v_null$effect == eff & v_null$n_obs == obs]
```

**Construction.** The three components are drawn at random and then centered so each satisfies the constraints its role implies: $\alpha$ and $\beta$ sum to zero over their own index, $\gamma$ over both. Without that a planted $\gamma$ would partly be an item or class main effect and the planted shares would be wrong before the measure saw them. Each is rescaled to hit a target share of the between-cell energy under the same weighting the measure uses. Cell means are then observed $n$ times each under isotropic noise, whose scale is set as a multiple of the mean per-cell energy. The grid crosses the planted shares (the same factorial Experiment 1 runs, mapped into share space, plus the three compositions Experiment 2 measures), item counts from 27 to 180, two and four class levels, widths from 8 to 2048, 10, 24 and 100 observations per cell, three noise levels, planted-orthogonal and planted-non-orthogonal effects, and five seeds: `r cm(nrow(val))` runs.

**Recovery.** At Experiment 2's regime, 24 observations per cell in 768 or more dimensions, the measured interaction matches the planted one across the whole range (Table \ref{tbl-val-recovery}), and is flagged in essentially every cell where anything was planted and none where nothing was.

```{r}
#| label: tbl-val-recovery
#| tbl-cap: "Recovery at 24 observations per cell in 768 or more dimensions, pooled over noise levels, item counts and widths. Brackets are the 5th and 95th percentiles across runs."
v_rec |> filter(planted_int == 0 | planted_int > 0.02) |> head(12) |>
  transmute(`planted` = g(planted_int, 3),
            runs = n,
            `measured` = sprintf("%.4f [%.4f, %.4f]", measured, lo, hi),
            `flagged` = sprintf("%.0f%%", 100 * flagged)) |>
  kbl(longtable = TRUE, booktabs = TRUE, align = "rrrr", linesep = "") |>
  kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

**Floors.** The interaction and the item effect read essentially zero when nothing is planted, at every sample size. The class effect does not, and this is the one quantity in the paper that carries a floor worth knowing (Table \ref{tbl-val-null}): at 24 observations per cell a true zero reads `r g(vget("class effect", 24, "med"), 4)`. That is the price of estimating the class effect as a squared length rather than a cross-product, and it is why the metaphor result in @sec-exp2-results is stated as an inability to distinguish the class effect from zero rather than as a measurement of something small. Metaphor's observed class effect, 0.0016 to 0.0062 at the deepest layer, sits at that floor.

```{r}
#| label: tbl-val-null
#| tbl-cap: "What the measure reports when an effect is planted at exactly zero, by observations per cell, pooled over every other axis. The interaction is flagged well below the nominal 5%; the class effect is not cross-denoised and carries a floor that shrinks with sample size."
v_null |>
  transmute(effect, `obs per cell` = n_obs,
            median = g(med, 5), `95th pct` = g(p95, 4),
            flagged = ifelse(is.na(flagged), "---", sprintf("%.1f%%", 100 * flagged))) |>
  arrange(effect, `obs per cell`) |>
  kbl(longtable = TRUE, booktabs = TRUE, align = "lrrrr", linesep = "") |>
  kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

**The chance level for overlap.** With the item and class effects planted orthogonal, the measured overlap tracks $r/d$ closely and sits a little below it (Table \ref{tbl-val-overlap}). This is the number Experiment 2's overlaps are read against: at $d = 2048$ a genuinely orthogonal pair reads `r g(v_ov$measured[v_ov$d == 2048], 5)`, and the values reported in @sec-exp2-results are two to seven times that.

```{r}
#| label: tbl-val-overlap
#| tbl-cap: "Measured item-class overlap when the two effects are planted orthogonal, at 24 observations per cell. The chance column is the analytic level of Appendix \\ref{sec-appendix-math}."
v_ov |>
  transmute(`width d` = d, measured = g(measured, 5), `analytic chance` = g(chance, 5)) |>
  kbl(longtable = TRUE, booktabs = TRUE, align = "rrr", linesep = "") |>
  kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

# A Construction That Is Not Same-Token {#sec-appendix-morph}

Every construction in the body keys the item on the surface form, so the token is identical at both levels of the distinction and the embedding layer has nothing to distinguish them with. That property is what licenses reading the depth profile as contextualization. It is worth seeing what the same measure does when the property fails.

Number and tense supply the test. A plural or a past tense is spelled with an extra subword, so *dog* and *dog-s* are different tokens and the distinction is present in the input before any contextualization occurs.

```{r}
#| label: tbl-morph
#| tbl-cap: "Number and tense, which are marked by an affix and therefore read at a different subword at the two levels. Unlike every construction in the body, the interaction is large at layer 0 and does not grow, which is what a distinction visible in the input looks like."
rd("llm_morph.csv") |>
  filter(regularity == "regular") |>
  group_by(model, feature) |>
  summarise(L0 = size_interaction[layer == 0],
            deep = size_interaction[layer == max(layer)], .groups = "drop") |>
  left_join(MODELS, by = "model") |>
  transmute(family, model = sub(".*/", "", model), feature,
            `layer 0` = g(L0, 3), `deepest` = g(deep, 3),
            `ratio` = g(deep / L0, 1)) |>
  arrange(feature, family, model) |>
  kbl(longtable = TRUE, booktabs = TRUE, align = "lllrrr", linesep = "") |>
  kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

The profile inverts (Table \ref{tbl-morph}). Where the same-token constructions begin near zero at layer 0 and grow by one to two orders of magnitude, number and tense begin large and grow barely or not at all. That is what a distinction carried by the input rather than built by the network looks like under this measure, and it is the pattern the body's constructions were designed to avoid.

It also illustrates why the surface-form keying of part of speech mattered. Keyed on the lemma, part of speech looked like this: a large layer-0 interaction and, for the BabyLM models, no growth. Keyed on the form it looks like role and metaphor instead. The measure was reporting the tokenization correctly in both cases; what changed was which question it was being asked.

# Model Details {#sec-appendix-models}

```{r}
#| label: tbl-models
#| tbl-cap: "The six models, paired by width. Training-token counts are approximate; the OPT-BabyLM models train for 20 epochs over a 100M-word corpus, so their effective token exposure is roughly twenty times the corpus size. Layers counts the transformer blocks, and the measure is applied to each block's output plus the embedding layer."
tibble::tribble(
  ~family,  ~model,          ~params, ~width, ~layers, ~`training data`,     ~`tokens`,
  "BabyLM", "opt-babylm-125m", "125M",   768,      12, "BabyLM (child-available)", "~2B",
  "BabyLM", "opt-babylm-350m", "350M",  1024,      24, "BabyLM (child-available)", "~2B",
  "BabyLM", "opt-babylm-1.3B", "1.3B",  2048,      24, "BabyLM (child-available)", "~2B",
  "Pythia", "pythia-160m",     "160M",   768,      12, "the Pile",                 "~300B",
  "Pythia", "pythia-410m",     "410M",  1024,      24, "the Pile",                 "~300B",
  "Pythia", "pythia-1.4b",     "1.4B",  2048,      24, "the Pile",                 "~300B") |>
  kbl(longtable = TRUE, booktabs = TRUE, align = "llrrrlr", linesep = "") |>
  kable_styling(font_size = 9, latex_options = c("repeat_header")) |>
  collapse_rows(columns = 1, latex_hline = "major")
```

The two families are matched on width rather than on parameter count, which is what the measure is sensitive to: the chance level for overlap is the class subspace rank over the width (Appendix \ref{sec-appendix-math}), so a comparison across models of different width has to be read against different chance levels. Pairing at 768, 1024 and 2048 removes that from the family comparison.

They differ in two ways that matter elsewhere in the paper. The training corpora differ by more than two orders of magnitude in size, which is the manipulation of interest. And they position tokens differently: Pythia uses rotary embeddings, applied inside attention, while OPT learns absolute position embeddings added at the input. Appendix \ref{sec-appendix-positions} uses that difference to rule out absolute position as the source of the interaction.

# Positional Embeddings Are Not the Source {#sec-appendix-positions}

One architectural explanation of the interaction has to be ruled out before it can be read as a fact about lexical representation. OPT-BabyLM adds a *learned absolute position embedding* at the input. The two levels of a linguistic distinction are not positionally interchangeable in a sentence, since subjects precede objects and nouns and verbs occupy different slots, so a representation that varies with position varies with the class. If that is what the decomposition is picking up, the interaction would be an artifact of where words sit rather than of how they are represented.

We re-extracted every OPT-BabyLM model with `decoder.embed_positions` zeroed and measured it identically. The ablation was verified per model rather than assumed: with the embedding zeroed, the same token at two different offsets must give identical layer-0 states, and it does, in every model (drift $0.0$ to machine precision). Pythia has no such module, so there is nothing to zero there.

The right quantity is the *portable fraction* used in @tbl-llm, the class effect's share of class-relevant structure. If positions were manufacturing the interaction, removing them should leave a larger share of what distinguishes the two levels in a direction shared across words. Absolute sizes cannot be compared here: zeroing the embeddings in a model that was trained with them is a heavy intervention, and it shrinks every component at once, the class effect included.

```{r}
#| label: tbl-posabl
#| tbl-cap: "Trained OPT-BabyLM at the deepest layer, with absolute position embeddings intact and zeroed. If positions produced the interaction, the portable fraction would rise when they are removed. It falls in all three models. Absolute sizes shrink because zeroing the embeddings damages a model trained with them, which is why the ratio rather than the level is the comparison."
rd("llm_unified_form_ablation.csv") |>
  filter(grepl("babylm", model), grepl("^pretrained", init)) |>
  group_by(model, init) |>
  slice_max(layer, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(model = sub(".*/", "", sub("-20eps-seed964", "", model)),
            positions = ifelse(grepl("noposemb$", init), "zeroed", "intact"),
            class = std_size_class, interaction = std_size_interaction,
            portable = std_size_class / (std_size_class + std_size_interaction)) |>
  arrange(match(model, c("opt-babylm-125m", "opt-babylm-350m", "opt-babylm-1.3B")),
          desc(positions)) |>
  transmute(model, positions, `class` = g(class, 3), `interaction` = g(interaction, 3),
            `portable fraction` = g(portable, 3)) |>
  kbl(longtable = TRUE, booktabs = TRUE, align = "llrrr", linesep = "") |>
  kable_styling(font_size = 9, latex_options = c("repeat_header"))
```

**Removing the position embeddings makes these representations *less* separable, not more.** The portable fraction falls in every model, from 0.439 to 0.264, from 0.468 to 0.253, and from 0.333 to 0.198 (Table \ref{tbl-posabl}). Whatever the absolute position embedding contributes, it is contributing to the part of class structure that *is* carried by a shared direction. Take it away and what remains is more word-specific, not less. The interaction is not an artifact of position.

A second line of evidence points the same way and needs no ablation at all. Pythia never had an absolute position embedding to remove, and its portable fractions are 0.253, 0.287 and 0.262, at or below every intact OPT-BabyLM value. The family without the mechanism shows the phenomenon at least as strongly as the family with it. An explanation resting on absolute position embeddings would have to account for the models that do not have them.

We tested this for part of speech, the construction where the positional worry is sharpest because noun and verb uses of a form occupy systematically different sentence positions. Grammatical role carries the same concern for the same reason, and behaves the same way. Metaphor does not, since literal and figurative uses of a word are not positionally distinguished.

