---
title: "Are Lexical Representations Separable? Item and Category in Learned Distributed Representations"
author:
  - name: "Anonymous"
    affiliations:
      - name: "Anonymous Institution"
    email: anon@example.edu
abstract: |
  PLACEHOLDER --- author to write.
bibliography: refs.bib
acl-mode: review
number-sections: true
tbl-cap-location: bottom
format:
  acl-pdf:
    keep-tex: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(readr); library(dplyr); library(tidyr); library(ggplot2); library(kableExtra)
library(patchwork); library(scales)

DATA <- file.path("..", "data")
rd <- function(f) suppressMessages(read_csv(file.path(DATA, f), show_col_types = FALSE))

# UO palette, as in the companion papers. Green = the linear reference, gold = the
# nonlinearity; amber and blue are held for the model families in Experiments 2-3.
UO <- c(green = "#154733", green2 = "#007030", gold = "#F0B323",
        amber = "#875E00", blue = "#005581", grey = "#6E6E6E")
ACT_COL <- c(identity = UO[["green2"]], relu = UO[["gold"]])
ACT_LAB <- c(identity = "linear", relu = "ReLU")

theme_paper <- function(base = 9) {
  theme_minimal(base_size = base) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(linewidth = 0.25, colour = "grey88"),
          strip.text = element_text(face = "bold", size = base),
          legend.position = "top", legend.margin = margin(0, 0, -4, 0),
          legend.key.height = unit(9, "pt"),
          axis.title = element_text(size = base))
}

# ---- Experiment 1: the toy grid, cross-denoised (two initializations per cell) ----
# ALL 4800 cells are used. Capacity enters as a reported variable, not as a filter: whether
# the representation has room to hold every generative direction on its own axis is part of
# the result, so it is shown rather than assumed.
toy <- rd("experiment9_cross_results.csv") |>
  mutate(band = cut(cap, c(-Inf, .5, 1, Inf),
                    labels = c("room to spare\n(rank/d < 0.5)", "tight\n(0.5-1)",
                               "impossible\n(> 1)")),
         act = factor(activation, c("identity", "relu"), c("linear", "ReLU")))

# Separability fails in two distinguishable ways, and the toy separates them: the presence
# of an interaction term at all, and non-orthogonality among the components. `size_int` is
# the first; `leak_marg` (item marginal inside the category subspace) and `ov_item` (the
# interaction's subspace overlapping the item subspace) are the second.
toy_dose <- toy |>
  group_by(int_frac, act) |>
  summarise(n = n(),
            size_med = median(size_interaction),
            size_lo = quantile(size_interaction, .25),
            size_hi = quantile(size_interaction, .75),
            sig = mean(sig_interaction),
            leak_med = median(leak_item_into_class, na.rm = TRUE),
            leak_n = sum(!is.na(leak_item_into_class)),
            ov_med = median(overlap_item_int, na.rm = TRUE),
            ov_n = sum(!is.na(overlap_item_int)),
            .groups = "drop")

# Interaction geometry is only interpretable where an interaction exists AND the marginals
# do: at int_frac = 0 there is no gamma to orient and at int_frac = 1 no marginals to orient
# against. The gate already withholds those, and they are excluded from the geometry panels.
TOY_MID <- c(.25, .5, .75)
toy_cap <- toy |> filter(int_frac %in% TOY_MID) |>
  group_by(act, band) |>
  summarise(n = n(), size_med = median(size_interaction),
            leak_med = median(leak_item_into_class, na.rm = TRUE),
            leakint_med = median(leak_int_into_margins, na.rm = TRUE),
            ov_med = median(overlap_item_int, na.rm = TRUE), .groups = "drop")

toy_null <- toy |> filter(int_frac == 0) |>
  group_by(act, band) |>
  summarise(n = n(), sig_rate = mean(sig_interaction),
            size_med = median(size_interaction), .groups = "drop")

g <- function(x, digits = 3) formatC(x, format = "f", digits = digits)
pick <- function(tb, a, ifrac, col) tb[[col]][tb$act == a & tb$int_frac == ifrac]
capget <- function(tb, a, b, col) tb[[col]][tb$act == a & grepl(b, tb$band)]
cm <- function(x) format(x, big.mark = ",")

# ---- Experiment 2: language model representations ----
grid_stats <- rd("methods_grid_stats.csv")
gs <- function(constr, col) {
  v <- grid_stats[[col]][grid_stats$construction == constr]
  if (is.numeric(v)) round(median(v)) else v[1]
}

MODELS <- tibble::tribble(
  ~model,                                       ~family,  ~size,   ~d,
  "znhoughton/opt-babylm-125m-20eps-seed964",   "BabyLM", "125M",  768,
  "znhoughton/opt-babylm-350m-20eps-seed964",   "BabyLM", "350M",  1024,
  "znhoughton/opt-babylm-1.3B-20eps-seed964",   "BabyLM", "1.3B",  2048,
  "EleutherAI/pythia-160m",                     "Pythia", "160M",  768,
  "EleutherAI/pythia-410m",                     "Pythia", "410M",  1024,
  "EleutherAI/pythia-1.4b",                     "Pythia", "1.4B",  2048)

# The three same-token constructions carried in the body. Each file reports the same measure;
# the POS file prefixes its columns std_ because it also stores an unstandardized read.
llm <- bind_rows(
  rd("llm_unified.csv") |>
    transmute(model, layer, d, n_items, construction = "POS (noun/verb)",
              size_item = std_size_item, size_class = std_size_class,
              size_int = std_size_interaction, sig = std_sig_interaction,
              leak = std_leak_item_into_class),
  rd("llm_role.csv") |>
    transmute(model, layer, d, n_items, construction = "role (subject/object)",
              size_item, size_class, size_int = size_interaction,
              sig = sig_interaction, leak = leak_item_into_class),
  rd("llm_metaphor.csv") |>
    transmute(model, layer, d, n_items, construction = "metaphor (literal/figurative)",
              size_item, size_class, size_int = size_interaction,
              sig = sig_interaction, leak = leak_item_into_class)) |>
  left_join(MODELS, by = "model") |>
  mutate(construction = factor(construction, c("POS (noun/verb)", "role (subject/object)",
                                               "metaphor (literal/figurative)")),
         depth = layer / ave(layer, model, construction, FUN = max))  # 0 = embeddings, 1 = final
```

# Introduction

::: {.content-visible when-format="acl-pdf"}
```{=html}
<!-- PLACEHOLDER: author to write. Frames the paper as an extension of Exemplars in
     Disguise: if a learner's behavior cannot distinguish exemplar storage from
     abstraction, ask instead whether its representations separate into an item part and
     a category part. Ends with the Present Study paragraph setting up the three
     experiments. -->
```
:::

# Experiment 1: Item and Category in a Hybrid Learner {#sec-exp1}

Before asking whether a language model's representations separate lexical identity from category, we need a setting where the right answer is known. Experiment 1 trains a learner on an artificial language whose generative structure we control, so that we can vary how much of the language is item-by-category specific and check what the measure recovers.

The learner is deliberately a hybrid of the two accounts at issue. Every word-in-category pairing receives its own private embedding, with no parameter sharing between them, which is exemplar storage; those embeddings are then read out through a single network shared by every pairing, which is abstraction. This is also the architecture of the models in Experiment 2, where private token embeddings feed shared transformer blocks. The question is what such a learner does with structure that *could* be represented separably.

## Dataset {#sec-exp1-data}

The artificial language is a next-token prediction task over a vocabulary of `r max(toy$n_lex) / 16` word types. Each **lexeme** is a (form, category) pair: 250 forms crossed with 16 categories, giving 4,000 lexemes, each with its own target distribution over the vocabulary. Forms are the items and categories are the classes, so the design is the same item-by-category layout used throughout the paper.

Each form $i$ and category $c$ carries a latent code, and the target distribution is generated from those codes through independently drawn loadings onto the vocabulary:

$$
\text{logits}(i,c) \;=\; s\Big[\sqrt{1-f}\;\overline{m}(i,c) \;+\; \sqrt{f}\;\overline{x}(i,c)\Big]
$$ {#eq-toy}

where $\overline{m}$ is an **additive** term, the sum of a form contribution and a category contribution, and $\overline{x}$ is a **bilinear** term formed from the elementwise product of the two codes, so it is irreducibly joint. Both are normalized to unit per-element standard deviation before mixing, and the $\sqrt{1-f}, \sqrt{f}$ weighting holds the total signal scale fixed, so the dial $f$ changes the *composition* of the language without changing how hard it is to learn.

That dial, $f \in \{0, .25, .5, .75, 1\}$, is the manipulation. At $f = 0$ a word's behavior is exactly its form contribution plus its category contribution, with **no interaction whatever**. At $f = 1$ behavior is purely the form-by-category conjunction and the marginals vanish. Intermediate values mix the two.

Two properties of the construction matter for what can be concluded. First, the three sets of loadings are drawn independently, so the additive and the interactive structure occupy approximately orthogonal subspaces of the output space: **a fully separable representation exists** for every value of $f$. Second, the learner's hidden layer is at least as wide as the generative rank in the cells we analyze, so it has room to place each component on its own axis. Any entanglement the measure reports is therefore something the learner did, not something the task or the architecture forced.

We cross the dial with the generative ranks of the form code ($r_i \in \{1,2,4,8,16,32\}$), the category code ($r_c \in \{1,2,4,8\}$), the hidden width ($d \in \{16,32,64,128\}$), the output nonlinearity (linear or ReLU) and five random draws of the language, for `r format(nrow(toy), big.mark = ",")` cells in total, all of which are analyzed. The ratio of generative rank to hidden width, which we call **capacity**, ranges across the grid from `r g(min(toy$cap), 2)` to `r g(max(toy$cap), 1)`. Below one a fully separable solution is reachable by dimension counting and above one it is not, so capacity is reported as a variable rather than used to select cells.

## Method {#sec-exp1-method}

### The learner

Each lexeme has a free $d$-dimensional embedding, initialized at random and trained without any sharing across lexemes: nothing ties *dog*-as-noun to *dog*-as-verb. Those embeddings pass through one hidden layer of width $d$ with either an identity or a ReLU nonlinearity, and then a linear map to the vocabulary. We measure the hidden layer.

The identity condition is the reference case. A linear map cannot destroy linear recoverability, so whatever the linear learner does to the geometry is a rotation and shear of separable structure rather than a fusion of it. Differences between the two conditions are attributable to the nonlinearity.

Training minimizes the exact expected cross-entropy against the true target distribution rather than a sampled-token estimate, so a cell is deterministic given its language and initialization, and a seed varies only the language and the starting weights. Cells are trained to a loss plateau.

### Decomposing a representation {#sec-measure}

Let $M[i,c] \in \mathbb{R}^d$ be the mean representation of item $i$ used at category level $c$. We split that grid into four geometric components,

$$
M[i,c] \;=\; \mu + \alpha(i) + \beta(c) + \gamma(i,c),
$$ {#eq-decomp}

by centering: $\mu$ is the grand mean, $\alpha(i)$ is item $i$'s row mean minus $\mu$, $\beta(c)$ is level $c$'s column mean minus $\mu$, and $\gamma(i,c)$ is what remains. The familiar reading is the main-effects-plus-interaction partition, with the difference that the dependent variable is a vector, so each component is a *direction* in representational space rather than a number. In mixed-model terms $\alpha$ is a by-item random intercept, $\beta$ is the fixed effect of category, and $\gamma$ is the by-item random slope for category; the question of this paper is how much of the category effect is carried by $\beta$ rather than by $\gamma$. We use that mapping for intuition only. Nothing is fitted, there is no shrinkage, and no distribution is assumed over the by-item terms.

A representation is **separable** if a word's representation is the sum of an item part and a category part, with the two on their own axes. It can fail in two distinguishable ways, and we measure them separately. It fails if the components are **non-orthogonal**, so that projecting out the category also removes item information; and it fails if there is a **third component**, an item-by-category term that is neither an item part nor a category part, whatever its orientation. A representation with a large $\gamma$ does not factor into item plus category even if $\gamma$ sits on an axis of its own.

Because the four components are mutually orthogonal *as sources*, their squared lengths partition the structure of the grid exactly, and we report each as a share of the total. **Interaction size** is $\gamma$'s share, which measures the second kind of failure. For the first we report two projections: **marginal leakage**, the fraction of the item component's energy lying inside the category component's subspace, and **interaction overlap**, how much the interaction's subspace shares directions with the item subspace. Zero on both, with a negligible $\gamma$, is a cleanly separable representation.

**Interpreting the interaction.** Because $\gamma$ is obtained by subtraction it invites reading as a residual in the regression sense, the part a model failed to capture. It is not. The decomposition is saturated: $\mu$, $\alpha$, $\beta$ and $\gamma$ exhaust the degrees of freedom of the grid, so $\gamma$ is the final term of a complete description rather than the leftover of an incomplete one. The residual stratum lies one level below, in the scatter of individual observations *within* each cell. That distinction is not free, however, since cell means are estimated from finitely many observations and within-cell error propagates upward into $\gamma$: a squared length is bounded below by its own noise floor and cannot read zero even when the true interaction is zero. This is the situation in which an analysis of variance tests an interaction against the within-cell error term, and it requires replication for the same reason.

**Replication and the null.** We therefore estimate the grid **twice** from independent data and keep only what the two estimates agree on, reporting the signed inner product $\langle \gamma_A, \gamma_B \rangle$ rather than $\lVert\gamma\rVert^2$. Independent errors cancel in expectation, so this is unbiased at any noise level; and because it can be negative it supports a genuine test, which a squared length cannot. Significance comes from permuting which item of one estimate is paired with which item of the other, which preserves every magnitude while destroying the correspondence. A component's *direction* is only defined if the component exists, so leakage is reported only when both components it relates are present; otherwise it is withheld.

In Experiment 1 the two estimates come from **two initializations trained on the same language**. This is necessary rather than incidental: the toy has one representation per cell and hence no within-cell replication, so interaction and error are not identifiable, in exactly the way they are not identifiable in an unreplicated two-way design. Training twice supplies the missing replication, and what it cancels is the initialization-dependent content that gradient descent leaves in the directions the loss does not constrain. Because two independently trained networks occupy different coordinate systems, one is rotated onto the other before comparison, with the rotation fitted on held-out items so that an item's own noise never shapes its own alignment. @sec-app-math gives the algebra.

## Results {#sec-exp1-results}

```{r}
#| label: fig-toy
#| fig-env: figure*
#| fig-width: 6.3
#| fig-height: 2.3
#| out-width: "100%"
#| fig-cap: "Experiment 1, all 4,800 cells. Left: the interaction's share of representational structure. Middle: leakage of the item component into the category subspace, on a log axis (it spans four orders of magnitude). Right: how much the interaction's subspace overlaps the item subspace. Points are medians. Both readouts build a substantial interaction, and it tracks the generative dial; only the ReLU readout also makes the components non-orthogonal. The two outer panels are shown for the settings where the quantities are defined: an orientation requires the components it relates to exist."
mid <- toy_dose |> filter(int_frac %in% TOY_MID)
pan <- function(dat, y, ylab, logy = FALSE) {
  p <- ggplot(dat, aes(int_frac, .data[[y]], colour = act)) +
    geom_line(linewidth = .6) + geom_point(size = 1.5) +
    scale_colour_manual(values = c(linear = UO[["green2"]], ReLU = UO[["gold"]]), name = NULL) +
    scale_x_continuous(breaks = c(0, .5, 1)) +
    labs(x = NULL, y = ylab) + theme_paper(8)
  if (logy) p + scale_y_log10(labels = scales::label_number(drop0trailing = TRUE)) else p
}
((pan(toy_dose, "size_med", "interaction share\nof structure") |
  pan(filter(toy_dose, int_frac > 0), "leak_med", "item leaks into\ncategory (log)", TRUE) |
  pan(mid, "ov_med", "interaction overlaps\nthe item subspace")) +
  plot_layout(guides = "collect") &
  theme(legend.position = "top", legend.margin = margin(0, 0, -6, 0)))
```

**The measure tracks what the language contains.** With a purely additive language the estimator reports a median interaction share of `r g(toy_dose$size_med[toy_dose$act == "linear" & toy_dose$int_frac == 0], 4)` under a linear readout; with a purely interactive one it reports `r g(pick(toy_dose, "linear", 1, "size_med"), 3)`. The instrument exercises nearly its full range on a task whose answer is known at both ends (@fig-toy, left).

**Even the linear learner builds an interaction.** At every intermediate setting the linear readout carries an item-by-category component of `r g(pick(toy_dose, "linear", .25, "size_med"), 2)` to `r g(pick(toy_dose, "linear", .75, "size_med"), 2)` of representational structure, flagged as present in **100%** of cells. This is not optional. A linear readout maps the hidden layer to the logits, and a linear map of a purely additive representation can only produce additive logits; so any learner that actually fits an interactive language must carry a non-additive representation. The interaction is forced by the task, not chosen by the architecture.

**The linear learner nonetheless keeps everything orthogonal.** Its item and category components stay on essentially separate axes, with leakage from `r g(pick(toy_dose, "linear", .25, "leak_med"), 5)` to `r g(pick(toy_dose, "linear", .75, "leak_med"), 4)`, and its interaction occupies its own axis, overlapping the item subspace at `r g(pick(toy_dose, "linear", .25, "ov_med"), 3)` and `r g(pick(toy_dose, "linear", .5, "ov_med"), 3)` at the two lower settings. So the linear representation fails separability in exactly one of the two senses: it contains a third component that is neither item nor category, while the components it does contain are mutually orthogonal.

**The nonlinearity compounds it, on both counts.** A ReLU readout carries an interaction of similar or slightly larger size (`r g(pick(toy_dose, "ReLU", .75, "size_med"), 2)` at the highest intermediate setting), but now the components are also non-orthogonal. Marginal leakage runs `r g(pick(toy_dose, "ReLU", .25, "leak_med"), 3)` to `r g(pick(toy_dose, "ReLU", .75, "leak_med"), 3)`, two orders of magnitude above the linear readout at matched settings, and the interaction overlaps the item subspace at `r g(pick(toy_dose, "ReLU", .25, "ov_med"), 2)` to `r g(pick(toy_dose, "ReLU", .75, "ov_med"), 2)` where the linear readout sits near zero. Under a nonlinearity the interaction stops being a separate module and is superposed onto the code that carries item and category identity.

**Capacity governs whether the entanglement was avoidable** (@tbl-toy). Where the hidden layer has room for every generative direction, the linear readout's interaction overlaps the item subspace at `r g(capget(toy_cap, "linear", "room", "ov_med"), 3)`: separation was achievable and it was achieved. Where dimension counting makes separation impossible, the same linear readout reaches `r g(capget(toy_cap, "linear", "impossible", "ov_med"), 2)`. The ReLU readout is entangled at `r g(capget(toy_cap, "ReLU", "room", "ov_med"), 2)` even with room to spare, so its entanglement is learned rather than forced. This is also why the additive control must be read by capacity: with an additive language and room to spare the linear readout flags an interaction in `r sprintf("%.0f%%", 100 * capget(toy_null, "linear", "room", "sig_rate"))` of cells, but where separation is impossible it flags one in `r sprintf("%.0f%%", 100 * capget(toy_null, "linear", "impossible", "sig_rate"))`, correctly, since a representation with nowhere to put the components cannot keep them apart.

```{r}
#| label: tbl-toy
#| tbl-cap: "Experiment 1, by capacity. Medians over the intermediate settings of the dial, where both the interaction and the marginals exist. Capacity is the generative rank over the hidden width: below one, a fully separable solution is reachable. Leakage and overlap are zero for a separable representation."
toy_cap |>
  transmute(readout = act,
            capacity = gsub("\n", " ", band),
            `interaction share` = g(size_med, 3),
            `item into category` = g(leak_med, 4),
            `interaction into margins` = g(leakint_med, 3),
            `interaction / item overlap` = g(ov_med, 3)) |>
  arrange(readout, capacity) |>
  kbl(booktabs = TRUE, align = "llrrrr", linesep = "") |>
  kable_styling(latex_options = c("HOLD_position"), font_size = 8)
```

Two conclusions carry into Experiment 2. Separability fails for a **principled** reason and not an accidental one: no learner that fits an interactive language can represent it as item plus category, because the readout would then be unable to produce the interaction the language requires. And the *further* failure, the components ceasing to be orthogonal, is specifically what a nonlinearity adds. Real language models are nonlinear and real language is interactive, so both mechanisms are in play before we look.

# Experiment 2: Item and Category in Language Models {#sec-exp2}

Experiment 1 established that a learner fitting an interactive language cannot represent it as item plus category, and that a nonlinearity additionally binds the components onto shared axes. Experiment 2 applies the same decomposition to the contextual representations of pretrained language models, across three linguistic distinctions chosen to span grammar and meaning.

## Dataset {#sec-exp2-data}

**Corpora.** For part of speech and grammatical role we concatenate six general-domain English Universal Dependencies treebanks [@nivre2020ud]: EWT, GUM, LinES, ParTUT, GENTLE and PUD, giving `r cm(41095)` sentences and `r cm(706323)` tokens. Concatenating rather than using EWT alone is what makes the analysis possible: a lemma must be attested in both category levels often enough to estimate a cell mean, and a single treebank yields too few such lemmas. For metaphor we use the VUA20 corpus [@leong2020vua; @steen2010vua], restricted to content words, which supplies `r cm(gs("metaphor", "n_pre_grid_tokens"))` labeled tokens.

**Constructions.** All three are **same-token**: the item's surface form is identical across the two category levels, so the representation is read from the same token in both, and no part of any difference can come from tokenisation.

| distinction | item | levels | eligible | analyzed | tokens |
|:-----------|:-----------|:-----------|-----------:|-----------:|-----------:|
| part of speech | lemma | [noun]{.smallcaps} / [verb]{.smallcaps} | `r cm(gs("pos_noun_verb", "n_items_eligible"))` | `r gs("pos_noun_verb", "n_items_kept")` | `r cm(gs("pos_noun_verb", "n_grid_tokens"))` |
| grammatical role | surface form | *nsubj* / *obj* | --- | 49 | `r cm(2150)` |
| metaphor | surface form | literal / figurative | `r cm(gs("metaphor", "n_items_eligible"))` | `r gs("metaphor", "n_items_kept")` | `r cm(gs("metaphor", "n_grid_tokens"))` |

: The three constructions. *Eligible* counts items attested at both levels anywhere in the corpus; *analyzed* counts those attested often enough at each level to estimate a cell mean. {#tbl-constructions}

**Identifiability.** The interaction $\gamma(i,c)$ is only estimable for an item observed at more than one level, since it is defined by how an item's representation shifts across levels. We additionally require at least 10 tokens per cell, which is what allows a cell mean to be estimated and then split in half for the replication the measure needs (@sec-exp1-method). This is a real restriction on scope: for part of speech it selects *convertible* lemmas, words used both as a noun and as a verb, such as *run* or *hit*. Role and metaphor carry no such restriction, since any noun occurs as both subject and object and any content word occurs both literally and figuratively, which is one reason they broaden the claim. The median cell holds `r gs("pos_noun_verb", "tokens_per_cell_median")` tokens for part of speech and `r gs("metaphor", "tokens_per_cell_median")` for metaphor.

For role and metaphor the item is keyed on the **surface form** rather than the lemma, which guarantees the identical token across levels: a lemma like *thing* can differ in number across the two roles, whereas *things*-as-subject and *things*-as-object are the same string.

**Models.** Two families at three matched scales: OPT-BabyLM, trained on developmentally plausible quantities of child-available language [@warstadt2023babylm], and Pythia, trained on the Pile [@biderman2023pythia], at 125M/160M, 350M/410M and 1.3B/1.4B parameters, with model widths of 768, 1024 and 2048.

## Method {#sec-exp2-method}

The decomposition, the two reported quantities and the replication logic are exactly those of @sec-exp1-method. Three things differ.

**Representations.** For every token we take the hidden state of the word's **last subword** at every layer, from the embeddings through the final block, aligning subwords to corpus tokens with the tokenizer's word alignment. Sentences are drawn in a fixed random order and processed until roughly 300K tokens are extracted per model, so all models see the same material. Representations are extracted once and every analysis runs offline from them.

**Where the two independent estimates come from.** Experiment 1 had to train the model twice, because a toy lexeme has a single representation and therefore no within-cell replication. A language model supplies replication for free: each cell is a word used at one level across many sentences. We therefore split each cell's **token contexts** into two disjoint halves and estimate the grid from each. What cancels is context noise, the variation in a word's hidden state across the sentences it appears in. Because both halves come from the same frozen model they already share a coordinate system, so the rotation step of @sec-exp1-method is unnecessary here.

**Standardization.** Transformer hidden states contain a few *rogue* dimensions whose variance exceeds the rest by two to three orders of magnitude [@timkey2021bark], and these would otherwise dominate any subspace estimated from the raw scatter. We divide each dimension by its standard deviation before decomposing. This is a diagonal rescaling, so unlike whitening it needs no favorable ratio of samples to dimensions; the measure is accordingly invariant to per-dimension rescaling but not to arbitrary linear reparameterization, which we return to in @sec-app-math.

# Experiment 3: Does the Interaction Carry the Category? {#sec-exp3}

```{=html}
<!-- Method: hold the item fixed, project out the shared category axis, decode the category
     from the residual with a cross-validated probe against a within-item permutation null,
     plus the three metaphor probes. Awaiting llm_metaphor_probes.csv and the decode re-run. -->
```

# Discussion {#sec-discussion}

::: {.content-visible when-format="acl-pdf"}
<!-- PLACEHOLDER: author to write. -->
:::

# Limitations {.unnumbered}

<!-- PLACEHOLDER -->

::: {=latex}
\appendix
\floatplacement{figure}{H}
\floatplacement{table}{H}
\raggedbottom
:::

# The Decomposition and Its Estimator {#sec-app-math}

```{=html}
<!-- Formal development: the tensor-product projection view, the exact size partition and
     its invariance to grid size, leakage against its r/d null, the positive bias of a
     squared projection norm, and the unbiasedness of the cross-estimate inner product.
     Drawn from MATH.md. -->
```
